### 2017-03-11 preproposal/preproposal.tex
update preproposal

%!TEX program = xelatex

\documentclass[11pt]{article}

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{titlesec}
\usepackage{wrapfig}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{sidecap}
\usepackage{clrscode}
\usepackage{xcolor}
\usepackage{float}
\usepackage{stackengine}

\usepackage[margin=0.75in]{geometry}

\newlength{\tindent}
\setlength{\tindent}{\parindent}
\setlength{\parindent}{0pt}
\renewcommand{\indent}{\hspace*{\tindent}}

\renewcommand{\thesubsection}{\thesection.\alph{subsection}}
\titleformat{\subsection}[runin]{\normalfont\large\bfseries}{\thesubsection}{1em}{}

\begin{document}

\begin{centering}
\Large
    \textbf{Time-series and Hierarchical Modeling for sc-RNA Expression Levels}  \\
    \vspace{2mm}
    \normalsize
    Skanda Koppula (\url{skoppula@mit.edu}), Karren Yang (\url{karren@mit.edu}) \\
    \vspace{2mm}
    \normalsize
    6.882 Project Pre-Proposal, \today \\
\end{centering}
\vspace{5mm}

Cancer cells implanted into mice exhibit patterned levels of RNA expression. The expression in each cancer cell varies over time, and the cells may naturally diverge in their expression during this time. Previous work has looked into an LDA-based clustering of expression vectors for particular time slices, resulting in interesting and biologically meaningful cell grouping shifts over time. \\

This project is interested in constructing the first generative model of this process. Specifically, we look to answer some of the following questions: 
\begin{enumerate}
\item Can we apply a topic model based probabalistic model for a collection of expression vectors for a particular time slice? Will a topic model based solution yield meaningful cell type assignments?
\item How can we apply time-series probabilistic models such as Chinese Restaurant to model cell types over time?
\item How can we model noise in expression level observations in our probabilistic models? Is it possible to use our model to establish estimates for missing coordinates in expression vectors (a common problem in high-throughput expression measurement machines)?
\item Is it possible to emulate expression data with a simple sampling processes (for the purposes of testing any models we create)?
\end{enumerate}
These were a few lines of inquiry we were interested in pursuing. The authors have access to an sc-RNA expression level dataset through their research lab.
\end{document}

### 2017-03-11 preproposal/build.sh

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+file="preproposal"
+author1="karren"
+author2="skoppula"
+totalfile="$file-$author1-$author2"
+
+# Compile the tex file
+pdflatex --shell-escape -jobname=$totalfile ${file}.tex
+pdflatex --shell-escape -jobname=$totalfile ${file}.tex
+
+if [ "$(uname)" == "Darwin" ]; then
+    open ${totalfile}.pdf &
+elif [ "$(expr substr $(uname -s) 1 5)" == "Linux" ]; then
+    evince ${totalfile}.pdf &
+elif [ "$(expr substr $(uname -s) 1 10)" == "MINGW32_NT" ]; then
+    echo 'Windows not supported!'
+fi
+
+rm *.aux *.log *.out

### 2017-03-11 proposal/build.sh
add preproposal

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+file="proposal"
+author1="karren"
+author2="skoppula"
+totalfile="$file-$author1-$author2"
+
+# Compile the tex file
+pdflatex --shell-escape -jobname=$totalfile ${file}.tex
+pdflatex --shell-escape -jobname=$totalfile ${file}.tex
+
+if [ "$(uname)" == "Darwin" ]; then
+    open ${totalfile}.pdf &
+elif [ "$(expr substr $(uname -s) 1 5)" == "Linux" ]; then
+    evince ${totalfile}.pdf &
+elif [ "$(expr substr $(uname -s) 1 10)" == "MINGW32_NT" ]; then
+    echo 'Windows not supported!'
+fi
+
+rm *.aux *.log *.out

### 2017-03-11 preproposal/preproposal.tex
Updated preproposal abstract

@@
 \Large
-    \textbf{Time-series and Hierarchical Modeling for sc-RNA Expression Levels}  \\
+    \textbf{Bayesian Non-Parametric Models for Analyzing Single-Cell RNA Sequencing Data}  \\
     \vspace{2mm}
@@
 
-Cancer cells implanted into mice exhibit patterned levels of RNA expression.
-The expression in each cancer cell varies over time, and the cells may naturally diverge in their expression during this time.
-Previous work has looked into an LDA-based clustering of expression vectors for particular time slices, resulting in interesting and biologically meaningful cell grouping shifts over time.
-\\
+Cancer cells injected in mice grow into tumors composed of distinct sub-populations of cells, which exhibit patterned levels of gene expression.
+In order to characterize sub-populations of tumor cells and identify useful gene modules, we propose applying Bayesian nonparametrics to model single-cell RNA-sequencing (scRNA-seq) data from mouse tumors.
+Parametric topic models based on matrix factorization have shown promise in identifying meaningful clusters of cells and genes in heterogeneous cell populations.
+However, the need to pre-determine the number of clusters is a major limitation, particularly when extending these models to time-series data, as sub-populations of cells may appear, disappear, or diverge in expression over time.
+We propose exploring the utility of nonparametric models based on the Hierarchical Dirichlet Process (HDP) and the Indian Buffet Process (IDP) for modeling scRNA-seq data from single time points.
+We plan to compare their performance to that of existing methods for scRNA-seq analysis, as well as that of the top parametric topic model, Latent Dirichlet Allocation (LDA).
+Time permitting, we would also like to explore the extension of these models to time-series data, in order to understand how sub-populations of cells in the tumor microenvironment evolve over time.
 
-This project is interested in constructing the first generative model of this process. Specifically, we look to answer some of the following questions: 
-\begin{enumerate}
-\item Can we apply a topic model based probabalistic model for a collection of expression vectors for a particular time slice? Will a topic model based solution yield meaningful cell type assignments?
-\item How can we apply time-series probabilistic models such as Chinese Restaurant to model cell types over time?
-\item How can we model noise in expression level observations in our probabilistic models?
-Is it possible to use our model to establish estimates for missing coordinates in expression vectors (a common problem in high-throughput expression measurement machines)?
-\item Is it possible to emulate expression data with a simple sampling processes (for the purposes of testing any models we create)?
-\end{enumerate}
-These were a few lines of inquiry we were interested in pursuing. The authors have access to an sc-RNA expression level dataset through their research lab.
 \end{document}
+\grid
+\grid

### 2017-03-11 preproposal/build.sh

@@ -7,7 +7,6 @@ totalfile="$file-$author1-$author2"
 
 # Compile the tex file
 pdflatex --shell-escape -jobname=$totalfile ${file}.tex
-pdflatex --shell-escape -jobname=$totalfile ${file}.tex
 
 if [ "$(uname)" == "Darwin" ]; then
     open ${totalfile}.pdf &

### 2017-03-21 preproposal/preproposal.tex
added objectives and tentative schedule

@@
 In order to characterize sub-populations of tumor cells and identify useful gene modules, we propose applying Bayesian nonparametrics to model single-cell RNA-sequencing (scRNA-seq) data from mouse tumors.
-Parametric topic models based on matrix factorization have shown promise in identifying meaningful clusters of cells and genes in heterogeneous cell populations.
-However, the need to pre-determine the number of clusters is a major limitation, particularly when extending these models to time-series data, as sub-populations of cells may appear, disappear, or diverge in expression over time.
+Parametric topic models based on matrix factorization have shown promise in identifying meaningful clusters of cells and genes in heterogeneous cell populations; however, the need to pre-determine the number of clusters is a major limitation, particularly when extending these models to time-series data, as sub-populations of cells may appear, disappear, or diverge in expression over time.
 We propose exploring the utility of nonparametric models based on the Hierarchical Dirichlet Process (HDP) and the Indian Buffet Process (IDP) for modeling scRNA-seq data from single time points.

### 2017-03-21 proposal/proposal.tex
add preproposal
added objectives and tentative schedule

--- 2017-03-11 preproposal/preproposal.tex
@@
 \Large
-    \textbf{Time-series and Hierarchical Modeling for sc-RNA Expression Levels}  \\
+    \textbf{Bayesian Clustering and Topic Modeling for Analyzing Single-Cell RNA Sequencing Data}  \\
     \vspace{2mm}
@@
     \normalsize
-    6.882 Project Pre-Proposal, \today \\
+    6.882 Project Proposal, \today \\
 \end{centering}
@@
 
-Cancer cells implanted into mice exhibit patterned levels of RNA expression.
-The expression in each cancer cell varies over time, and the cells may naturally diverge in their expression during this time.
-Previous work has looked into an LDA-based clustering of expression vectors for particular time slices, resulting in interesting and biologically meaningful cell grouping shifts over time.
-\\
+\begin{abstract}
 
-This project is interested in constructing the first generative model of this process. Specifically, we look to answer some of the following questions: 
+Tumors are composed of many sub-populations of cells, which exhibit different patterns of gene expression.
+In order to characterize sub-populations of tumor cells and identify functional gene modules, we propose applying Bayesian methods (parametric and non-parametric) for clustering and topic modeling to single-cell RNA-sequencing (scRNA-seq) data from tumors.
+We will compare these methods to standard methods for analyzing scRNA-seq data based their ability to identify known cell sub-populations and gene modules from the literature as well as their robustness to noise.
+We will furthermore determine if integrating the clustering model with the topic model results in superior performance in these tasks.
+Time permitting, we will attempt to extend these models to accommodate time-series data.
+
+\end{abstract}
+
+\section{Overview}
+
+Using single-cell RNA-sequencing (scRNA-seq) data, it is possible to obtain the expression levels of 10,000+ genes for 1000+ cells in a single experiment.
+Given such data, we are primarily interested in answering the following questions:
+
 \begin{enumerate}
-\item Can we apply a topic model based probabalistic model for a collection of expression vectors for a particular time slice? Will a topic model based solution yield meaningful cell type assignments?
-\item How can we apply time-series probabilistic models such as Chinese Restaurant to model cell types over time?
-\item How can we model noise in expression level observations in our probabilistic models?
-Is it possible to use our model to establish estimates for missing coordinates in expression vectors (a common problem in high-throughput expression measurement machines)?
-\item Is it possible to emulate expression data with a simple sampling processes (for the purposes of testing any models we create)?
+    \item Can Bayesian nonparametric topic models, such as the Hierarchical Dirichlet Process (HDP) and the Indian Buffet Process Compound Dirichlet Process (IBP-CDP), extract functional gene modules from this data?
+Do Bayesian nonparametric topic models perform as well as parametric topic models, such as Latent Dirichlet Allocation (LDA), on this data?
+    \item Can we get meaningful cell-type assignments by clustering cells, using finite or infinite mixture models, based on the output of the topic models?
+    \item How robust are these methods to dropout, a phenomenon commonly observed in real scRNA-seq data?
+    \item Can integrated models of cell clustering and topic modeling yield superior results in these tasks?
+    \item Can we extend any of these models to time-series data?
+
 \end{enumerate}
-These were a few lines of inquiry we were interested in pursuing. The authors have access to an sc-RNA expression level dataset through their research lab.
+
+Risk: For a 6-7 week project, this list seems ambitious, but we have mitigated the risk in several ways.
+(1) Although we are primarily interested in the utility of nonparametric methods for topic (gene module) modeling, we are also using Latent Dirichlet Allocation so that subsequent steps can proceed even if the nonparametric methods do not work well.
+(2) If we become stuck on steps 4-5, we can submit a good report from objectives 1-3.
+
+\section{Tentative Schedule}
+\subsection{Apply LDA and HDP to scRNA-seq data - SK}
+\subsection{Apply IBP-CDP to scRNA-seq data - KY}
+\subsection{Evaluate biological relevance of gene modules from topic models - KY}
+\subsection{Evaluate robustness of topic models to dropout - SK}
+\subsection{Apply finite mixture model to cluster topic model output - SK}
+\subsection{Apply infinite mixture model to cluster topic model output - KY}
+\subsection{Integrate topic modeling and cell clustering - TBA}
+\subsection{Extend model to time-series data - TBA}
+
+\begin{enumerate}
+    \item Referenced paper, outside/related work
+    \item Outline work to be done
+    \item 4 steps per team member, internal deadlines
+    \item Risks: what might be more difficult than planned, thoughts to mitigate these
+    \item How to evaluate our methods?
+\end{enumerate}
 \end{document}

### 2017-03-22 proposal/proposal.tex
done

--- 2017-03-11 preproposal/preproposal.tex
@@
 
+\newcommand\tab[1][1cm]{\hspace*{#1}}
+\newcommand{\SK}{\textbf{[SK]}\tab}
+\newcommand{\KY}{\textbf{[KY]}\tab}
+\newcommand{\TBD}{\textbf{[TBD]}\tab}
+\newcommand\cdate[1][1cm]{\textbf{[#1]}}
+
 \renewcommand{\thesubsection}{\thesection.\alph{subsection}}
@@
 \Large
-    \textbf{Time-series and Hierarchical Modeling for sc-RNA Expression Levels}  \\
+    \textbf{Bayesian Clustering and Topic Modeling of Single-Cell RNA}  \\
     \vspace{2mm}
@@
 
-Cancer cells implanted into mice exhibit patterned levels of RNA expression.
-The expression in each cancer cell varies over time, and the cells may naturally diverge in their expression during this time.
-Previous work has looked into an LDA-based clustering of expression vectors for particular time slices, resulting in interesting and biologically meaningful cell grouping shifts over time.
+\section{Overview}
+
+Tumors are composed of different sub-populations of cells.
+These sub-populations exhibit shared patterns of gene expression, and prior research has applied basic measures of statistical distance to cluster cell groups based on samples of single-cell RNA-sequencing (scRNA-seq) data taken from tumors \cite{nature}.
+With contemporary sequencing machines, it is possible to obtain the expression levels of 10,000+ genes for 1000+ cells in a single experiment.
+Good clustering of sc-RNA data often has
+biologically meaningful results, and as such, a wide variety of correlation based methods have been used in prior art to derive meaning from scRNA-seq data \cite{coexpression, consensus, profiling}.
 \\
 
-This project is interested in constructing the first generative model of this process. Specifically, we look to answer some of the following questions: 
+In this project, we aim to apply Bayesian methods to study scRNA-seq datasets from the author's lab.
+Specifically, we intend to explore the use of a hierarchical topic model and combined clustering-topic model (e.g.
+MGCTM \cite{pengtao}) to analyze both real and simulated scRNA-seq data.
+We will compare the results of these methods with standard methods used in prior work, and evaluate these methods' robustness to noise.
+Time permitting, we will explore the use of non-parametric models, and study time-series sc-RNA data.
+\\
+
+In summary, and in order of importance, we are primarily interested in answering the following questions:
+
 \begin{enumerate}
-\item Can we apply a topic model based probabalistic model for a collection of expression vectors for a particular time slice? Will a topic model based solution yield meaningful cell type assignments?
-\item How can we apply time-series probabilistic models such as Chinese Restaurant to model cell types over time?
-\item How can we model noise in expression level observations in our probabilistic models?
-Is it possible to use our model to establish estimates for missing coordinates in expression vectors (a common problem in high-throughput expression measurement machines)?
-\item Is it possible to emulate expression data with a simple sampling processes (for the purposes of testing any models we create)?
+    \item Can we get meaningful cell-type assignments by clustering cells, using finite or infinite mixture models, based on the output of the topic models? How does this compare to existing methods?
+    \item How robust are these methods to data dropout (commonly observed in real scRNA-seq data)?
+    \item Do integrated models of cell clustering and topic modeling yield superior results in these tasks? 
+    \item Can nonparametric topic models, such as the Hierarchical Dirichlet Process (HDP) and the Indian Buffet Process Compound Dirichlet Process (IBP-CDP), extract functional gene modules from this data?
+Do nonparametric topic models perform as well as parametric topic models on this data?
+    \item Can we extend any of these models to time-series data?
 \end{enumerate}
-These were a few lines of inquiry we were interested in pursuing. The authors have access to an sc-RNA expression level dataset through their research lab.
+
+\section{Tentative Schedule and Task Breakdown}
+\begin{enumerate}
+    \item \cdate[04/02] \SK Apply LDA and HDP to scRNA-seq data
+    \item \cdate[04/08] \KY Apply IBP-CDP to scRNA-seq data
+    \item \cdate[04/10] \KY Evaluate biological relevance of gene modules from topic models
+    \item \cdate[04/13] \SK Evaluate robustness of topic models to dropout
+    \item \cdate[04/22] \SK Apply finite mixture model to cluster topic model output
+    \item \cdate[04/25] \KY Apply infinite mixture model to cluster topic model output
+    \item \cdate[05/05] \TBD Integrate topic modeling and cell clustering (MGCTM)
+    \item \cdate[05/05] \TBD Extend model to time-series data
+\end{enumerate}
+
+\section{Risk and Evaluation}
+For a 6-7 week project, this list seems ambitious, but we have mitigated the risk in several ways.
+(1) Although we are primarily interested in the utility of nonparametric methods for topic (gene module) modeling, we are also using Latent Dirichlet Allocation so that subsequent steps can proceed even if the nonparametric methods do not work well.
+(2) If we become stuck on steps 4-5, we can submit a good report from objectives 1-3.
+\\
+
+In addition to standards methods to evaluate model fit (e.g.
+likelihood of held-out data, model visualization, etc.), we can cross-reference prior work, and, when possible, cross-check the biological interpretation of our results.
+
+\begin{thebibliography}{9}
+    \bibitem{nature} 
+        Spearman-based hierarchical clustering of scRNA-seq.
+        \url{http://www.nature.com/nsmb/journal/v24/n3/fig_tab/nsmb.3365_SF1.html}
+
+    \bibitem{coexpression} 
+        Exploiting single-cell expression to characterize co-expression replicability.
+        \url{https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-0964-6}
+
+    \bibitem{consensus} 
+        SC3 - consensus clustering of single-cell RNA-Seq data.
+        \url{http://biorxiv.org/content/early/2016/09/02/036558}
+
+    \bibitem{profiling} 
+        SINCERA: A Pipeline for Single-Cell RNA-Seq Profiling Analysis.
+        \url{http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004575}
+
+    \bibitem{pengtao} 
+        Integrating Document Clustering and Topic Modeling.
+        \url{https://arxiv.org/pdf/1309.6874.pdf}
+\end{thebibliography}
+
 \end{document}

### 2017-04-04 hbp.py
add preproces template and hbp template

@@ -1,0 +1,4 @@
+from preprocess_dataset import load_dataset
+
+data, genes, cells = load_dataset()
+print("Data shape:", data.shape)

### 2017-04-04 preprocess_dataset.py

@@ -1,0 +1,74 @@
+
+# coding: utf-8
+
+# In[1]:
+
+import pandas as pd
+import os.path
+import sys
+import numpy as np
+
+
+# In[2]:
+
+def load_dataset(preprocessed_path='dataset/preprocessed/human_melanoma_data.npy',
+                 raw_path='dataset/human_melanoma_data.tsv'):
+
+    if os.path.isfile(preprocessed_path):
+        return np.load(preprocessed_path),             np.load(preprocessed_path.split('.')[0] + "_column_lbls.npy"),             np.load(preprocessed_path.split('.')[0] + "_row_lbls.npy")
+
+    elif os.path.isfile(raw_path):
+        expr_data = pd.read_csv(raw_path, delimiter='\t').T
+        print("Number of cells:", expr_data.shape[0])
+        print("Number of genes:", expr_data.shape[1])
+
+        # excise stop genes
+        STOP_GENES = ['ACTB', 'B2M', 'EEF1A1', 'MTRNR2L1', 'MTRNR2L2',
+                  'MTRNR2L8', 'RPL41', 'RPLP1', 'RPS18', 'RPS27', 'TMSB4X']
+        expr_data = expr_data.drop(STOP_GENES, 1)
+        print("After cutting out stop genes:")
+        print("\tNumber of cells:", expr_data.shape[0])
+        print("\tNumber of genes:", expr_data.shape[1])
+
+        genes = list(expr_data.columns.values)
+        ribosomal_genes = [gene for gene in genes if gene.startswith('RPL')]
+        mito_ribosomal_genes = [gene for gene in genes if gene.startswith('MRPL')]
+        excise_genes = ribosomal_genes + mito_ribosomal_genes
+        expr_data = expr_data.drop(excise_genes, 1)
+        print("After cutting out ribosomal protein genes:")
+        print("\tNumber of cells:", expr_data.shape[0])
+        print("\tNumber of genes:", expr_data.shape[1])
+
+        # excise zero cells
+        expr_data = expr_data.loc[~(expr_data == 0).all(axis=1)]
+        print("After cutting out all-zero rows [cells]:")
+        print("\tNumber of cells:", expr_data.shape[0])
+        print("\tNumber of genes:", expr_data.shape[1])
+
+        # excise zero genes
+        expr_data = expr_data.loc[:, ~(expr_data == 0).all(axis=0)]
+        print("After cutting out all-zero columns [genes]:")
[…]

### 2017-04-04 proposal/proposal.tex

@@
 \newcommand{\KY}{\textbf{[KY]}\tab}
-\newcommand{\TBD}{\textbf{[TBD]}\tab}
+\newcommand{\CO}{\textbf{[CO]}\tab}
 \newcommand\cdate[1][1cm]{\textbf{[#1]}}
@@
 \Large
-    \textbf{Bayesian Clustering and Topic Modeling of Single-Cell RNA}  \\
+    \textbf{Bayesian Clustering and Topic Modeling of Gene Expression Data}  \\
     \vspace{2mm}
@@
 
-Tumors are composed of different sub-populations of cells.
-These sub-populations exhibit shared patterns of gene expression, and prior research has applied basic measures of statistical distance to cluster cell groups based on samples of single-cell RNA-sequencing (scRNA-seq) data taken from tumors \cite{nature}.
+Tumors are composed of different sub-populations of cells, and these sub-populations exhibit shared patterns of gene expression.
 With contemporary sequencing machines, it is possible to obtain the expression levels of 10,000+ genes for 1000+ cells in a single experiment.
-Good clustering of sc-RNA data often has
-biologically meaningful results, and as such, a wide variety of correlation based methods have been used in prior art to derive meaning from scRNA-seq data \cite{coexpression, consensus, profiling}.
+In this project, we aim to apply Bayesian methods for clustering and topic modeling to discover meaningful cell clusters and gene modules from single-cell RNA-sequencing (scRNA-seq) datasets from tumors.
+We first intend to explore the use of a hierarchical topic model, Latent Dirichlet Allocation (LDA) \cite{LDA}, as well as two non-parametric topic models, Hierarchical Dirichlet Processes (HDP) \cite{HDP} and the Indian Buffet Process Compound Dirichlet Process (IBP-CDP) \cite{IBP} for analyzing scRNA-seq data.
+We will evaluate these models based on their ability to discover meaningful functional gene modules.
+Subsequently, we plan to explore the use of mixture models to cluster cells based on the output of the topic models, which we will evaluate based on their ability to differentiate cell types (e.g.
+immune, cancer, non-cancerous).
+Finally, we will train a combined clustering-topic model (e.g.
+MGCTM \cite{pengtao}) to see if it outperforms the individual models.
+We will compare the results of these methods with standard methods used in prior work, and evaluate these methods' robustness to noise.
+Time permitting, we will study extensions of these models appropriate for time-series scRNA-seq data \cite{TS-DDP}.
 \\
 
-In this project, we aim to apply Bayesian methods to study scRNA-seq datasets from the author's lab.
-Specifically, we intend to explore the use of a hierarchical topic model and combined clustering-topic model (e.g.
-MGCTM \cite{pengtao}) to analyze both real and simulated scRNA-seq data.
-We will compare the results of these methods with standard methods used in prior work, and evaluate these methods' robustness to noise.
-Time permitting, we will explore the use of non-parametric models, and study time-series sc-RNA data.
+Prior research has applied basic measures of statistical distance to cluster cell groups or gene modules based on samples of single-cell RNA-sequencing (scRNA-seq) data taken from tumors \cite{nature}.
+Good clustering of scRNA-seq data often has biologically meaningful results, and as such, a wide variety of correlation based methods have been used in prior work to derive meaning from scRNA-seq data \cite{coexpression, consensus, profiling}.
+We believe that Bayesian methods for clustering and topic modeling can be more illuminating than these previous methods, due to their consideration of higher-order relationships within the data.
 \\
@@
 \begin{enumerate}
+    \item Do Bayesian topic models such as LDA, HDP and IBP-CDP work well on scRNA-seq data? Can we extract functional gene modules from the topics? \\
+    Topics are to documents what gene modules are to cells; therefore, we will apply these topic models to scRNA-seq datasets to discover genes that tend to be co-expressed in cells, which suggests they are functionally related.
+We are interested in using LDA, as it is the top-performing parametric topic model and its implementation is available online \cite{LDA}.
+We are interested in HDP because it is a non-parametric extension of LDA, and it might be more flexible as we would not need to designate the number of gene modules beforehand \cite{HDP}.
+Its implementation is also available online.
+Finally, we are interested in IBP-CDP because it offers the nonparametric flexibility of HDP, plus it overcomes the assumptions of HDP that gene modules active in many cells must also be highly expressed \cite{IBP}.
+Its implementation is not available online, but we plan to ask the authors for their code.
+
     \item Can we get meaningful cell-type assignments by clustering cells, using finite or infinite mixture models, based on the output of the topic models? How does this compare to existing methods?
-    \item How robust are these methods to data dropout (commonly observed in real scRNA-seq data)?
-    \item Do integrated models of cell clustering and topic modeling yield superior results in these tasks? 
-    \item Can nonparametric topic models, such as the Hierarchical Dirichlet Process (HDP) and the Indian Buffet Process Compound Dirichlet Process (IBP-CDP), extract functional gene modules from this data?
-Do nonparametric topic models perform as well as parametric topic models on this data?
-    \item Can we extend any of these models to time-series data?
+    \\
+    Just as document clustering can be done on topic proportions, we plan to cluster cells based on their gene module proportions.
+We plan to use a finite Gaussian mixture model, as well as an infinite Gaussian mixture model based on the Dirichlet Process.
+
+    \item How robust are these methods to data dropout? \\
+    Real scRNA-seq data suffers from dropout (i.e.
+a gene that was actually expressed in a cell has zero count for that cell, due to noise and limitations in experimental detection).
+We hypothesize that clustering based on gene module proportions will be more robust to dropout than clustering based on individual gene counts.
+
+    \item Do integrated models of cell clustering and topic modeling yield superior results in these tasks? \\
+    We plan to compose our best topic model and best clustering method into an integrated model, following the example of the MGCTM \cite{pengtao}. 
+    \item Can we extend any of these models to time-series data? \\
+    We hope to extend our models to time-series data, for example, using Hidden Markov Models or Dependent Dirichlet Process \cite{TS-DDP}.
+
 \end{enumerate}
@@
     \item \cdate[04/25] \KY Apply infinite mixture model to cluster topic model output
-    \item \cdate[05/05] \TBD Integrate topic modeling and cell clustering (MGCTM)
-    \item \cdate[05/05] \TBD Extend model to time-series data
+    \item \cdate[05/05] \CO Integrate topic modeling and cell clustering \\
+    This task will be planned and divided once the authors have a clearer idea of the most successful models from the earlier steps.
+Both authors are expected to deliberate on which models will be most appropriate to integrate based on the outcomes of previous steps in this project.
+Both authors will contribute to the implementation.
+More will be announced in the progress report.
+    \item \cdate[05/05] \CO Extend model to time-series data \\
+    Similar to the above, this task will be planned and divided once the authors have a clearer idea of the most successful models from the earlier steps.
+Both authors are expected to deliberate on which models will be most appropriate to extend to time-series data based on the outcomes of previous steps in this project.
+Both authors will contribute to the implementation.
+More will be announced in the progress report.
+
 \end{enumerate}
@@
         \url{https://arxiv.org/pdf/1309.6874.pdf}
+
+    \bibitem{LDA}
+        Latent Dirichlet Allocation.
+        \url{http://www.jmlr.org/papers/volume3/blei03a/blei03a.pdf}
+
+    \bibitem{HDP}
+        Hierarchical Dirichlet Processes
+        \url{http://people.eecs.berkeley.edu/~jordan/papers/hdp.pdf}
+
+    \bibitem{IBP}
+        The IBP Compound Dirichlet Process and its Application to Focused Topic Modeling
+        \url{http://www.cs.columbia.edu/~blei/papers/WilliamsonWangHellerBlei2010.pdf}
+
+    \bibitem{TS-DDP}
+        A Time-Series DDP for Functional Proteomics Profiles
+        \url{https://www.ma.utexas.edu/users/pmueller/pap/NM12.pdf}
+
 \end{thebibliography}

### 2017-04-09 hbp.py
add lda

@@ -1,4 +1,0 @@
-from preprocess_dataset import load_dataset
-
-data, genes, cells = load_dataset()
-print("Data shape:", data.shape)

### 2017-04-09 preprocess_dataset.py

@@ -1,74 +1,0 @@
-
-# coding: utf-8
-
-# In[1]:
-
-import pandas as pd
-import os.path
-import sys
-import numpy as np
-
-
-# In[2]:
-
-def load_dataset(preprocessed_path='dataset/preprocessed/human_melanoma_data.npy',
-                 raw_path='dataset/human_melanoma_data.tsv'):
-
-    if os.path.isfile(preprocessed_path):
-        return np.load(preprocessed_path),             np.load(preprocessed_path.split('.')[0] + "_column_lbls.npy"),             np.load(preprocessed_path.split('.')[0] + "_row_lbls.npy")
-
-    elif os.path.isfile(raw_path):
-        expr_data = pd.read_csv(raw_path, delimiter='\t').T
-        print("Number of cells:", expr_data.shape[0])
-        print("Number of genes:", expr_data.shape[1])
-
-        # excise stop genes
-        STOP_GENES = ['ACTB', 'B2M', 'EEF1A1', 'MTRNR2L1', 'MTRNR2L2',
-                  'MTRNR2L8', 'RPL41', 'RPLP1', 'RPS18', 'RPS27', 'TMSB4X']
-        expr_data = expr_data.drop(STOP_GENES, 1)
-        print("After cutting out stop genes:")
-        print("\tNumber of cells:", expr_data.shape[0])
-        print("\tNumber of genes:", expr_data.shape[1])
-
-        genes = list(expr_data.columns.values)
-        ribosomal_genes = [gene for gene in genes if gene.startswith('RPL')]
-        mito_ribosomal_genes = [gene for gene in genes if gene.startswith('MRPL')]
-        excise_genes = ribosomal_genes + mito_ribosomal_genes
-        expr_data = expr_data.drop(excise_genes, 1)
-        print("After cutting out ribosomal protein genes:")
-        print("\tNumber of cells:", expr_data.shape[0])
-        print("\tNumber of genes:", expr_data.shape[1])
-
-        # excise zero cells
-        expr_data = expr_data.loc[~(expr_data == 0).all(axis=1)]
-        print("After cutting out all-zero rows [cells]:")
-        print("\tNumber of cells:", expr_data.shape[0])
-        print("\tNumber of genes:", expr_data.shape[1])
-
-        # excise zero genes
-        expr_data = expr_data.loc[:, ~(expr_data == 0).all(axis=0)]
-        print("After cutting out all-zero columns [genes]:")
[…]

### 2017-04-11 progress/update.tex
add update, small mdl

--- 2017-04-04 proposal/proposal.tex
@@
 
-Tumors are composed of different sub-populations of cells, and these sub-populations exhibit shared patterns of gene expression.
-With contemporary sequencing machines, it is possible to obtain the expression levels of 10,000+ genes for 1000+ cells in a single experiment.
-In this project, we aim to apply Bayesian methods for clustering and topic modeling to discover meaningful cell clusters and gene modules from single-cell RNA-sequencing (scRNA-seq) datasets from tumors.
-We first intend to explore the use of a hierarchical topic model, Latent Dirichlet Allocation (LDA) \cite{LDA}, as well as two non-parametric topic models, Hierarchical Dirichlet Processes (HDP) \cite{HDP} and the Indian Buffet Process Compound Dirichlet Process (IBP-CDP) \cite{IBP} for analyzing scRNA-seq data.
-We will evaluate these models based on their ability to discover meaningful functional gene modules.
-Subsequently, we plan to explore the use of mixture models to cluster cells based on the output of the topic models, which we will evaluate based on their ability to differentiate cell types (e.g.
-immune, cancer, non-cancerous).
-Finally, we will train a combined clustering-topic model (e.g.
-MGCTM \cite{pengtao}) to see if it outperforms the individual models.
-We will compare the results of these methods with standard methods used in prior work, and evaluate these methods' robustness to noise.
-Time permitting, we will study extensions of these models appropriate for time-series scRNA-seq data \cite{TS-DDP}.
-\\
-
-Prior research has applied basic measures of statistical distance to cluster cell groups or gene modules based on samples of single-cell RNA-sequencing (scRNA-seq) data taken from tumors \cite{nature}.
-Good clustering of scRNA-seq data often has biologically meaningful results, and as such, a wide variety of correlation based methods have been used in prior work to derive meaning from scRNA-seq data \cite{coexpression, consensus, profiling}.
-We believe that Bayesian methods for clustering and topic modeling can be more illuminating than these previous methods, due to their consideration of higher-order relationships within the data.
-\\
-
-In summary, and in order of importance, we are primarily interested in answering the following questions:
+The progress report (one per team) will be a written pdf document, about 2 pages long,
+reporting on what has been accomplished so far and outlining the work that remains to be done.
+You should write clearly about how you’ve spent your time since the proposal and any preliminary
+findings (positive or negative). If there are multiple participants, the division of time spent
+so far should be made clear. Then you should include a plan (updated and/or revised from the
+proposal, as appropriate) for the time remaining. If there are multiple participants, the division of
+responsibility going forward should be made clear.
 

### 2017-04-11 preprocess_dataset.py

@@ -1,0 +1,80 @@
+
+# coding: utf-8
+
+# In[1]:
+
+import pandas as pd
+import os.path
+import sys
+import numpy as np
+
+
+# In[4]:
+
+def cut_out_zero_rows(data):
+    return data.loc[~(data == 0).all(axis=1)]
+
+def cut_out_zero_cols(data):
+    return data.loc[:, ~(data == 0).all(axis=0)]
+
+def load_dataset(preprocessed_path='dataset/preprocessed/human_melanoma_data.npy',
+                 raw_path='dataset/human_melanoma_data.tsv'):
+
+    if os.path.isfile(preprocessed_path):
+        return np.load(preprocessed_path),             np.load(preprocessed_path.split('.')[0] + "_column_lbls.npy"),             np.load(preprocessed_path.split('.')[0] + "_row_lbls.npy")
+
+    elif os.path.isfile(raw_path):
+        expr_data = pd.read_csv(raw_path, delimiter='\t').T
+        print("Number of cells:", expr_data.shape[0])
+        print("Number of genes:", expr_data.shape[1])
+
+        # excise stop genes
+        STOP_GENES = ['ACTB', 'B2M', 'EEF1A1', 'MTRNR2L1', 'MTRNR2L2',
+                  'MTRNR2L8', 'RPL41', 'RPLP1', 'RPS18', 'RPS27', 'TMSB4X']
+        expr_data = expr_data.drop(STOP_GENES, 1)
+        print("After cutting out stop genes:")
+        print("\tNumber of cells:", expr_data.shape[0])
+        print("\tNumber of genes:", expr_data.shape[1])
+
+        genes = list(expr_data.columns.values)
+        ribosomal_genes = [gene for gene in genes if gene.startswith('RPL')]
+        mito_ribosomal_genes = [gene for gene in genes if gene.startswith('MRPL')]
+        excise_genes = ribosomal_genes + mito_ribosomal_genes
+        expr_data = expr_data.drop(excise_genes, 1)
+        print("After cutting out ribosomal protein genes:")
+        print("\tNumber of cells:", expr_data.shape[0])
+        print("\tNumber of genes:", expr_data.shape[1])
+
+        # excise zero cells
+        expr_data = cut_out_zero_rows(expr_data)
+        print("After cutting out all-zero rows [cells]:")
+        print("\tNumber of cells:", expr_data.shape[0])
+        print("\tNumber of genes:", expr_data.shape[1])
+
+        # excise zero genes
[…]

### 2017-04-11 progress/build.sh

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+file="update"
+author1="karren"
+author2="skoppula"
+totalfile="$file-$author1-$author2"
+
+# Compile the tex file
+pdflatex --shell-escape -jobname=$totalfile ${file}.tex
+pdflatex --shell-escape -jobname=$totalfile ${file}.tex
+
+if [ "$(uname)" == "Darwin" ]; then
+    open ${totalfile}.pdf &
+elif [ "$(expr substr $(uname -s) 1 5)" == "Linux" ]; then
+    evince ${totalfile}.pdf &
+elif [ "$(expr substr $(uname -s) 1 10)" == "MINGW32_NT" ]; then
+    echo 'Windows not supported!'
+fi
+
+rm *.aux *.log *.out

### 2017-04-11 IBP-DP/IBPCompoundDirichlet.py
added IBP-DP model based on Williamson paper (2010)

@@ -1,0 +1,428 @@
+import sampyl as smp
+import autograd.numpy as np
+import autograd.scipy.special as scis
+import csv
+# testing
+np.seterr(all='raise')
+
+class IBPCompoundDirichlet(object):
+
+	def __init__(self):
+		pass
+		
+	
+	def get_topics(self):
+		self.beta = []		
+		for k in range(self.n_topics):
+			beta_k = []			
+			for v in range(self.n_vocab):
+				beta_k.append(np.count_nonzero(self.WS[self.ZS == k] == v))
+			try:
+				beta_k = np.array(beta_k)/sum(beta_k)
+			except FloatingPointError:
+				beta_k = np.array(beta_k)
+			self.beta.append(beta_k)
+		self.beta = np.array(self.beta)
+		print(self.beta)
+
+	def get_topic_assignments(self):
+		self.theta = []
+		for m in range(self.n_docs):
+			theta_m = []
+			for k in range(self.n_topics):
+				theta_m.append(np.count_nonzero(self.ZS[self.DS == m] == k))
+			theta_m = np.array(theta_m)/sum(theta_m)
+			self.theta.append(theta_m)
+		self.theta = np.array(self.theta)
+		print(self.theta)
+			
+	
+	
+	def fit_data(self, data, n_iter, gamma, alpha, eta):
+		# number of iterations
+		self.n_iter = n_iter
+
+		# Set hyperparameters
+		self.gamma = gamma
+		self.alpha = alpha
+		self.eta = eta
+
+		# initialize all latent variables
+		self._initialize(data)
+
+		# do posterior inference
+		self._posterior_inference()
+
+
+	def _initialize(self, data):
+    
+	    # FROM LDA UTIL - for converting word-document matrix into list of words and list of document indices
+	    # Rows are documents, columns are vocabulary
+	    def _matrix_to_lists(doc_word): 
+	        ii, jj = np.nonzero(doc_word)
+	        ss = tuple(doc_word[i, j] for i, j in zip(ii, jj))
+	        n_tokens = int(doc_word.sum())
+	        DS = np.repeat(ii, ss).astype(np.int)
+	        WS = np.empty(n_tokens, dtype=np.int)
+	        startidx = 0
+	        for i, cnt in enumerate(ss):
+	            WS[startidx:startidx + cnt] = jj[i]
+	            startidx += cnt
[…]

### 2017-04-12 IBP-DP/IBPCompoundDirichlet.py
added simulation files, fixed bugs

@@ -1,6 +1,5 @@
-import sampyl as smp
-import autograd.numpy as np
-import autograd.scipy.special as scis
+import numpy as np
+import scipy.special as scis
 import csv
 # testing
 np.seterr(all='raise')
@@ -10,7 +9,32 @@ class IBPCompoundDirichlet(object):
 	def __init__(self):
 		pass
 		
-	
+	def write_to_file(self, file_root):
+		self.get_topics()
+		self.get_topic_assignments()
+
+		topics_file = file_root + "_topics"
+		assignments_file = file_root + "_assignments"
+		pi_phi_file = file_root + "_pi_phi"
+
+		with open(topics_file, 'w') as csvfile:
+			writer = csv.writer(csvfile, delimiter = ' ')
+			for i in range(len(self.beta)):
+				writer.writerow(self.beta[i])
+
+		with open(assignments_file, 'w') as csvfile:
+			writer = csv.writer(csvfile, delimiter = ' ')
+			for i in range(len(self.theta)):
+				writer.writerow(self.theta[i])
+
+		with open(pi_phi_file, 'w') as csvfile:
+			writer = csv.writer(csvfile, delimiter = ' ')
+			writer.writerow(self.pi)
+			writer.writerow(self.phi)
+
+
+
+
 	def get_topics(self):
 		self.beta = []		
 		for k in range(self.n_topics):
@@ -23,7 +47,9 @@ class IBPCompoundDirichlet(object):
 				beta_k = np.array(beta_k)
 			self.beta.append(beta_k)
 		self.beta = np.array(self.beta)
-		print(self.beta)
+
+
+
 
 	def get_topic_assignments(self):
 		self.theta = []
@@ -31,11 +57,14 @@ class IBPCompoundDirichlet(object):
 			theta_m = []
 			for k in range(self.n_topics):
 				theta_m.append(np.count_nonzero(self.ZS[self.DS == m] == k))
-			theta_m = np.array(theta_m)/sum(theta_m)
+			try:
+				theta_m = np.array(theta_m)/sum(theta_m)
+			except FloatingPointError:
+				theta_m = np.array(theta_m)
 			self.theta.append(theta_m)
 		self.theta = np.array(self.theta)
-		print(self.theta)
-			
+
+
 	
 	
 	def fit_data(self, data, n_iter, gamma, alpha, eta):
@@ -73,8 +102,11 @@ class IBPCompoundDirichlet(object):
[…]

### 2017-04-12 IBP-DP/IBP_Reuters_test.py

@@ -1,0 +1,24 @@
+import numpy as np
+import lda.datasets
+
+X = lda.datasets.load_reuters()
+vocab = lda.datasets.load_reuters_vocab()
+titles = lda.datasets.load_reuters_titles()
+print(X.shape)
+
+data = X[0:50,0:200]
+
+
+import IBPCompoundDirichlet
+import time
+
+t_start = time.time()
+
+model = IBPCompoundDirichlet.IBPCompoundDirichlet()
+model.fit_data(data, 500, 5, 1, 0.1)
+
+t_end = time.time()
+
+print(t_end-t_start)
+
+model.write_to_file('reuters_100docs_2000words_results')

### 2017-04-17 lda.py
add everything

@@ -1,0 +1,192 @@
+
+# coding: utf-8
+
+# In[16]:
+
+from preprocess_dataset import load_dataset
+from preprocess_dataset import cut_out_zero_rows
+from preprocess_dataset import cut_out_zero_cols
+import pandas as pd
+import numpy as np
+import pickle
+from sklearn.model_selection import train_test_split
+from sklearn.decomposition import LatentDirichletAllocation
+from sklearn.externals import joblib
+import os
+import errno
+import time
+
+
+# In[17]:
+
+def mkdir(path):
+    try:
+        os.makedirs(path)
+    except OSError as exception:
+        if exception.errno != errno.EEXIST:
+            raise
+
+def compute_stats(data):
+    tmp = np.count_nonzero(data, axis=1)
+    print("Average/Min/Max non-zero count for each cell: ", np.mean(tmp), np.min(tmp), np.max(tmp))
+    tmp = np.count_nonzero(data, axis=0)
+    print("Average/Min/Max non-zero count for each gene: ", np.mean(tmp), np.min(tmp), np.max(tmp))
+    if np.max(tmp) == data.shape[0]: print("there exists a gene active in all cells!")
+    tmp = np.max(data, axis=0)
+    print("Average/Min/Max max-expression value for each gene: ", np.mean(tmp), np.min(tmp), np.max(tmp))
+    print("Number of genes with max-expression of 1:", np.where(tmp==1)[0].shape[0])
+    # mean non-zero expression value of each gene
+    tmp = np.true_divide(data.sum(0),(data!=0).sum(0))
+    print("Average/Min/Max mean non-zero expression value for each gene: ", np.mean(tmp), np.min(tmp), np.max(tmp))
+    
+def normalize_along_columns(data):
+    return (data - data.mean(axis=0)) / data.std(axis=0)
+
+def split_test(data, test_size=0.1):
+    data_trn,data_te,_,_ = train_test_split(data, np.ones(data.shape[0]), test_size=test_size, random_state=42)
+    return data_trn,data_te
+
+def gene_dropout(data_trn,data_te,dropout=0.1):
+    column_dropout = 0.1
+    rand_idx = np.random.randint(data_trn.shape[1], size=int(data_trn.shape[1]*column_dropout))
+    trimmed_data_trn = data_trn[:,rand_idx]
+    trimmed_data_te = data_te[:,rand_idx]
[…]

### 2017-04-17 preprocess_dataset.py

@@ -1,7 +1,7 @@
 
 # coding: utf-8
 
-# In[1]:
+# In[3]:
 
 import pandas as pd
 import os.path
@@ -9,7 +9,7 @@ import sys
 import numpy as np
 
 
-# In[4]:
+# In[26]:
 
 def cut_out_zero_rows(data):
     return data.loc[~(data == 0).all(axis=1)]
@@ -18,10 +18,11 @@ def cut_out_zero_cols(data):
     return data.loc[:, ~(data == 0).all(axis=0)]
 
 def load_dataset(preprocessed_path='dataset/preprocessed/human_melanoma_data.npy',
-                 raw_path='dataset/human_melanoma_data.tsv'):
+                 raw_path='dataset/human_melanoma_data.tsv', filter_std=None, load=True):
 
-    if os.path.isfile(preprocessed_path):
-        return np.load(preprocessed_path),             np.load(preprocessed_path.split('.')[0] + "_column_lbls.npy"),             np.load(preprocessed_path.split('.')[0] + "_row_lbls.npy")
+    preprocessed_data_path = preprocessed_path.split('.')[0] + '_' + str(filter_std) + '.npy'
+    if os.path.isfile(preprocessed_data_path) and load:
+        return np.load(preprocessed_data_path),  np.load(preprocessed_path.split('.')[0] + "_column_lbls.npy"), np.load(preprocessed_path.split('.')[0] + "_row_lbls.npy")
 
     elif os.path.isfile(raw_path):
         expr_data = pd.read_csv(raw_path, delimiter='\t').T
@@ -56,13 +57,22 @@ def load_dataset(preprocessed_path='dataset/preprocessed/human_melanoma_data.npy
         print("After cutting out all-zero columns [genes]:")
         print("\tNumber of cells:", expr_data.shape[0])
         print("\tNumber of genes:", expr_data.shape[1])
-
+        
+        # remove low STD cols (genes)
+        stds = np.std(expr_data, axis=0)
+        keep_idxs = stds.argsort()[int(len(stds)*filter_std):]
+        threshold = np.sort(stds)[int(len(stds)*filter_std)]
+        expr_data = np.take(expr_data,keep_idxs, axis=1)
+        print("After cutting out the bottom %f fraction of genes with low STD (<%f):" % (filter_std, threshold))
+        print("\tNumber of cells:", expr_data.shape[0])
+        print("\tNumber of genes:", expr_data.shape[1])
+        
[…]

### 2017-04-17 update/update.tex
finish shit

%!TEX program = xelatex

\documentclass[11pt]{article}

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{titlesec}
\usepackage{wrapfig}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{sidecap}
\usepackage{clrscode}
\usepackage{xcolor}
\usepackage{float}
\usepackage{stackengine}

\usepackage[margin=0.75in]{geometry}

\newlength{\tindent}
\setlength{\tindent}{\parindent}
\setlength{\parindent}{0pt}
\renewcommand{\indent}{\hspace*{\tindent}}

\newcommand\tab[1][1cm]{\hspace*{#1}}
\newcommand{\SK}{\textbf{[SK]}\tab}
\newcommand{\KY}{\textbf{[KY]}\tab}
\newcommand{\CO}{\textbf{[CO]}\tab}
\newcommand\cdate[1][1cm]{\textbf{[#1]}}

\renewcommand{\thesubsection}{\thesection.\alph{subsection}}
\titleformat{\subsection}[runin]{\normalfont\large\bfseries}{\thesubsection}{1em}{}

\begin{document}

\begin{centering}
\Large
    \textbf{Bayesian Clustering and Topic Modeling of Gene Expression Data}  \\
    \vspace{2mm}
    \normalsize
    Skanda Koppula (\url{skoppula@mit.edu}), Karren Yang (\url{karren@mit.edu}) \\
    \vspace{2mm}
    \normalsize
    6.882 Project Proposal, \today \\
\end{centering}
\vspace{5mm}

\section{Overview}

We explored the usage of various topic and clustering models in the analysis of gene expression data. For the former, we explored whether topic modeling could identify biologically relevant `topics'. In this context, a `topic' is a set of genes that together perform a specific biological function (a `gene module'); we compared our results to modules found by biology literate. For the latter, we explored whether we are able to cluster cells
accurately based on their gene expression levels. \footnote{The genes in gene modules generally move in tandem: so when a gene module is upregulated, all genes have higher expression (i.e. frequency). This motivates our use of topic models to capture this relationship.}.

\section{Topic Model: Latent Dirichlet Allocation}

We were unable to yield any results using Python's out-of-the-box implementation of \texttt{lda} \cite{lda}. With our 250 MB dataset, the collapsed Gibbs sampler used in the implementation was taking too long to produce samples, even on larger server-grade machines and with a small number of LDA clusters. We explored modifying the source to parallelize the sampling, but found the source to be crabbed and hard to follow. \\

In our search for an alternative, we explored another implementation that used online variational bayes for posterior estimate \cite{ovb, online}, and a broken \texttt{C++} Gibbs sampler for LDA that we modified to work \cite{plda}. We were able to get this latter sampler running parallel across multiple cores, with a burn-in of 100 iterations. We compare these two posterior estimation approaches using our entire dataset, with a 10\% held-out testing partition. We experimented with $k=5,10,25,\text{ and } 50$ clusters.

Figure \ref{fig:time} in the Appendix shows the time to complete each estimation method. As expected, Gibbs scales poorly with the parameter dimensionality and is strictly worse than online variational Bayes across all studied topic counts.

Figure \ref{fig:time} in the Appendix shows the time to complete each estimation method. As expected, Gibbs scales poorly with the parameter dimensionality and is strictly worse than online variational Bayes across all studied topic counts.

Note that these $p$-values do \textit{not} factor for multiple hypothesis corrections, so could be misleadingly significant.

\section{Division of Work}

\section{Remaining Work and Schedule}
- robust to dropout
- time series

\section{Appendix}
\begin{figure}
    \label{fig:time}
    \centering
    \includegraphics[width=0.5\textwidth]{time}
    \caption{Comparison of the running times of each of the posterior estimation methods across various cluster sizes.}
\end{figure}

\begin{figure}
    \label{fig:ll}
    \centering
    \includegraphics[width=0.5\textwidth]{ll}
    \caption{Log-likelihood of the held-out testing set, across various cluster sizes.}
\end{figure}

\begin{thebibliography}{9}
    \bibitem{nature} 
        The XL-mHG Test For Enrichment: A Technical Report.
        \url{https://arxiv.org/pdf/1507.07905.pdf}

    \bibitem{msigdb} 
        Molecular Signatures Database v6.0.
        \url{http://software.broadinstitute.org/gsea/msigdb}

    \bibitem{lda} 
        lda: Topic modeling with latent Dirichlet Allocation.
        \url{http://pythonhosted.org/lda/}

    \bibitem{ovb} 
        Online Latent Dirichlet Allocation with variational inference.
        \url{https://github.com/scikit-learn/scikit-learn/blob/master/sklearn/decomposition/online_lda.py}

    \bibitem{plda} 
        C++ implementation of Latent Dirichlet Allocation
        \url{https://github.com/openbigdatagroup/plda/blob/master/lda.cc}

    \bibitem{online}
        Online Learning for Latent Dirichlet Allocation.
        \url{https://pdfs.semanticscholar.org/157a/ef34d39c85d6576028f29df1ea4c6480a979.pdf}

    \bibitem{HDP}
        Hierarchical Dirichlet Processes
        \url{http://people.eecs.berkeley.edu/~jordan/papers/hdp.pdf}

    \bibitem{IBP}
        The IBP Compound Dirichlet Process and its Application to Focused Topic Modeling
        \url{http://www.cs.columbia.edu/~blei/papers/WilliamsonWangHellerBlei2010.pdf}

    \bibitem{TS-DDP}
        A Time-Series DDP for Functional Proteomics Profiles
        \url{https://www.ma.utexas.edu/users/pmueller/pap/NM12.pdf}

\end{thebibliography}

\end{document}

### 2017-04-17 enrichment_scoring/enrichment_mHGtest.py

@@ -5,67 +5,67 @@ import matplotlib.pyplot as plt
 import xlmhg
 
 def enrichment_mHGtest(vocab_file_name, ranked_genes_file_name, save_file_name = 'enrichment', n_sig_pathways = 6, enrichment_minimum_n = 5, lowest_cutoff = 100):
-	vocab = np.load(vocab_file_name)
+    vocab = np.load(vocab_file_name)
 
-	gene_sets = []
-	gene_set_sizes = []
-	gene_set_names = []
-	
-	print('Loading gene sets...(can take a few minutes)')
-	# load gene sets from msigdb file
-	with open('c2.all.v6.0.symbols.gmt', 'r') as file:
-		reader = csv.reader(file, delimiter = '\t')
-		for idx, row in enumerate(reader):
-			gene_set_names.append(row[0])
-			genes = np.array(row[2:])
-			genes = genes[np.in1d(genes,vocab)]
-			gene_sets.append(genes)
-			gene_set_sizes.append(len(genes))
-	print('Done')
-	gene_set_names = np.array(gene_set_names)
-	gene_set_sizes = np.array(gene_set_sizes)
-	gene_subsize = gene_set_sizes[gene_set_sizes>10]
+    gene_sets = []
+    gene_set_sizes = []
+    gene_set_names = []
 
-	idx = np.argwhere(gene_set_sizes>10).squeeze()
+    print('Loading gene sets...(can take a few minutes)')
+    # load gene sets from msigdb file
+    with open('c2.all.v6.0.symbols.gmt', 'r') as file:
+        reader = csv.reader(file, delimiter = '\t')
+        for idx, row in enumerate(reader):
+            gene_set_names.append(row[0])
+            genes = np.array(row[2:])
+            genes = genes[np.in1d(genes,vocab)]
+            gene_sets.append(genes)
+            gene_set_sizes.append(len(genes))
+    print('Done')
+    gene_set_names = np.array(gene_set_names)
+    gene_set_sizes = np.array(gene_set_sizes)
+    gene_subsize = gene_set_sizes[gene_set_sizes>10]
 
-	names = gene_set_names[idx]
-	sets = [gene_sets[i] for i in idx]
-	sizes = gene_set_sizes[idx]
-	
-	ranked_gene_lists = np.load(ranked_genes_file_name)
+    idx = np.argwhere(gene_set_sizes>10).squeeze()
 
-	all_pvals = []
-	all_pathways = []
-	
-	print('Calculating enrichment scores')
[…]

### 2017-04-17 lda.py

@@ -1,192 +1,0 @@
-
-# coding: utf-8
-
-# In[16]:
-
-from preprocess_dataset import load_dataset
-from preprocess_dataset import cut_out_zero_rows
-from preprocess_dataset import cut_out_zero_cols
-import pandas as pd
-import numpy as np
-import pickle
-from sklearn.model_selection import train_test_split
-from sklearn.decomposition import LatentDirichletAllocation
-from sklearn.externals import joblib
-import os
-import errno
-import time
-
-
-# In[17]:
-
-def mkdir(path):
-    try:
-        os.makedirs(path)
-    except OSError as exception:
-        if exception.errno != errno.EEXIST:
-            raise
-
-def compute_stats(data):
-    tmp = np.count_nonzero(data, axis=1)
-    print("Average/Min/Max non-zero count for each cell: ", np.mean(tmp), np.min(tmp), np.max(tmp))
-    tmp = np.count_nonzero(data, axis=0)
-    print("Average/Min/Max non-zero count for each gene: ", np.mean(tmp), np.min(tmp), np.max(tmp))
-    if np.max(tmp) == data.shape[0]: print("there exists a gene active in all cells!")
-    tmp = np.max(data, axis=0)
-    print("Average/Min/Max max-expression value for each gene: ", np.mean(tmp), np.min(tmp), np.max(tmp))
-    print("Number of genes with max-expression of 1:", np.where(tmp==1)[0].shape[0])
-    # mean non-zero expression value of each gene
-    tmp = np.true_divide(data.sum(0),(data!=0).sum(0))
-    print("Average/Min/Max mean non-zero expression value for each gene: ", np.mean(tmp), np.min(tmp), np.max(tmp))
-    
-def normalize_along_columns(data):
-    return (data - data.mean(axis=0)) / data.std(axis=0)
-
-def split_test(data, test_size=0.1):
-    data_trn,data_te,_,_ = train_test_split(data, np.ones(data.shape[0]), test_size=test_size, random_state=42)
-    return data_trn,data_te
-
-def gene_dropout(data_trn,data_te,dropout=0.1):
-    column_dropout = 0.1
-    rand_idx = np.random.randint(data_trn.shape[1], size=int(data_trn.shape[1]*column_dropout))
-    trimmed_data_trn = data_trn[:,rand_idx]
-    trimmed_data_te = data_te[:,rand_idx]
[…]

### 2017-04-17 lda/preprocess_dataset.py

@@ -1,0 +1,102 @@
+
+# coding: utf-8
+
+# In[3]:
+
+import pandas as pd
+import os.path
+import sys
+import numpy as np
+
+
+# In[26]:
+
+def cut_out_zero_rows(data):
+    return data.loc[~(data == 0).all(axis=1)]
+
+def cut_out_zero_cols(data):
+    return data.loc[:, ~(data == 0).all(axis=0)]
+
+def load_dataset(preprocessed_path='dataset/preprocessed/human_melanoma_data.npy',
+                 raw_path='dataset/human_melanoma_data.tsv', filter_std=None, load=True):
+
+    preprocessed_data_path = preprocessed_path.split('.')[0] + '_' + str(filter_std) + '.npy'
+    if os.path.isfile(preprocessed_data_path) and load:
+        return np.load(preprocessed_data_path),  np.load(preprocessed_path.split('.')[0] + "_column_lbls.npy"), np.load(preprocessed_path.split('.')[0] + "_row_lbls.npy")
+
+    elif os.path.isfile(raw_path):
+        expr_data = pd.read_csv(raw_path, delimiter='\t').T
+        print("Number of cells:", expr_data.shape[0])
+        print("Number of genes:", expr_data.shape[1])
+
+        # excise stop genes
+        STOP_GENES = ['ACTB', 'B2M', 'EEF1A1', 'MTRNR2L1', 'MTRNR2L2',
+                  'MTRNR2L8', 'RPL41', 'RPLP1', 'RPS18', 'RPS27', 'TMSB4X']
+        expr_data = expr_data.drop(STOP_GENES, 1)
+        print("After cutting out stop genes:")
+        print("\tNumber of cells:", expr_data.shape[0])
+        print("\tNumber of genes:", expr_data.shape[1])
+
+        genes = list(expr_data.columns.values)
+        ribosomal_genes = [gene for gene in genes if gene.startswith('RPL')]
+        mito_ribosomal_genes = [gene for gene in genes if gene.startswith('MRPL')]
+        excise_genes = ribosomal_genes + mito_ribosomal_genes
+        expr_data = expr_data.drop(excise_genes, 1)
+        print("After cutting out ribosomal protein genes:")
+        print("\tNumber of cells:", expr_data.shape[0])
+        print("\tNumber of genes:", expr_data.shape[1])
+
+        # excise zero cells
+        expr_data = cut_out_zero_rows(expr_data)
+        print("After cutting out all-zero rows [cells]:")
[…]

### 2017-04-17 preprocess_dataset.py

@@ -1,102 +1,0 @@
-
-# coding: utf-8
-
-# In[3]:
-
-import pandas as pd
-import os.path
-import sys
-import numpy as np
-
-
-# In[26]:
-
-def cut_out_zero_rows(data):
-    return data.loc[~(data == 0).all(axis=1)]
-
-def cut_out_zero_cols(data):
-    return data.loc[:, ~(data == 0).all(axis=0)]
-
-def load_dataset(preprocessed_path='dataset/preprocessed/human_melanoma_data.npy',
-                 raw_path='dataset/human_melanoma_data.tsv', filter_std=None, load=True):
-
-    preprocessed_data_path = preprocessed_path.split('.')[0] + '_' + str(filter_std) + '.npy'
-    if os.path.isfile(preprocessed_data_path) and load:
-        return np.load(preprocessed_data_path),  np.load(preprocessed_path.split('.')[0] + "_column_lbls.npy"), np.load(preprocessed_path.split('.')[0] + "_row_lbls.npy")
-
-    elif os.path.isfile(raw_path):
-        expr_data = pd.read_csv(raw_path, delimiter='\t').T
-        print("Number of cells:", expr_data.shape[0])
-        print("Number of genes:", expr_data.shape[1])
-
-        # excise stop genes
-        STOP_GENES = ['ACTB', 'B2M', 'EEF1A1', 'MTRNR2L1', 'MTRNR2L2',
-                  'MTRNR2L8', 'RPL41', 'RPLP1', 'RPS18', 'RPS27', 'TMSB4X']
-        expr_data = expr_data.drop(STOP_GENES, 1)
-        print("After cutting out stop genes:")
-        print("\tNumber of cells:", expr_data.shape[0])
-        print("\tNumber of genes:", expr_data.shape[1])
-
-        genes = list(expr_data.columns.values)
-        ribosomal_genes = [gene for gene in genes if gene.startswith('RPL')]
-        mito_ribosomal_genes = [gene for gene in genes if gene.startswith('MRPL')]
-        excise_genes = ribosomal_genes + mito_ribosomal_genes
-        expr_data = expr_data.drop(excise_genes, 1)
-        print("After cutting out ribosomal protein genes:")
-        print("\tNumber of cells:", expr_data.shape[0])
-        print("\tNumber of genes:", expr_data.shape[1])
-
-        # excise zero cells
-        expr_data = cut_out_zero_rows(expr_data)
-        print("After cutting out all-zero rows [cells]:")
[…]

### 2017-04-17 progress/build.sh

@@ -1,20 +1,0 @@
-#!/bin/bash
-
-file="update"
-author1="karren"
-author2="skoppula"
-totalfile="$file-$author1-$author2"
-
-# Compile the tex file
-pdflatex --shell-escape -jobname=$totalfile ${file}.tex
-pdflatex --shell-escape -jobname=$totalfile ${file}.tex
-
-if [ "$(uname)" == "Darwin" ]; then
-    open ${totalfile}.pdf &
-elif [ "$(expr substr $(uname -s) 1 5)" == "Linux" ]; then
-    evince ${totalfile}.pdf &
-elif [ "$(expr substr $(uname -s) 1 10)" == "MINGW32_NT" ]; then
-    echo 'Windows not supported!'
-fi
-
-rm *.aux *.log *.out

### 2017-04-17 update/update.tex
finish everything

@@
 \renewcommand{\indent}{\hspace*{\tindent}}
+\floatplacement{figure}{H}
 
@@
 
-In our search for an alternative, we explored another implementation that used online variational bayes for posterior estimate \cite{ovb, online}, and a broken \texttt{C++} Gibbs sampler for LDA that we modified to work \cite{plda}.
+In our search for an alternative, we explored another implementation that used online mean-field variational bayes for posterior estimate \cite{ovb, online}, and a broken \texttt{C++} Gibbs sampler for LDA that we modified to work \cite{plda}.
 We were able to get this latter sampler running parallel across multiple cores, with a burn-in of 100 iterations.
 We compare these two posterior estimation approaches using our entire dataset, with a 10\% held-out testing partition.
-We experimented with $k=5,10,25,\text{ and } 50$ clusters.
+We experimented with
+$k=5,10,25,\text{ and } 50$ clusters. \\
 
@@
 As expected, Gibbs scales poorly with the parameter dimensionality and is strictly worse than online variational Bayes across all studied topic counts.
+\\
 
-Figure \ref{fig:time} in the Appendix shows the time to complete each estimation method.
-As expected, Gibbs scales poorly with the parameter dimensionality and is strictly worse than online variational Bayes across all studied topic counts.
+We had time to calculate perplexity and log-likelihood for the OVB parameters, and very surprisingly, we find an increasing log-likelihood for larger cluster sizes on a held-out test set (Figure \ref{fig:ll}).
+We are currently trying to understand whether this is because of programming error or a legitimate result.
+\\
 
-Note that these $p$-values do \textit{not} factor for multiple hypothesis corrections, so could be misleadingly significant.
+As is standard in computational gene module method validation, we use a hypergeometric test to compare our topics' collections of genes, with existing collections of genes catalogued in the comprehensive gene module database, \texttt{MSigDB} \cite{msigdb}.
+Figure \ref{fig:pathways} in the Appendix shows topic-to-\texttt{MSigDB} module matches for which the $p$-value of the match is at least less than 0.3.
+Models with larger clusters tend to have more matches with higher significance.
+There
+a very slightly higher number of matches using the topics discovered by parallelized Gibbs.
+There are repeat matches, which is slightly disheartening.
+We note however that these $p$-values do \textit{not} factor for multiple hypothesis corrections (yet), so could be misleadingly significant!
 
-\section{Division of Work}
+\section{Clustering Model: Finite-Sized Mixture Model}
+
+We are currently implementing from scratch the mixture model specified by the plate model in Figure \ref{fig:plate}.
+The underlying assumption of behind the model is that the cell cluster assignment, $z_i$, would be indicative of it's gene expression values as well.
+Our implementation uses \texttt{numpy}, and we are exploring backends to make posterior inference computationally feasible (our current implementation, which uses a Gibbs sampler, overflows memory, so we are
+implementing stochastic variational methods, like used in our LDA experiments).
 
 \section{Remaining Work and Schedule}
-- robust to dropout
-- time series
+We are making progress as per the schedule outlined in our proposal.
+Apart from continuing with work on the finite-sized mixture model, we plan on running a few experiments to test robustness of our methods to dropout, and scaling IBP and HDP to the full dataset.
 
@@
 \begin{figure}
-    \label{fig:time}
     \centering
@@
     \caption{Comparison of the running times of each of the posterior estimation methods across various cluster sizes.}
+    \label{fig:time}
 \end{figure}
@@
 \begin{figure}
-    \label{fig:ll}
     \centering
@@
     \caption{Log-likelihood of the held-out testing set, across various cluster sizes.}
+    \label{fig:ll}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1\textwidth]{pathways}
+    \caption{Matches between the gene collections found in LDA topics and published gene sets in \texttt{MSigDB}.}
+    \label{fig:pathways}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.3\textwidth]{mixture}
+    \caption{Finite \texttt{K}-sized mixture model currently implemented.
+$\theta$ is the parameter for every cluster component, represented from a categorical draw of over all genes.
+$z_i$ is the cluster assignment, and $\phi$ is the distribution of clusters.
+As usual, $\alpha, \beta$ are hyper-parameters.}
+    \label{fig:plate}
 \end{figure}
@@
 \begin{thebibliography}{9}
-    \bibitem{nature} 
+    \bibitem{hg} 
         The XL-mHG Test For Enrichment: A Technical Report.

### 2017-04-18 update/update.tex
Completed progress report

--- 2017-04-17 update/update.tex
@@
 
-We explored the usage of various topic and clustering models in the analysis of gene expression data.
-For the former, we explored whether topic modeling could identify biologically relevant `topics'.
-In this context, a `topic' is a set of genes that together perform a specific biological function (a `gene module'); we compared our results to modules found by biology literate.
-For the latter, we explored whether we are able to cluster cells
-accurately based on their gene expression levels.
-\footnote{The genes in gene modules generally move in tandem: so when a gene module is upregulated, all genes have higher expression (i.e.
-frequency).
-This motivates our use of topic models to capture this relationship.}.
+We have made preliminary assessments of three topic models -- LDA, the Hierarchical Dirichlet Process (HDP), and the Indian Buffet Process Compound Dirichlet Process (IDP-DP) -- for discovering topics in single-cell RNA-sequencing (scRNA-seq) data.
+In this context, topics are sets of genes that are co-expressed across cells (i.e.
+gene modules) \footnote{The genes in gene modules generally move in tandem: so when a gene module is upregulated, all genes have higher expression.
+This motivates our use of topic models to capture these relationships.}.
+Since we would like topics to correpond to biologically useful gene modules, our main evaluation metric is the correpondence of the learned topics to known functional gene sets curated from literature (i.e.
+from the \texttt{MSigDB} database).
+While HDP and the IBP-DP offer greater flexibility in that the number of topics does not have to be defined a priori, we do not find evidence that they yield significantly more coherent topics; moreover, we find in practice that training the models and tuning the hyper-parameters to maintain a reasonable number of topics is less time-efficient than applying LDA over a range of number of topics.
+Moving forward, we will focus on implementing and evaluating a unified framework for clustering and topic modeling based on LDA.
 
-\section{Topic Model: Latent Dirichlet Allocation}
+\section{Results}
+\subsection*{Running LDA}
 
-We were unable to yield any results using Python's out-of-the-box implementation of \texttt{lda} \cite{lda}.
+We explored two implementations of posterior inference for LDA on large datasets: one using online mean-field variational bayes for posterior estimate \cite{ovb, online}, and a broken \texttt{C++} Gibbs sampler for LDA that we modified to work \cite{plda}.
+We were able to get this latter sampler running parallel across multiple cores, with a burn-in of 100 iterations.
+We compare these two posterior estimation approaches using our entire dataset, with a 10\% held-out testing partition.
+We experimented with
+$k=5, 10, 25, 50$ topics. \\
+
+\iffalse
+It was nececssary to use these implementations as we were unable to obtain any results using Python's out-of-the-box implementation of \texttt{lda} \cite{lda}.
 With our 250 MB dataset, the collapsed Gibbs sampler used in the implementation was taking too long to produce samples, even on larger server-grade machines and with a small number of LDA clusters.
@@
 \\
+\fi
 
-In our search for an alternative, we explored another implementation that used online variational bayes for posterior estimate \cite{ovb, online}, and a broken \texttt{C++} Gibbs sampler for LDA that we modified to work \cite{plda}.
-We were able to get this latter sampler running parallel across multiple cores, with a burn-in of 100 iterations.
-We compare these two posterior estimation approaches using our entire dataset, with a 10\% held-out testing partition.
-We experimented with $k=5,10,25,\text{ and } 50$ clusters.
+Figure \ref{fig:time} shows the time to complete each estimation method.
+As expected, Gibbs scales poorly with the parameter dimensionality and is strictly worse than online variational Bayes across all studied topic counts.
+We also calculated perplexity and log-likelihood for the OVB parameters, and very surprisingly, we found that log-likelihood decreased as number of topics increased on a held-out test set (Figure \ref{fig:ll}).
+We are currently trying to understand whether this is because of programming error.
 
-Figure \ref{fig:time} in the Appendix shows the time to complete each estimation method.
-As expected, Gibbs scales poorly with the parameter dimensionality and is strictly worse than online variational Bayes across all studied topic counts.
+\subsection*{Evaluating Biological Significance of LDA Topics}
 
-Figure \ref{fig:time} in the Appendix shows the time to complete each estimation method.
-As expected, Gibbs scales poorly with the parameter dimensionality and is strictly worse than online variational Bayes across all studied topic counts.
+We did gene set enrichment analysis using the minimum hypergeometric test \cite{hg} to compare our topics, which are ranked lists of genes, with existing collections of genes catalogued in the comprehensive gene module database, \texttt{MSigDB} \cite{msigdb}.
+Figure \ref{fig:pathways} in the Appendix shows topic-to-\texttt{MSigDB} module matches for which the $p$-value of the match is at least less than 0.3.
+Models with more topics tended to have more matches with higher significance.
+These results suggest that we need to train the model with more topics.
+We note that the $p$-values do \textit{not} factor for multiple hypothesis corrections, so at the moment we are only using them as relative measures of model quality.
 
-Note that these $p$-values do \textit{not} factor for multiple hypothesis corrections, so could be misleadingly significant.
+\subsection*{LDA vs. HDP}
 
-\section{Division of Work}
+We trained an HDP topic model and compared it to LDA models with similar numbers of topics, using held-out perplexity and gene set enrichment analysis as described above as our evaluation metrics.
+Since the available implementation of the posterior inference algorithm for HDP was too slow to run on the entire dataset, we did feature selection using the Seurat toolkit in R \cite{seurat} to reduce the number of genes in our dataset with low variance across cells, and trained all models on this subset of data.
+We tuned the concentration parameters in HDP to obtain a reasonable number of topics (i.e.
+~150); subsequently, we trained LDA models with similar numbers of topics for comparison.
+We found that the HDP model had higher perplexity on a held-out dataset (Figure \ref{fig:hdp-perplexity}).
+Moreover, the topics in the HDP model tended to have less significant enrichment for known gene sets from \texttt{MSigDB}, and we found that this difference to be significant (Figure \ref{fig:hdp-enrichment}, Mann-Whitney-U test, $p<0.0001$).
+Overall, the LDA model proved superior to the HDP model in this experiment.
+
+\subsection*{LDA vs. IBP-DP}
+One drawback to the HDP is that there tends to be a correlation between how frequently a topic appears across all documents and how prevalent this topic is within documents that it appears in.
+Williamson et al.
+\cite{IBP} proposed the 'focused topic model' to overcome this drawback.
+In their model, each topic $k = 1,2,...$ has a relative prevalence $\phi_k \sim Gamma(\gamma, 1)$ and a population frequency $\pi_k = \prod^k_{j=1} \mu_k$, where each $\mu_k \sim Beta(\alpha, 1) $.
+For each document $m$, whether topic $k$ appears is sampled as $b_{mk} \sim Bernoulli(\pi_k)$, and the topic proportions are sampled as $\theta_m \sim Dirichlet(b_m \cdot \phi)$.
+\\
+
+Since the code from the original IBP-DP paper was not available, we implemented an inference algorithm using collapsed Gibbs sampling \cite{IBP}.
+Due to the non-conjugacy of the model, sampling each latent variable from its full conditional required using another sampling method.
+To sample the topics parameters $\pi$ and $\phi$, we used slice sampling based on the semi-ordered stick-breaking representation of the model \cite{IBP2}.\\
+
+We tested our code on a subset of the Reuters-21578 dataset, using several different values of the concentration hyper-parameter $\alpha$, which influences the number of clusters.
+Although higher values of $\alpha$ yielded better log-likelihood values, we found that it resulted in a large number of very small topics, which are not very useful (Figure \ref{fig:ibp_words_in_topic}).
+Qualitatively, we did not find the topics from IBP-DP (Figure \ref{fig:ibptopics}) to be more coherent than topics than LDA (Figure \ref{fig:ldatopics}).
+The most prevalent topics from the IBP-DP each corresponded to similar topics from LDA; less prevalent topics tended to consist of a few unrelated words.
+These results discouraged us from optimizing the code to train this model on our scRNA-seq dataset, as we do not think it would yield more coherent gene modules than LDA.
+We emphasize that these results are not completely unexpected, as the authors of the IBP-DP paper did not show any topics from their model, nor did they assess the quality of their model with metrics other than perplexity.
 
 \section{Remaining Work and Schedule}
-- robust to dropout
-- time series
+
+In the next month, we will focus on implementing and evaluating a unified framework for clustering and topic modeling based on LDA.
+Currently, non-negative matrix factorization appears to be the primary unsupervised method for combined learning of gene modules and cell clusters from scRNA-seq data (ref); typically, the pipeline for scRNA-seq analysis involves first clustering cells, and then using differential analysis to find gene modules that distinguish each cluster (ref).
+We are currently exploring and implementing Bayesian mixture models for clustering cells; specifically, we are implementing the finite mixture model specified by the plate diagram in Figure \ref{fig:plate}.
+Our implementation uses \texttt{numpy}, and we are exploring backends to make posterior inference computationally feasible.
+If we are successful, we will combine these models with LDA and test our unified framework against existing methods of cell clustering and gene module discovery.
 
@@
 \begin{figure}
-    \label{fig:time}
     \centering
     \includegraphics[width=0.5\textwidth]{time}
-    \caption{Comparison of the running times of each of the posterior estimation methods across various cluster sizes.}
+    \caption{Comparison of the running times of each of the posterior estimation methods across various numbers of topics.}
+    \label{fig:time}
 \end{figure}
@@
 \begin{figure}
-    \label{fig:ll}
     \centering
     \includegraphics[width=0.5\textwidth]{ll}
-    \caption{Log-likelihood of the held-out testing set, across various cluster sizes.}
+    \caption{Log-likelihood of the held-out testing set, across various numbers of topics.}
+    \label{fig:ll}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1\textwidth]{pathways}
+    \caption{Matches between the gene collections found in LDA topics and published gene sets in \texttt{MSigDB}. Cluster size refers to the number of topics in the model}
+    \label{fig:pathways}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{hdp-perplexity}
+    \caption{Comparison of log-likelihood of the held-out testing set, under various LDA models and the HDP model.}
+    \label{fig:hdp-perplexity}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{hdp-enrichment}
+    \caption{Comparison of distributions of p-values from gene set enrichment analysis between LDA models and the HDP model.}
+    \label{fig:hdp-enrichment}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{ibp_words_in_topic}
+    \caption{Beeswarm plot of number of words per topic, for 3 different IBP-DP models with different concentration parameters. Each point represents one topic from its model}
+    \label{fig:ibp_words_in_topic}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1\textwidth]{ibptopics}
+    \caption{Topics from IBP-DP model trained on subset of Reuters dataset.}
+    \label{fig:ibptopics}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1\textwidth]{ldatopics}
+    \caption{Topics from LDA model with 15 topics trained on subset of Reuters dataset.}
+    \label{fig:ldatopics}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.3\textwidth]{mixture}
+    \caption{Finite \texttt{K}-sized mixture model currently implemented.
+$\theta$ is the parameter for every cluster component, represented from a categorical draw of over all genes.
+$z_i$ is the cluster assignment, and $\phi$ is the distribution of clusters.
+As usual, $\alpha, \beta$ are hyper-parameters.}
+    \label{fig:plate}
 \end{figure}
@@
 
+    \bibitem{IBP2}
+        Stick-breaking Construction for the Indian Buffet Process
+        \url{http://mlg.eng.cam.ac.uk/zoubin/papers/TehGorGha07.pdf}
+
+    \bibitem{seurat}
+        Spatial reconstruction of single-cell gene expression data
+        \url{http://www.nature.com/nbt/journal/v33/n5/full/nbt.3192.html}
+
 \end{thebibliography}

### 2017-05-09 finalreport/paper.tex
add final report

\documentclass{article}

% For figures
\usepackage{graphicx} % more modern
\usepackage{subfigure}

% For citations
\usepackage{natbib}

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}

\usepackage{hyperref}

% Packages hyperref and algorithmic misbehave sometimes.  We can fix
% this with the following command.
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% \usepackage{sty/icml2013} 
% Employ this version of the ``usepackage'' statement after the paper has
% been accepted, when creating the final version.  This will set the
% note in the first column to ``Proceedings of the...''
\usepackage[accepted]{sty/icml2013}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Submission and Formatting Instructions for ICML 2013}

\begin{document}

\twocolumn[
\icmltitle{Submission and Formatting Instructions for the Thirtieth \\ 
           International Conference on Machine Learning (ICML 2013)}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2013
% package.
\icmlauthor{Karren Dai Yang, Skanda Koppula}{\{karren, skoppula\}@mit.edu}
\icmladdress{Massachusetts Institute of Technology,
            Cambridge, MA 02139 USA}

% You may provide any keywords that you 
% find helpful for describing your paper; these are used to populate 
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{boring formatting information, machine learning, ICML}

\vskip 0.3in
]

\begin{abstract} 
There are three review cycles for ICML 2013, with full paper submissions 
due on October 1, 2012; December 15, 2012; and February 15, 2013. Reviewing 
will be blind to the identities of the authors, and therefore identifying
information must not appear in any way in papers submitted for review. 
Submissions must be in PDF, with an 8 page length limit (upto 9 pages 
including references).
\end{abstract}

\section{Electronic Submission}
\label{submission}

As in the past few years, ICML will rely exclusively on
electronic formats for submission and review.

\subsection{Templates for Papers}

Electronic templates for producing papers for submission are available
for \LaTeX\/ and Microsoft Word. Templates are accessible on the World
Wide Web at:\\
\textbf{\texttt{http://icml.cc/2013/}}

\noindent
Send questions about these electronic templates to
\texttt{program@icml.cc}.

The formatting instructions below will be enforced for initial submissions and camera-ready copies. 
\begin{itemize}
\item The maximum paper length is 8 pages excluding references, and 
9 pages including references.
\item Do not alter the style template; in particular, do not compress the paper format 
by reducing the vertical spaces.
\item Do not include author information or acknowledgments in your
  initial submission. 
\item Place figure captions {\em under} the figure (and omit titles from
  inside the graphic file itself).  Place table captions {\em over}
  the table.
\item References must include page numbers whenever possible and be as
  complete as possible.  Place multiple citations in chronological order.  
\end{itemize}
Please see below for details on each of these items.

\subsection{Submitting Papers}

Submission to ICML 2013 will be entirely electronic, via a web site
(not email).  The URL and information about the submission process
are available on the conference web site at

\textbf{\texttt{http://icml.cc/2013/}}

{\bf Paper Deadline:} The deadline for paper submission to ICML 2013
is at 23:59 Universal Time (3:59 Pacific Daylight Time) on the due date
(October 1, December 15, or February 15, depending on the review cycle).  
If your full submission does not reach us by this time, it will 
not be considered for publication. There is no separate abstract submission.

{\bf Anonymous Submission:} To facilitate blind review, no identifying
author information should appear on the title page or in the paper
itself.  Section~\ref{author info} will explain the details of how to
format this.

{\bf Simultaneous Submission:} ICML will not accept any paper which,
at the time of submission, is under review for another conference or
has already been published. This policy also applies to papers that
overlap substantially in technical content with conference papers
under review or previously published. ICML submissions must not be
submitted to other conferences during ICML's review period. Authors
may submit to ICML substantially different versions of journal papers
that are currently under review by the journal, but not yet accepted
at the time of submission. Informal publications, such as technical
reports or papers in workshop proceedings which do not appear in
print, do not fall under these restrictions.

\medskip

To ensure our ability to print submissions, authors must provide their
manuscripts in \textbf{PDF} format.  Furthermore, please make sure
that files contain only Type-1 fonts (e.g.,~using the program {\tt
  pdffonts} in linux or using File/DocumentProperties/Fonts in
Acrobat).  Other fonts (like Type-3) might come from graphics files
imported into the document.

Authors using \textbf{Word} must convert their document to PDF.  Most
of the latest versions of Word have the facility to do this
automatically.  Submissions will not be accepted in Word format or any
format other than PDF. Really. We're not joking. Don't send Word.

Those who use \textbf{\LaTeX} to format their accepted papers need to
pay close attention to the typefaces used.  Specifically, when
producing the PDF by first converting the dvi output of \LaTeX\ to Postscript
the default behavior is to use non-scalable Type-3 PostScript bitmap
fonts to represent the standard \LaTeX\ fonts. The resulting document
is difficult to read in electronic form; the type appears fuzzy. To
avoid this problem, dvips must be instructed to use an alternative
font map.  This can be achieved with
something like the following commands:\\[0.5em]
{\bf dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi}\\
{\bf ps2pdf paper.ps}\\[0.5em]
Note that it is a zero following the ``-G''.  This tells dvips to use
the config.pdf file (and this file refers to a better font mapping).

Another alternative is to use the \textbf{pdflatex} program instead of
straight \LaTeX. This program avoids the Type-3 font problem, however
you must ensure that all of the fonts are embedded (use {\tt
pdffonts}). If they are not, you need to configure pdflatex to use a
font map file that specifies that the fonts be embedded. Also you
should ensure that images are not downsampled or otherwise compressed
in a lossy way.

Note that the 2013 style files use the {\tt hyperref} package to
make clickable links in documents.  If this causes problems for you,
add {\tt nohyperref} as one of the options to the {\tt icml2013}
usepackage statement.

\subsection{Reacting to Reviews}
We will continue the ICML tradition in which the authors are given the
option of providing a short reaction to the initial reviews. These
reactions will be taken into account in the discussion among the
reviewers and area chairs.

\subsection{Submitting Final Camera-Ready Copy}

The final versions of papers accepted for publication should follow the
same format and naming convention as initial submissions, except of
course that the normal author information (names and affiliations)
should be given.  See Section~\ref{final author} for details of how to
format this.

The footnote, ``Preliminary work.  Under review by the International
Conference on Machine Learning (ICML).  Do not distribute.'' must be
modified to ``\textit{Proceedings of the
$\mathit{30}^{th}$ International Conference on Machine Learning},
Atlanta, Georgia, USA, 2013.  JMLR: W\&CP volume 28. 
Copyright 2013 by the author(s).''

For those using the \textbf{\LaTeX} style file, simply change
$\mathtt{\backslash usepackage\{icml2013\}}$ to

\verb|\usepackage[accepted]{icml2013}|

\noindent
Authors using \textbf{Word} must edit the
footnote on the first page of the document themselves.

Camera-ready copies should have the title of the paper as running head
on each page except the first one.  The running title consists of a
single line centered above a horizontal rule which is $1$ point thick.
The running head should be centered, bold and in $9$ point type.  The
rule should be $10$ points above the main text.  For those using the
\textbf{\LaTeX} style file, the original title is automatically set as running
head using the {\tt fancyhdr} package which is included in the ICML
2013 style file package.  In case that the original title exceeds the
size restrictions, a shorter form can be supplied by using

\verb|\icmltitlerunning{...}|

just before $\mathtt{\backslash begin\{document\}}$.
Authors using \textbf{Word} must edit the header of the document themselves.

\section{Format of the Paper}

All submissions must follow the same format to ensure the printer can
reproduce them without problems and to let readers more easily find
the information that they desire.

\subsection{Length and Dimensions}

Papers must not exceed eight (8) pages, including all figures, tables,
and appendices, but excluding references. When references are included,
the paper must not exceed nine (9) pages. Any submission that exceeds 
this page limit or that diverges significantly from the format specified 
herein will be rejected without review.

The text of the paper should be formatted in two columns, with an
overall width of 6.75 inches, height of 9.0 inches, and 0.25 inches
between the columns. The left margin should be 0.75 inches and the top
margin 1.0 inch (2.54~cm). The right and bottom margins will depend on
whether you print on US letter or A4 paper, but all final versions
must be produced for US letter size.

The paper body should be set in 10~point type with a vertical spacing
of 11~points. Please use Times Roman typeface throughout the text.

\subsection{Title}

The paper title should be set in 14~point bold type and centered
between two horizontal rules that are 1~point thick, with 1.0~inch
between the top rule and the top edge of the page. Capitalize the
first letter of content words and put the rest of the title in lower
case.

\subsection{Author Information for Submission}
\label{author info}

To facilitate blind review, author information must not appear.  If
you are using \LaTeX\/ and the \texttt{icml2013.sty} file, you may use
\verb+\icmlauthor{...}+ to specify authors.  The author information
will simply not be printed until {\tt accepted} is an argument to the
style file. Submissions that include the author information will not
be reviewed.

\subsubsection{Self-Citations}

If your are citing published papers for which you are an author, refer
to yourself in the third person. In particular, do not use phrases
that reveal your identity (e.g., ``in previous work \cite{langley00}, we 
have shown \ldots'').

Do not anonymize citations in the reference section by removing or
blacking out author names. The only exception are manuscripts that are
not yet published (e.g. under submission). If you choose to refer to
such unpublished manuscripts \cite{anonymous}, anonymized copies have 
to be submitted
as Supplementary Material via CMT. However, keep in mind that an ICML
paper should be self contained and should contain sufficient detail
for the reviewers to evaluate the work. In particular, reviewers are
not required to look a the Supplementary Material when writing their
review.

\subsubsection{Camera-Ready Author Information}
\label{final author}

If a paper is accepted, a final camera-ready copy must be prepared.
%
For camera-ready papers, author information should start 0.3~inches
below the bottom rule surrounding the title. The authors' names should
appear in 10~point bold type, electronic mail addresses in 10~point
small capitals, and physical addresses in ordinary 10~point type.
Each author's name should be flush left, whereas the email address
should be flush right on the same line. The author's physical address
should appear flush left on the ensuing line, on a single line if
possible. If successive authors have the same affiliation, then give
their physical address only once.

A sample file (in PDF) with author names is included in the ICML2013 
style file package.

\subsection{Abstract}

The paper abstract should begin in the left column, 0.4~inches below
the final address. The heading `Abstract' should be centered, bold,
and in 11~point type. The abstract body should use 10~point type, with
a vertical spacing of 11~points, and should be indented 0.25~inches
more than normal on left-hand and right-hand margins. Insert
0.4~inches of blank space after the body. Keep your abstract brief and 
self-contained,
limiting it to one paragraph and no more than six or seven sentences.

\subsection{Partitioning the Text}

You should organize your paper into sections and paragraphs to help
readers place a structure on the material and understand its
contributions.

\subsubsection{Sections and Subsections}

Section headings should be numbered, flush left, and set in 11~pt bold
type with the content words capitalized. Leave 0.25~inches of space
before the heading and 0.15~inches after the heading.

Similarly, subsection headings should be numbered, flush left, and set
in 10~pt bold type with the content words capitalized. Leave
0.2~inches of space before the heading and 0.13~inches afterward.

Finally, subsubsection headings should be numbered, flush left, and
set in 10~pt small caps with the content words capitalized. Leave
0.18~inches of space before the heading and 0.1~inches after the
heading.

Please use no more than three levels of headings.

\subsubsection{Paragraphs and Footnotes}

Within each section or subsection, you should further partition the
paper into paragraphs. Do not indent the first line of a given
paragraph, but insert a blank line between succeeding ones.

You can use footnotes\footnote{For the sake of readability, footnotes
should be complete sentences.} to provide readers with additional
information about a topic without interrupting the flow of the paper. 
Indicate footnotes with a number in the text where the point is most
relevant. Place the footnote in 9~point type at the bottom of the
column in which it appears. Precede the first footnote in a column
with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
appear in each column, in the same order as they appear in the text,
but spread them across columns and pages if possible.}

\begin{figure}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{figs/examplegraph}}
\caption{Historical locations and number of accepted papers for International
  Machine Learning Conferences (ICML 1993 -- ICML 2008) and
  International Workshops on Machine Learning (ML 1988 -- ML
  1992). At the time this figure was produced, the number of
  accepted papers for ICML 2008 was unknown and instead estimated.}
\label{icml-historical}
\end{center}
\vskip -0.2in
\end{figure}

\subsection{Figures}

You may want to include figures in the paper to help readers visualize
your approach and your results. Such artwork should be centered,
legible, and separated from the text. Lines should be dark and at
least 0.5~points thick for purposes of reproduction, and text should
not appear on a gray background.

Label all distinct components of each figure. If the figure takes the
form of a graph, then give a name for each axis and include a legend
that briefly describes each curve. Do not include a title inside the
figure; instead, the caption should serve this function.

Number figures sequentially, placing the figure number and caption
{\it after\/} the graphics, with at least 0.1~inches of space before
the caption and 0.1~inches after it, as in
Figure~\ref{icml-historical}.  The figure caption should be set in
9~point type and centered unless it runs two or more lines, in which
case it should be flush left.  You may float figures to the top or
bottom of a column, and you may set wide figures across both columns
(use the environment {\tt figure*} in \LaTeX), but always place
two-column figures at the top or bottom of the page.

\subsection{Algorithms}

If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic'' 
environments to format pseudocode. These require 
the corresponding stylefiles, algorithm.sty and 
algorithmic.sty, which are supplied with this package. 
Algorithm~\ref{alg:example} shows an example.

\begin{algorithm}[tb]
   \caption{Bubble Sort}
   \label{alg:example}
\begin{algorithmic}
   \STATE {\bfseries Input:} data $x_i$, size $m$
   \REPEAT
   \STATE Initialize $noChange = true$.
   \FOR{$i=1$ {\bfseries to} $m-1$}
   \IF{$x_i > x_{i+1}$} 
   \STATE Swap $x_i$ and $x_{i+1}$
   \STATE $noChange = false$
   \ENDIF
   \ENDFOR
   \UNTIL{$noChange$ is $true$}
\end{algorithmic}
\end{algorithm}

\subsection{Tables}

You may also want to include tables that summarize material. Like 
figures, these should be centered, legible, and numbered consecutively. 
However, place the title {\it above\/} the table with at least 
0.1~inches of space before the title and the same after it, as in 
Table~\ref{sample-table}. The table title should be set in 9~point 
type and centered unless it runs two or more lines, in which case it
should be flush left.

% Note use of \abovespace and \belowspace to get reasonable spacing 
% above and below tabular lines.

\begin{table}[t]
\caption{Classification accuracies for naive Bayes and flexible 
Bayes on various data sets.}
\label{sample-table}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcccr}
\hline
\abovespace\belowspace
Data set & Naive & Flexible & Better? \\
\hline
\abovespace
Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
\belowspace
Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
\hline
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

Tables contain textual material that can be typeset, as contrasted 
with figures, which contain graphical material that must be drawn. 
Specify the contents of each row and column in the table's topmost
row. Again, you may float tables to a column's top or bottom, and set
wide tables across both columns, but place two-column tables at the
top or bottom of the page.

\subsection{Citations and References}

Please use APA reference format regardless of your formatter
or word processor. If you rely on the \LaTeX\/ bibliographic 
facility, use {\tt natbib.sty} and {\tt icml2013.bst} 
included in the style-file package to obtain this format.

Citations within the text should include the authors' last names and
year. If the authors' names are included in the sentence, place only
the year in parentheses, for example when referencing Arthur Samuel's
pioneering work \yrcite{Samuel59}. Otherwise place the entire
reference in parentheses with the authors and year separated by a
comma \cite{Samuel59}. List multiple references separated by
semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
construct only for citations with three or more authors or after
listing all authors to a publication in an earlier reference \cite{MachineLearningI}.

Authors should cite their own work in the third person
in the initial version of their paper submitted for blind review.
Please refer to Section~\ref{author info} for detailed instructions on how to
cite your own papers.

Use an unnumbered first-level section heading for the references, and 
use a hanging indent style, with the first line of the reference flush
against the left margin and subsequent lines indented by 10 points. 
The references at the end of this document give examples for journal
articles \cite{Samuel59}, conference publications \cite{langley00}, book chapters \cite{Newell81}, books \cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, 
technical reports \cite{mitchell80}, and dissertations \cite{kearns89}.

Alphabetize references by the surnames of the first authors, with
single author entries preceding multiple author entries. Order
references for the same authors by year of publication, with the
earliest first. Make sure that each reference includes all relevant
information (e.g., page numbers).

\subsection{Software and Data}

We strongly encourage the publication of software and data with the
camera-ready version of the paper whenever appropriate.  This can be
done by including a URL in the camera-ready copy.  However, do not
include URLs that reveal your institution or identity in your
submission for review.  Instead, provide an anonymous URL or upload
the material as ``Supplementary Material'' into the CMT reviewing
system.  Note that reviewers are not required to look a this material
when writing their review.

% Acknowledgements should only appear in the accepted version. 
\section*{Acknowledgments}

\textbf{Do not} include acknowledgements in the initial version of
the paper submitted for blind review.

If a paper is accepted, the final camera-ready version can (and
probably should) include acknowledgements. In this case, please
place such acknowledgements in an unnumbered section at the
end of the paper. Typically, this will include thanks to reviewers
who gave useful comments, to colleagues who contributed to the ideas, 
and to funding agencies and corporate sponsors that provided financial 
support.

% In the unusual situation where you want a paper to appear in the
% references without citing it in the main text, use \nocite
\nocite{langley00}

\bibliography{citations}
\bibliographystyle{sty/icml2013}

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was
% created by Lise Getoor and Tobias Scheffer, it was slightly modified  
% from the 2010 version by Thorsten Joachims & Johannes Fuernkranz, 
% slightly modified from the 2009 version by Kiri Wagstaff and 
% Sam Roweis's 2008 version, which is slightly modified from 
% Prasad Tadepalli's 2007 version which is a lightly 
% changed version of the previous year's version by Andrew Moore, 
% which was in turn edited from those of Kristian Kersting and 
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2017-05-09 finalreport/build.sh

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+file="finalreport"
+author1="karren"
+author2="skoppula"
+totalfile="$file-$author1-$author2"
+
+# Compile the tex file
+pdflatex --shell-escape -jobname=$totalfile paper.tex
+pdflatex --shell-escape -jobname=$totalfile paper.tex
+
+if [ "$(uname)" == "Darwin" ]; then
+    open ${totalfile}.pdf &
+elif [ "$(expr substr $(uname -s) 1 5)" == "Linux" ]; then
+    evince ${totalfile}.pdf &
+elif [ "$(expr substr $(uname -s) 1 10)" == "MINGW32_NT" ]; then
+    echo 'Windows not supported!'
+fi
+
+rm *.aux *.log *.out

### 2017-05-10 finalreport/paper.tex
finish shit

@@
 % For citations
-\usepackage{natbib}
+% \usepackage{natbib}
 
@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Submission and Formatting Instructions for ICML 2013}
+\icmltitlerunning{Clustering and Topic Discovery in Gene Expression Data}
 
@@
 \twocolumn[
-\icmltitle{Submission and Formatting Instructions for the Thirtieth \\ 
-           International Conference on Machine Learning (ICML 2013)}
+\icmltitle{Bayesian Clustering and Topic Discovery \\ 
+           in Melonoma Cell Gene Expression Data}
 
@@
 \begin{abstract} 
-There are three review cycles for ICML 2013, with full paper submissions 
-due on October 1, 2012; December 15, 2012; and February 15, 2013. Reviewing 
-will be blind to the identities of the authors, and therefore identifying
-information must not appear in any way in papers submitted for review. 
-Submissions must be in PDF, with an 8 page length limit (upto 9 pages 
-including references).
+    We present a report of the methods and results of our use of various Bayesian models to analyze  gene expression data.
+We explored parallelized LDA, mixture models, dynamic-time topic models, topic clustering models, and non-parametric models, implemented in \texttt{numpy}, \texttt{C++}, and \texttt{Edward}.
+Overcoming problems associated with the size and dimensionality of our dataset and difficulty of posterior inference, we report held-out likelihood, performance metrics, posterior predictive checks and, most notably, meaningful biologically-meaningful topics.
 \end{abstract}
 
-\section{Electronic Submission}
-\label{submission}
+\section{Introduction}
+\label{intro}
 
-As in the past few years, ICML will rely exclusively on
-electronic formats for submission and review.
+Tumors are composed of different sub-populations of cells, and these sub-populations often exhibit shared patterns of gene expression.
+With contemporary sequencing machines, it is possible to obtain the expression levels of 10,000+ genes for 1000+ cells in a single experiment.
 
-\subsection{Templates for Papers}
+% In this project, we applied Bayesian methods for clustering and topic discovery to discover meaningful cell clusters and gene modules from single-cell RNA-sequencing (scRNA-seq) datasets from tumors.
+We first intend to explore the use of a hierarchical topic model, Latent Dirichlet Allocation (LDA) \cite{LDA}, as well as two non-parametric topic models, Hierarchical Dirichlet Processes (HDP) \cite{HDP} and the Indian Buffet Process Compound Dirichlet Process (IBP-CDP) \cite{IBP} for analyzing scRNA-seq data.
+We will evaluate these models based on their ability to discover meaningful functional gene modules.
+Subsequently, we plan to explore the use of mixture models to cluster cells based on the output of the topic models, which we will evaluate based on their ability to differentiate cell types (e.g.
+immune, cancer, non-cancerous).
+Finally, we will train a combined clustering-topic model (e.g.
+MGCTM \cite{pengtao}) to see if it outperforms the individual models.
+We will compare the results of these methods with standard methods used in prior work, and evaluate these methods' robustness to noise.
+Time permitting, we will study extensions of these models appropriate for time-series scRNA-seq data \cite{TS-DDP}.
+\\
 
-Electronic templates for producing papers for submission are available
-for \LaTeX\/ and Microsoft Word. Templates are accessible on the World
-Wide Web at:\\
-\textbf{\texttt{http://icml.cc/2013/}}
-
-\noindent
-Send questions about these electronic templates to
-\texttt{program@icml.cc}.
-
-The formatting instructions below will be enforced for initial submissions and camera-ready copies. 
-\begin{itemize}
-\item The maximum paper length is 8 pages excluding references, and 
-9 pages including references.
-\item Do not alter the style template; in particular, do not compress the paper format 
-by reducing the vertical spaces.
-\item Do not include author information or acknowledgments in your
-  initial submission. 
-\item Place figure captions {\em under} the figure (and omit titles from
-  inside the graphic file itself).  Place table captions {\em over}
-  the table.
-\item References must include page numbers whenever possible and be as
-  complete as possible.  Place multiple citations in chronological order.  
-\end{itemize}
-Please see below for details on each of these items.
-
-\subsection{Submitting Papers}
-
-Submission to ICML 2013 will be entirely electronic, via a web site
-(not email).  The URL and information about the submission process
-are available on the conference web site at
-
-\textbf{\texttt{http://icml.cc/2013/}}
-
-{\bf Paper Deadline:} The deadline for paper submission to ICML 2013
-is at 23:59 Universal Time (3:59 Pacific Daylight Time) on the due date
-(October 1, December 15, or February 15, depending on the review cycle).  
-If your full submission does not reach us by this time, it will 
-not be considered for publication. There is no separate abstract submission.
-
-{\bf Anonymous Submission:} To facilitate blind review, no identifying
-author information should appear on the title page or in the paper
-itself.  Section~\ref{author info} will explain the details of how to
-format this.
-
-{\bf Simultaneous Submission:} ICML will not accept any paper which,
-at the time of submission, is under review for another conference or
-has already been published. This policy also applies to papers that
-overlap substantially in technical content with conference papers
-under review or previously published. ICML submissions must not be
-submitted to other conferences during ICML's review period. Authors
-may submit to ICML substantially different versions of journal papers
-that are currently under review by the journal, but not yet accepted
-at the time of submission. Informal publications, such as technical
-reports or papers in workshop proceedings which do not appear in
-print, do not fall under these restrictions.
-
-\medskip
-
-To ensure our ability to print submissions, authors must provide their
-manuscripts in \textbf{PDF} format.  Furthermore, please make sure
-that files contain only Type-1 fonts (e.g.,~using the program {\tt
-  pdffonts} in linux or using File/DocumentProperties/Fonts in
-Acrobat).  Other fonts (like Type-3) might come from graphics files
-imported into the document.
-
-Authors using \textbf{Word} must convert their document to PDF.  Most
-of the latest versions of Word have the facility to do this
-automatically.  Submissions will not be accepted in Word format or any
-format other than PDF. Really. We're not joking. Don't send Word.
-
-Those who use \textbf{\LaTeX} to format their accepted papers need to
-pay close attention to the typefaces used.  Specifically, when
-producing the PDF by first converting the dvi output of \LaTeX\ to Postscript
-the default behavior is to use non-scalable Type-3 PostScript bitmap
-fonts to represent the standard \LaTeX\ fonts. The resulting document
-is difficult to read in electronic form; the type appears fuzzy. To
-avoid this problem, dvips must be instructed to use an alternative
-font map.  This can be achieved with
-something like the following commands:\\[0.5em]
-{\bf dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi}\\
-{\bf ps2pdf paper.ps}\\[0.5em]
-Note that it is a zero following the ``-G''.  This tells dvips to use
-the config.pdf file (and this file refers to a better font mapping).
-
-Another alternative is to use the \textbf{pdflatex} program instead of
-straight \LaTeX. This program avoids the Type-3 font problem, however
-you must ensure that all of the fonts are embedded (use {\tt
-pdffonts}). If they are not, you need to configure pdflatex to use a
-font map file that specifies that the fonts be embedded. Also you
-should ensure that images are not downsampled or otherwise compressed
-in a lossy way.
-
-Note that the 2013 style files use the {\tt hyperref} package to
-make clickable links in documents.  If this causes problems for you,
-add {\tt nohyperref} as one of the options to the {\tt icml2013}
-usepackage statement.
-
-\subsection{Reacting to Reviews}
-We will continue the ICML tradition in which the authors are given the
-option of providing a short reaction to the initial reviews. These
-reactions will be taken into account in the discussion among the
-reviewers and area chairs.
-
-\subsection{Submitting Final Camera-Ready Copy}
-
-The final versions of papers accepted for publication should follow the
-same format and naming convention as initial submissions, except of
-course that the normal author information (names and affiliations)
-should be given.  See Section~\ref{final author} for details of how to
-format this.
-
-The footnote, ``Preliminary work.  Under review by the International
-Conference on Machine Learning (ICML).  Do not distribute.'' must be
-modified to ``\textit{Proceedings of the
-$\mathit{30}^{th}$ International Conference on Machine Learning},
-Atlanta, Georgia, USA, 2013.  JMLR: W\&CP volume 28. 
-Copyright 2013 by the author(s).''
-
-For those using the \textbf{\LaTeX} style file, simply change
-$\mathtt{\backslash usepackage\{icml2013\}}$ to
-
-\verb|\usepackage[accepted]{icml2013}|
-
-\noindent
-Authors using \textbf{Word} must edit the
-footnote on the first page of the document themselves.
-
-Camera-ready copies should have the title of the paper as running head
-on each page except the first one.  The running title consists of a
-single line centered above a horizontal rule which is $1$ point thick.
-The running head should be centered, bold and in $9$ point type.  The
-rule should be $10$ points above the main text.  For those using the
-\textbf{\LaTeX} style file, the original title is automatically set as running
-head using the {\tt fancyhdr} package which is included in the ICML
-2013 style file package.  In case that the original title exceeds the
-size restrictions, a shorter form can be supplied by using
-
-\verb|\icmltitlerunning{...}|
-
-just before $\mathtt{\backslash begin\{document\}}$.
-Authors using \textbf{Word} must edit the header of the document themselves.
+\section{Prior work}
+% Prior research has applied basic measures of statistical distance to cluster cell groups or gene modules based on samples of single-cell RNA-sequencing (scRNA-seq) data taken from tumors \cite{nature}.
+Good clustering of scRNA-seq data often has biologically meaningful results, and as such, a wide variety of correlation based methods have been used in prior work to derive meaning from scRNA-seq data \cite{coexpression, consensus, profiling}.
+We believe that Bayesian methods for clustering and topic modeling can be more illuminating than these previous methods, due to their consideration of higher-order relationships within the data.
+\\
 
@@
 
-All submissions must follow the same format to ensure the printer can
-reproduce them without problems and to let readers more easily find
-the information that they desire.
-
-\subsection{Length and Dimensions}
-
-Papers must not exceed eight (8) pages, including all figures, tables,
-and appendices, but excluding references. When references are included,
-the paper must not exceed nine (9) pages. Any submission that exceeds 
-this page limit or that diverges significantly from the format specified 
-herein will be rejected without review.
-
-The text of the paper should be formatted in two columns, with an
-overall width of 6.75 inches, height of 9.0 inches, and 0.25 inches
-between the columns. The left margin should be 0.75 inches and the top
-margin 1.0 inch (2.54~cm). The right and bottom margins will depend on
-whether you print on US letter or A4 paper, but all final versions
-must be produced for US letter size.
-
-The paper body should be set in 10~point type with a vertical spacing
-of 11~points. Please use Times Roman typeface throughout the text.
-
-\subsection{Title}
-
-The paper title should be set in 14~point bold type and centered
-between two horizontal rules that are 1~point thick, with 1.0~inch
-between the top rule and the top edge of the page. Capitalize the
-first letter of content words and put the rest of the title in lower
-case.
-
-\subsection{Author Information for Submission}
-\label{author info}
-
-To facilitate blind review, author information must not appear.  If
-you are using \LaTeX\/ and the \texttt{icml2013.sty} file, you may use
-\verb+\icmlauthor{...}+ to specify authors.  The author information
-will simply not be printed until {\tt accepted} is an argument to the
-style file. Submissions that include the author information will not
-be reviewed.
-
-\subsubsection{Self-Citations}
-
-If your are citing published papers for which you are an author, refer
-to yourself in the third person. In particular, do not use phrases
-that reveal your identity (e.g., ``in previous work \cite{langley00}, we 
-have shown \ldots'').
-
-Do not anonymize citations in the reference section by removing or
-blacking out author names. The only exception are manuscripts that are
-not yet published (e.g. under submission). If you choose to refer to
-such unpublished manuscripts \cite{anonymous}, anonymized copies have 
-to be submitted
-as Supplementary Material via CMT. However, keep in mind that an ICML
-paper should be self contained and should contain sufficient detail
-for the reviewers to evaluate the work. In particular, reviewers are
-not required to look a the Supplementary Material when writing their
-review.
-
-\subsubsection{Camera-Ready Author Information}
-\label{final author}
-
-If a paper is accepted, a final camera-ready copy must be prepared.
-%
-For camera-ready papers, author information should start 0.3~inches
-below the bottom rule surrounding the title. The authors' names should
-appear in 10~point bold type, electronic mail addresses in 10~point
-small capitals, and physical addresses in ordinary 10~point type.
-Each author's name should be flush left, whereas the email address
-should be flush right on the same line. The author's physical address
-should appear flush left on the ensuing line, on a single line if
-possible. If successive authors have the same affiliation, then give
-their physical address only once.
-
-A sample file (in PDF) with author names is included in the ICML2013 
-style file package.
-
-\subsection{Abstract}
-
-The paper abstract should begin in the left column, 0.4~inches below
-the final address. The heading `Abstract' should be centered, bold,
-and in 11~point type. The abstract body should use 10~point type, with
-a vertical spacing of 11~points, and should be indented 0.25~inches
-more than normal on left-hand and right-hand margins. Insert
-0.4~inches of blank space after the body. Keep your abstract brief and 
-self-contained,
-limiting it to one paragraph and no more than six or seven sentences.
-
 \subsection{Partitioning the Text}
-
-You should organize your paper into sections and paragraphs to help
-readers place a structure on the material and understand its
-contributions.
-
-\subsubsection{Sections and Subsections}
-
-Section headings should be numbered, flush left, and set in 11~pt bold
-type with the content words capitalized. Leave 0.25~inches of space
-before the heading and 0.15~inches after the heading.
-
-Similarly, subsection headings should be numbered, flush left, and set
-in 10~pt bold type with the content words capitalized. Leave
-0.2~inches of space before the heading and 0.13~inches afterward.
-
-Finally, subsubsection headings should be numbered, flush left, and
-set in 10~pt small caps with the content words capitalized. Leave
-0.18~inches of space before the heading and 0.1~inches after the
-heading.
-
-Please use no more than three levels of headings.
-
-\subsubsection{Paragraphs and Footnotes}
-
-Within each section or subsection, you should further partition the
-paper into paragraphs. Do not indent the first line of a given
-paragraph, but insert a blank line between succeeding ones.
-
-You can use footnotes\footnote{For the sake of readability, footnotes
-should be complete sentences.} to provide readers with additional
-information about a topic without interrupting the flow of the paper. 
-Indicate footnotes with a number in the text where the point is most
-relevant. Place the footnote in 9~point type at the bottom of the
-column in which it appears. Precede the first footnote in a column
-with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
-appear in each column, in the same order as they appear in the text,
-but spread them across columns and pages if possible.}
 
@@
 \end{figure}
-
-\subsection{Figures}
-
-You may want to include figures in the paper to help readers visualize
-your approach and your results. Such artwork should be centered,
-legible, and separated from the text. Lines should be dark and at
-least 0.5~points thick for purposes of reproduction, and text should
-not appear on a gray background.
-
-Label all distinct components of each figure. If the figure takes the
-form of a graph, then give a name for each axis and include a legend
-that briefly describes each curve. Do not include a title inside the
-figure; instead, the caption should serve this function.
-
-Number figures sequentially, placing the figure number and caption
-{\it after\/} the graphics, with at least 0.1~inches of space before
-the caption and 0.1~inches after it, as in
-Figure~\ref{icml-historical}.  The figure caption should be set in
-9~point type and centered unless it runs two or more lines, in which
-case it should be flush left.  You may float figures to the top or
-bottom of a column, and you may set wide figures across both columns
-(use the environment {\tt figure*} in \LaTeX), but always place
-two-column figures at the top or bottom of the page.
 
@@
 
-\subsection{Tables}
-
-You may also want to include tables that summarize material. Like 
-figures, these should be centered, legible, and numbered consecutively. 
-However, place the title {\it above\/} the table with at least 
-0.1~inches of space before the title and the same after it, as in 
-Table~\ref{sample-table}. The table title should be set in 9~point 
-type and centered unless it runs two or more lines, in which case it
-should be flush left.
-
-% Note use of \abovespace and \belowspace to get reasonable spacing 
-% above and below tabular lines.
-
-\begin{table}[t]
-\caption{Classification accuracies for naive Bayes and flexible 
-Bayes on various data sets.}
-\label{sample-table}
-\vskip 0.15in
-\begin{center}
-\begin{small}
-\begin{sc}
-\begin{tabular}{lcccr}
-\hline
-\abovespace\belowspace
-Data set & Naive & Flexible & Better? \\
-\hline
-\abovespace
-Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
-Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
-Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
-Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
-Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
-Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
-Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
-\belowspace
-Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
-\hline
-\end{tabular}
-\end{sc}
-\end{small}
-\end{center}
-\vskip -0.1in
-\end{table}
-
-Tables contain textual material that can be typeset, as contrasted 
-with figures, which contain graphical material that must be drawn. 
-Specify the contents of each row and column in the table's topmost
-row. Again, you may float tables to a column's top or bottom, and set
-wide tables across both columns, but place two-column tables at the
-top or bottom of the page.
-
 \subsection{Citations and References}
 
-Please use APA reference format regardless of your formatter
-or word processor. If you rely on the \LaTeX\/ bibliographic 
-facility, use {\tt natbib.sty} and {\tt icml2013.bst} 
-included in the style-file package to obtain this format.
-
-Citations within the text should include the authors' last names and
-year. If the authors' names are included in the sentence, place only
-the year in parentheses, for example when referencing Arthur Samuel's
-pioneering work \yrcite{Samuel59}. Otherwise place the entire
-reference in parentheses with the authors and year separated by a
-comma \cite{Samuel59}. List multiple references separated by
-semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
-construct only for citations with three or more authors or after
-listing all authors to a publication in an earlier reference \cite{MachineLearningI}.
-
-Authors should cite their own work in the third person
-in the initial version of their paper submitted for blind review.
-Please refer to Section~\ref{author info} for detailed instructions on how to
-cite your own papers.
-
-Use an unnumbered first-level section heading for the references, and 
-use a hanging indent style, with the first line of the reference flush
-against the left margin and subsequent lines indented by 10 points. 
-The references at the end of this document give examples for journal
-articles \cite{Samuel59}, conference publications \cite{langley00}, book chapters \cite{Newell81}, books \cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, 
-technical reports \cite{mitchell80}, and dissertations \cite{kearns89}.
-
-Alphabetize references by the surnames of the first authors, with
-single author entries preceding multiple author entries. Order
-references for the same authors by year of publication, with the
-earliest first. Make sure that each reference includes all relevant
-information (e.g., page numbers).
-
-\subsection{Software and Data}
-
-We strongly encourage the publication of software and data with the
-camera-ready version of the paper whenever appropriate.  This can be
-done by including a URL in the camera-ready copy.  However, do not
-include URLs that reveal your institution or identity in your
-submission for review.  Instead, provide an anonymous URL or upload
-the material as ``Supplementary Material'' into the CMT reviewing
-system.  Note that reviewers are not required to look a this material
-when writing their review.
-
-% Acknowledgements should only appear in the accepted version. 
 \section*{Acknowledgments}
 
-\textbf{Do not} include acknowledgements in the initial version of
-the paper submitted for blind review.
-
-If a paper is accepted, the final camera-ready version can (and
-probably should) include acknowledgements. In this case, please
-place such acknowledgements in an unnumbered section at the
-end of the paper. Typically, this will include thanks to reviewers
-who gave useful comments, to colleagues who contributed to the ideas, 
-and to funding agencies and corporate sponsors that provided financial 
 support.
@@
 % references without citing it in the main text, use \nocite
-\nocite{langley00}
+% \nocite{langley00}
 
-\bibliography{citations}
-\bibliographystyle{sty/icml2013}
+% \bibliography{citations}
+% \bibliographystyle{sty/icml2013}
+
+\begin{thebibliography}{9}
+    \bibitem{nature} 
+        Spearman-based hierarchical clustering of scRNA-seq.
+        \url{http://www.nature.com/nsmb/journal/v24/n3/fig_tab/nsmb.3365_SF1.html}
+
+    \bibitem{coexpression} 
+        Exploiting single-cell expression to characterize co-expression replicability.
+        \url{https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-0964-6}
+
+    \bibitem{consensus} 
+        SC3 - consensus clustering of single-cell RNA-Seq data.
+        \url{http://biorxiv.org/content/early/2016/09/02/036558}
+
+    \bibitem{profiling} 
+        SINCERA: A Pipeline for Single-Cell RNA-Seq Profiling Analysis.
+        \url{http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004575}
+
+    \bibitem{pengtao} 
+        Integrating Document Clustering and Topic Modeling.
+        \url{https://arxiv.org/pdf/1309.6874.pdf}
+
+    \bibitem{LDA}
+        Latent Dirichlet Allocation.
+        \url{http://www.jmlr.org/papers/volume3/blei03a/blei03a.pdf}
+
+    \bibitem{HDP}
+        Hierarchical Dirichlet Processes
+        \url{http://people.eecs.berkeley.edu/~jordan/papers/hdp.pdf}
+
+    \bibitem{IBP}
+        The IBP Compound Dirichlet Process and its Application to Focused Topic Modeling
+        \url{http://www.cs.columbia.edu/~blei/papers/WilliamsonWangHellerBlei2010.pdf}
+
+    \bibitem{TS-DDP}
+        A Time-Series DDP for Functional Proteomics Profiles
+        \url{https://www.ma.utexas.edu/users/pmueller/pap/NM12.pdf}
+
+\end{thebibliography}
 
 \end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was
-% created by Lise Getoor and Tobias Scheffer, it was slightly modified  
-% from the 2010 version by Thorsten Joachims & Johannes Fuernkranz, 
-% slightly modified from the 2009 version by Kiri Wagstaff and 
-% Sam Roweis's 2008 version, which is slightly modified from 
-% Prasad Tadepalli's 2007 version which is a lightly 
-% changed version of the previous year's version by Andrew Moore, 
-% which was in turn edited from those of Kristian Kersting and 
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2017-05-10 finalreport/paper.tex
fix citations

@@
 \icmltitle{Bayesian Clustering and Topic Discovery \\ 
-           in Melonoma Cell Gene Expression Data}
+           in Gene Expression Data}
 
@@
 \begin{abstract} 
-    We present a report of the methods and results of our use of various Bayesian models to analyze  gene expression data.
-We explored parallelized LDA, mixture models, dynamic-time topic models, topic clustering models, and non-parametric models, implemented in \texttt{numpy}, \texttt{C++}, and \texttt{Edward}.
-Overcoming problems associated with the size and dimensionality of our dataset and difficulty of posterior inference, we report held-out likelihood, performance metrics, posterior predictive checks and, most notably, meaningful biologically-meaningful topics.
+    We report the results using various Bayesian models to analyze gene expression data.
+Our exploration includes parallelized LDA, mixture models, dynamic-time topic models, topic-clustering models, and non-parametric models, implemented in \texttt{numpy}, \texttt{C++}, and \texttt{Edward}.
+We evaluate our methods using held-out likelihood, posterior predictive checks, and biological meaningfulness testing.
 \end{abstract}
@@
 
-% In this project, we applied Bayesian methods for clustering and topic discovery to discover meaningful cell clusters and gene modules from single-cell RNA-sequencing (scRNA-seq) datasets from tumors.
-We first intend to explore the use of a hierarchical topic model, Latent Dirichlet Allocation (LDA) \cite{LDA}, as well as two non-parametric topic models, Hierarchical Dirichlet Processes (HDP) \cite{HDP} and the Indian Buffet Process Compound Dirichlet Process (IBP-CDP) \cite{IBP} for analyzing scRNA-seq data.
+In this project, we applied Bayesian methods for clustering and topic discovery to discover meaningful cell clusters and gene modules from single-cell RNA-sequencing (scRNA-seq) datasets from tumors.
+We first intend to explore the use of a hierarchical topic model, Latent Dirichlet Allocation (LDA) \cite{LDA}, as well as two non-parametric topic models, Hierarchical Dirichlet Processes (HDP) \cite{HDP} and the Indian Buffet Process Compound Dirichlet Process (IBP-CDP) \cite{IBP} for analyzing
+scRNA-seq data.
 We will evaluate these models based on their ability to discover meaningful functional gene modules.
@@
 MGCTM \cite{pengtao}) to see if it outperforms the individual models.
-We will compare the results of these methods with standard methods used in prior work, and evaluate these methods' robustness to noise.
-Time permitting, we will study extensions of these models appropriate for time-series scRNA-seq data \cite{TS-DDP}.
+We will compare the
+results of these methods with standard methods used in prior work, and evaluate these methods' robustness to noise.
+Time permitting, we will study extensions of these models appropriate for time-series scRNA-seq data \cite{tsdpp}.
 \\
@@
 \section{Prior work}
-% Prior research has applied basic measures of statistical distance to cluster cell groups or gene modules based on samples of single-cell RNA-sequencing (scRNA-seq) data taken from tumors \cite{nature}.
+Prior research has applied basic measures of statistical distance to cluster cell groups or gene modules based on samples of single-cell RNA-sequencing (scRNA-seq) data taken from tumors \cite{nature}.
 Good clustering of scRNA-seq data often has biologically meaningful results, and as such, a wide variety of correlation based methods have been used in prior work to derive meaning from scRNA-seq data \cite{coexpression, consensus, profiling}.
@@
 
-% \bibliography{citations}
-% \bibliographystyle{sty/icml2013}
-
-\begin{thebibliography}{9}
-    \bibitem{nature} 
-        Spearman-based hierarchical clustering of scRNA-seq.
-        \url{http://www.nature.com/nsmb/journal/v24/n3/fig_tab/nsmb.3365_SF1.html}
-
-    \bibitem{coexpression} 
-        Exploiting single-cell expression to characterize co-expression replicability.
-        \url{https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-0964-6}
-
-    \bibitem{consensus} 
-        SC3 - consensus clustering of single-cell RNA-Seq data.
-        \url{http://biorxiv.org/content/early/2016/09/02/036558}
-
-    \bibitem{profiling} 
-        SINCERA: A Pipeline for Single-Cell RNA-Seq Profiling Analysis.
-        \url{http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004575}
-
-    \bibitem{pengtao} 
-        Integrating Document Clustering and Topic Modeling.
-        \url{https://arxiv.org/pdf/1309.6874.pdf}
-
-    \bibitem{LDA}
-        Latent Dirichlet Allocation.
-        \url{http://www.jmlr.org/papers/volume3/blei03a/blei03a.pdf}
-
-    \bibitem{HDP}
-        Hierarchical Dirichlet Processes
-        \url{http://people.eecs.berkeley.edu/~jordan/papers/hdp.pdf}
-
-    \bibitem{IBP}
-        The IBP Compound Dirichlet Process and its Application to Focused Topic Modeling
-        \url{http://www.cs.columbia.edu/~blei/papers/WilliamsonWangHellerBlei2010.pdf}
-
-    \bibitem{TS-DDP}
-        A Time-Series DDP for Functional Proteomics Profiles
-        \url{https://www.ma.utexas.edu/users/pmueller/pap/NM12.pdf}
-
-\end{thebibliography}
+\bibliography{citations}
+\bibliographystyle{sty/icml2013}
 

### 2017-05-10 finalreport/build.sh

@@ -7,6 +7,8 @@ totalfile="$file-$author1-$author2"
 
 # Compile the tex file
 pdflatex --shell-escape -jobname=$totalfile paper.tex
+bibtex $totalfile
+pdflatex --shell-escape -jobname=$totalfile paper.tex
 pdflatex --shell-escape -jobname=$totalfile paper.tex
 
 if [ "$(uname)" == "Darwin" ]; then
@@ -17,4 +19,4 @@ elif [ "$(expr substr $(uname -s) 1 10)" == "MINGW32_NT" ]; then
     echo 'Windows not supported!'
 fi
 
-rm *.aux *.log *.out
+rm *.aux *.log *.out *.bbl *.blg

### 2017-05-10 finalreport/paper.tex
finish shit

@@
 \twocolumn[
-\icmltitle{Bayesian Clustering and Topic Discovery \\ 
-           in Gene Expression Data}
+\icmltitle{Bayesian Clustering and Topic Discovery: \\ 
+           Adventures with Gene Expression Data}
 
@@
 
-\begin{abstract} 
-    We report the results using various Bayesian models to analyze gene expression data.
-Our exploration includes parallelized LDA, mixture models, dynamic-time topic models, topic-clustering models, and non-parametric models, implemented in \texttt{numpy}, \texttt{C++}, and \texttt{Edward}.
-We evaluate our methods using held-out likelihood, posterior predictive checks, and biological meaningfulness testing.
-\end{abstract}
-
 \section{Introduction}
 \label{intro}
-
+% copied from proposal
 Tumors are composed of different sub-populations of cells, and these sub-populations often exhibit shared patterns of gene expression.
@@
 
+\subsection{Prior work}
+% copied from proposal
+Prior research has applied basic measures of statistical distance to cluster cell groups or gene modules based on samples of single-cell RNA-sequencing (scRNA-seq) data taken from tumors \cite{nature}.
+Good clustering of scRNA-seq data often has biologically meaningful results, and as such, a wide variety of correlation based methods have been used in prior work to derive meaning from scRNA-seq data \cite{coexpression, consensus, profiling}.
+We believe that Bayesian methods for clustering and topic modeling can be more illuminating than these previous methods, due to their consideration of higher-order relationships within the data.
+\\
+
+\subsection{Description of Data}
+
+\begin{figure}[ht]
+\vskip 0.2in
+\begin{center}
+\centerline{\includegraphics[width=\columnwidth]{figs/examplegraph}}
+\caption{This is a demo figure.}
+\label{alabel}
+\end{center}
+\vskip -0.2in
+\end{figure}
+
+\subsection{Structure of Report}
+% copied from proposal
 In this project, we applied Bayesian methods for clustering and topic discovery to discover meaningful cell clusters and gene modules from single-cell RNA-sequencing (scRNA-seq) datasets from tumors.
@@
 
-\section{Prior work}
-Prior research has applied basic measures of statistical distance to cluster cell groups or gene modules based on samples of single-cell RNA-sequencing (scRNA-seq) data taken from tumors \cite{nature}.
-Good clustering of scRNA-seq data often has biologically meaningful results, and as such, a wide variety of correlation based methods have been used in prior work to derive meaning from scRNA-seq data \cite{coexpression, consensus, profiling}.
-We believe that Bayesian methods for clustering and topic modeling can be more illuminating than these previous methods, due to their consideration of higher-order relationships within the data.
-\\
+    We report the results using various Bayesian models to analyze gene expression data.
+Our exploration includes parallelized LDA, mixture models, dynamic-time topic models, topic-clustering models, and non-parametric models, implemented in \texttt{numpy}, \texttt{C++}, and \texttt{Edward}.
+We evaluate our methods using held-out likelihood, posterior predictive checks, and biological meaningfulness testing.
 
-\section{Format of the Paper}
-
-\subsection{Partitioning the Text}
-
-\begin{figure}[ht]
-\vskip 0.2in
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{figs/examplegraph}}
-\caption{Historical locations and number of accepted papers for International
-  Machine Learning Conferences (ICML 1993 -- ICML 2008) and
-  International Workshops on Machine Learning (ML 1988 -- ML
-  1992). At the time this figure was produced, the number of
-  accepted papers for ICML 2008 was unknown and instead estimated.}
-\label{icml-historical}
-\end{center}
-\vskip -0.2in
-\end{figure}
-
-\subsection{Algorithms}
-
-If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic'' 
-environments to format pseudocode. These require 
-the corresponding stylefiles, algorithm.sty and 
-algorithmic.sty, which are supplied with this package. 
-Algorithm~\ref{alg:example} shows an example.
+\section{Latent Dirichlet Allocation} 
+\subsection{Model Description} 
+    Algorithm~\ref{alg:example} describes the generative process for LDA.
 
     \begin{algorithm}[tb]
-   \caption{Bubble Sort}
+       \caption{Latent Dirichlet Allocation}
        \label{alg:example}
@@
     \end{algorithm}
+\subsection{Implementation} 
+\subsection{Experiments}
 
-\subsection{Citations and References}
+\section{Dirichlet Mixture Model} 
+\subsection{Model Description} 
+    Algorithm~\ref{alg:example} describes the generative process for the mixture odel.
 
-\section*{Acknowledgments}
+    \begin{algorithm}[tb]
+       \caption{Mixture Model}
+       \label{alg:example}
+    \begin{algorithmic}
+       \STATE {\bfseries Input:} data $x_i$, size $m$
+       \REPEAT
+       \STATE Initialize $noChange = true$.
+       \FOR{$i=1$ {\bfseries to} $m-1$}
+       \IF{$x_i > x_{i+1}$} 
+       \STATE Swap $x_i$ and $x_{i+1}$
+       \STATE $noChange = false$
+       \ENDIF
+       \ENDFOR
+       \UNTIL{$noChange$ is $true$}
+    \end{algorithmic}
+    \end{algorithm}
+\subsection{Implementation} 
+\subsection{Experiments}
 
-support.
+\section{Dynamic Time Model} 
+\subsection{Model Description} 
+\subsection{Implementation} 
+\subsection{Experiments}
+
+\section{Document Topic-Clustering Model} 
+\subsection{Model Description} 
+\subsection{Implementation} 
+\subsection{Experiments}
+
+\section{Non-parametric Models: IBP and HDP} 
+\subsection{Model Description} 
+\subsection{Implementation} 
+\subsection{Experiments}
 

### 2017-05-10 finalreport/paper.tex
finish introduction

@@
 
-% You may provide any keywords that you 
-% find helpful for describing your paper; these are used to populate 
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{boring formatting information, machine learning, ICML}
+\icmlkeywords{topic models, bayesian clustering, gene expression, gene modules}
 
@@
 \label{intro}
-% copied from proposal
-Tumors are composed of different sub-populations of cells, and these sub-populations often exhibit shared patterns of gene expression.
-With contemporary sequencing machines, it is possible to obtain the expression levels of 10,000+ genes for 1000+ cells in a single experiment.
+Tumors cell lines are composed of different sub-populations of cells which often exhibit shared patterns of gene expression.
+Biologists are interested in two key questions: given gene expressions values, (1) can we identify cell clusterings, and (2) can we identify clusters of biologically-related genes?
+Our goal with this project was to answer both these two questions using Bayesian methods.
+
+In brief, we explored the use of three Bayesian clustering methods (a vanilla mixture, integrated topic-mixture, and non-parametric models) to address the first question, and two topic models (vanilla and a dynamic-topic LDA) to address the second.
+
+\subsection{Description of Data}
+Using a contemporary gene sequencing machine, we obtained samples of single-cell RNA-sequencing data (scRNA-seq) taken from tumors in mice.
+Our data consisted of the expression values of 22,712 genes for each of 4645 cells.
+In total, the dataset amounted to 0.86 GB, presenting significant computational challenges when attempting posterior estimation.
+Apart from a few computational tricks (online inference, multi-core parallelization), for most methods, we pruned non-informative genes from the dataset, ranking based on the deviation in value across cells \footnote{We recognize that this can bias towards noisy genes.
+We favor this method because it is simple and easy to implement, and a practice used in literature \cite{pruning}}.
+We eschewed dimensionality reduction techniques such as PCA because of loss of its direct feature interpretability.
+The complete dataset, as well our preprocessed and pruned versions, are available at \cite{dataset}.
 
 \subsection{Prior work}
-% copied from proposal
-Prior research has applied basic measures of statistical distance to cluster cell groups or gene modules based on samples of single-cell RNA-sequencing (scRNA-seq) data taken from tumors \cite{nature}.
-Good clustering of scRNA-seq data often has biologically meaningful results, and as such, a wide variety of correlation based methods have been used in prior work to derive meaning from scRNA-seq data \cite{coexpression, consensus, profiling}.
-We believe that Bayesian methods for clustering and topic modeling can be more illuminating than these previous methods, due to their consideration of higher-order relationships within the data.
-\\
+Prior research has explored the use of various computational techniques to analyze gene expression data.
+Most commonly, Spearman and Pearson correlation metrics are frequently used infer sets of genes that cluster together \cite{profiling, nature}.
+Other techniques, including PCA followed by linear regression, has been used for expression-based cell clustering \cite{transcriptomics}.
+Yu et al.
+propose a unsupervised classifier ensemble as another approach to cell clustering \cite{consensus}.
 
-\subsection{Description of Data}
-
-\begin{figure}[ht]
-\vskip 0.2in
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{figs/examplegraph}}
-\caption{This is a demo figure.}
-\label{alabel}
-\end{center}
-\vskip -0.2in
-\end{figure}
+More Bayesian approaches have also been tried in prior work from the Pe'er lab.
+\cite{bayesge2} uses a Heirarchical Dirichlet Mixture Model to learn cell clusterings.
+\cite{bayesge1} builds on this to jointly learn optimal normalization pre-processing of the data.
+Bayesian networks have also been used in an attempt to learn gene dependencies from expression data \cite{bayesge3}.
 
 \subsection{Structure of Report}
-% copied from proposal
-In this project, we applied Bayesian methods for clustering and topic discovery to discover meaningful cell clusters and gene modules from single-cell RNA-sequencing (scRNA-seq) datasets from tumors.
-We first intend to explore the use of a hierarchical topic model, Latent Dirichlet Allocation (LDA) \cite{LDA}, as well as two non-parametric topic models, Hierarchical Dirichlet Processes (HDP) \cite{HDP} and the Indian Buffet Process Compound Dirichlet Process (IBP-CDP) \cite{IBP} for analyzing
-scRNA-seq data.
-We will evaluate these models based on their ability to discover meaningful functional gene modules.
-Subsequently, we plan to explore the use of mixture models to cluster cells based on the output of the topic models, which we will evaluate based on their ability to differentiate cell types (e.g.
-immune, cancer, non-cancerous).
-Finally, we will train a combined clustering-topic model (e.g.
-MGCTM \cite{pengtao}) to see if it outperforms the individual models.
-We will compare the
-results of these methods with standard methods used in prior work, and evaluate these methods' robustness to noise.
-Time permitting, we will study extensions of these models appropriate for time-series scRNA-seq data \cite{tsdpp}.
-\\
-
-    We report the results using various Bayesian models to analyze gene expression data.
-Our exploration includes parallelized LDA, mixture models, dynamic-time topic models, topic-clustering models, and non-parametric models, implemented in \texttt{numpy}, \texttt{C++}, and \texttt{Edward}.
-We evaluate our methods using held-out likelihood, posterior predictive checks, and biological meaningfulness testing.
+    We first discuss our experiments using Bayesian topic models to discover related sets of genes in a 'topic': LDA in Section~\ref{ldasec} and Dynamic Time Models in Section~\ref{dtmsec}.
+We discuss our experiments in clustering: Dirichlet Mixtures in Section~\ref{mmsec}, Integrated Topic-Clustering in Section~\ref{intsec}, and non-parametric models in Section~\ref{nonparametricsec}.
+Finally, we conclude our paper with our obversations from across all studies.
 
 \section{Latent Dirichlet Allocation} 
+\label{ldasec}
 \subsection{Model Description} 
@@
 
+\begin{figure}[ht]
+\vskip 0.2in
+\begin{center}
+\centerline{\includegraphics[width=\columnwidth]{figs/examplegraph}}
+\caption{This is a demo figure.}
+\label{alabel}
+\end{center}
+\vskip -0.2in
+\end{figure}
+
 \section{Dirichlet Mixture Model} 
+\label{mmsec}
 \subsection{Model Description} 
@@
 \section{Dynamic Time Model} 
+\label{dtmsec}
 \subsection{Model Description} 
@@
 
-\section{Document Topic-Clustering Model} 
+\section{Integrated Topic-Clustering Model} 
+\label{intsec}
 \subsection{Model Description} 
@@
 \section{Non-parametric Models: IBP and HDP} 
+\label{nonparametricsec}
 \subsection{Model Description} 
@@
 
+\section{Appendix} 
+\subsection{Latent Dirichlet Allocation}
+
 \end{document}

### 2017-05-10 finalreport/paper.tex
update appendix, lda, intro

@@
 \subsection{Model Description} 
-    Algorithm~\ref{alg:example} describes the generative process for LDA.
+In the generative process for LDA, the topic assigned to a word in a document is a drawn from the document's topic distribution.
+The identity of the word is drawn from topic's word distribution.
+Assuming the reader is familiar with LDA, we relegate further details and formalization of the model to \cite{LDA}.
 
-    \begin{algorithm}[tb]
-       \caption{Latent Dirichlet Allocation}
-       \label{alg:example}
-    \begin{algorithmic}
-       \STATE {\bfseries Input:} data $x_i$, size $m$
-       \REPEAT
-       \STATE Initialize $noChange = true$.
-       \FOR{$i=1$ {\bfseries to} $m-1$}
-       \IF{$x_i > x_{i+1}$} 
-       \STATE Swap $x_i$ and $x_{i+1}$
-       \STATE $noChange = false$
-       \ENDIF
-       \ENDFOR
-       \UNTIL{$noChange$ is $true$}
-    \end{algorithmic}
-    \end{algorithm}
+In the context of analyzing gene expression data, we are interested in discovering 'topics' that comprise of a set of top-$N$ genes within the topic distribution that are biologically related.
+For example, together the genes may direct a specific chemical function in a cell.
+Biologists denote such sets of genes as 'gene modules' which can be cross-referenced with existing gene module databases.
+
 \subsection{Implementation}
+
 \subsection{Experiments}
 
-\begin{figure}[ht]
+\begin{figure}[h]
 \vskip 0.2in

### 2017-05-10 finalreport/paper.tex
finish lda shit

@@
 
+1.
+Here's the breakdown of work done insofar: LDA implementation, LDA experiments, and LDA biological relevance benchmarking was finished by Skanda (first two sections in the results).
+Karren completed the experiments involved HDP, the IBP-DP, Gene Set Enrichment implementation, and the IBP-DP experiments (second two sections in the results).
+
+Breakdown of work remaining: we plan on slightly shifting our milestones from our proposal.
+Specifically, we plan on shifting our more experimentation based milestones (e.g.
+how robust are these parametric/non-parametrics to data dropout, wrestling with pre-existing mixture model implementations to scale to our full datasets), and focus more on model implementation and formulation.
+Specifically, we were interested in jumping to our last two proposal points: formulating a dynamic topic
+model for RNA-seq [1], and finite-sized joint clustering and topic model [2].
+
+Specifically, in the upcoming work, there's two sub-investigations, which we were going to assign one to each of us.
+One person would formulate and implement the dynamic topic model with LDA to try on a subset of our RNA-Seq data.
+The other member would be responsible for formulating and implementing finite-sized joint clustering and topic model for our subset of our RNA-Seq data.
+
+2.
+More details about the Gibbs sampler: We had a fixed burn-in of number of samples (200), a fixed number of sampling iterations after that (500).
+We didn't extensively explore varying these values, but trying out significantly more iterations (700) didn't seem to change the topic's word distributions significantly.
+We could explore this in more detail for the final report.
+There was one sampling chain per core, with four cores.
+Not sure exactly what you had in mind by reporting
+'performance details', other than time for estimation/core, but maybe a somewhat more fair comparison, like test perplexity/instructions executed during posterior estimation?
+
+3.
+More details about biological significance: Good question about # of topics affecting the results for the hypergeometric test for biological relevance/correlation with existing biologically annotated.
+This is something I'm not too sure about, and why we experience the results we do.
+That said, # topics is something that we were experimented with, but soon realized we didn't consider a broad enough range, after seeing the results from our non-parametric model afterward.
+More
+investigation and thought will need to go into this for the final report.
+
 \section{Dynamic Time Model} 

### 2017-05-10 finalreport/paper.tex
finish update integration

--- 2017-05-10 finalreport/paper.tex
@@
 We favor this method because it is simple and easy to implement, and a practice used in literature \cite{pruning}}.
+The Seurat biological toolkit in R was also used for the purpose of low-variance feature selection \cite{seurat}.
 We eschewed dimensionality reduction techniques such as PCA because of loss of its direct feature interpretability.
+Biologists have hand-labeled every cell in our dataset using existing tool; there are a total of 9 cell categories.
 The complete dataset, as well our preprocessed and pruned versions, are available at \cite{dataset}.
@@
 \subsection{Implementation} 
+We explored two implementations of posterior inference for LDA on large datasets: one using online mean-field variational bayes for posterior estimate \cite{ovb, online}, and a broken \texttt{C++} Gibbs sampler for LDA that we modified to work \cite{plda}.
+We were able to get this latter sampler running parallel across multiple cores, with a burn-in of 100 iterations.
+We compare these two posterior estimation approaches using our entire dataset, with a 10\% held-out testing partition.
+We experimented with
+$k=5, 10, 25, 50$ topics. \\
+
+More details about the Gibbs sampler: We had a fixed burn-in of number of samples (200), a fixed number of sampling iterations after that (500).
+We didn't extensively explore varying these values, but trying out significantly more iterations (700) didn't seem to change the topic's word distributions significantly.
+We could explore this in more detail for the final report.
+There was one sampling chain per core, with four cores.
+Not sure exactly what you had in mind by reporting
+'performance details', other than time for estimation/core, but maybe a somewhat more fair comparison, like test perplexity/instructions executed during posterior estimation?
 
 \subsection{Experiments} 
+- posterior predictive checks
+- no intrusion testing meaningful
+- hypergeometric tests
+- likelihood
+- performance metrics
+
+We did gene set enrichment analysis using the minimum hypergeometric test \cite{hg} to compare our topics, which are ranked lists of genes, with existing collections of genes catalogued in the comprehensive gene module database, \texttt{MSigDB} \cite{msigdb}.
+Figure \ref{fig:pathways} in the Appendix shows topic-to-\texttt{MSigDB} module matches for which the $p$-value of the match is at least less than 0.3.
+Models with more topics tended to have more matches with higher significance.
+These results suggest that we need to train the model with more topics.
+We note that the $p$-values do \textit{not} factor for multiple hypothesis corrections, so at the moment we are only using them as relative measures of model quality.
+
+3.
+More details about biological significance: Good question about \# of topics affecting the results for the hypergeometric test for biological relevance/correlation with existing biologically annotated.
+This is something I'm not too sure about, and why we experience the results we do.
+That said, \# topics is something that we were experimented with, but soon realized we didn't consider a broad enough range, after seeing the results from our non-parametric model afterward.
+More
+investigation and thought will need to go into this for the final report.
+
+Figure \ref{fig:time} shows the time to complete each estimation method.
+As expected, Gibbs scales poorly with the parameter dimensionality and is strictly worse than online variational Bayes across all studied topic counts.
+We also calculated perplexity and log-likelihood for the OVB parameters, and very surprisingly, we found that log-likelihood decreased as number of topics increased on a held-out test set (Figure \ref{fig:ll}).
+We are currently trying to understand whether this is because of programming error.
 
@@
 \section{Appendix}
-\subsection{Latent Dirichlet Allocation}
+
+\subsection{Team's Breakdown of Work}
+The implementation(s) and experiments involving LDA, Dirichlet Mixture, and Dynamic Time Model was completed by Skanda (Section~\ref{ldasec}, Section~\ref{mmsec}, and Section~\ref{dtmsec}).
+Karren completed the implementations and experiments involved Integrated Topic-Clustering, Non-Parametric models, and the shared gene enrichment test implementation (Section~\ref{intsec} and Section~\ref{nonparametricsec}.
+We believe the authors contributed equally in this work.
+
+\subsection{LDA vs. IBP-DP}
+One drawback to the HDP is that there tends to be a correlation between how frequently a topic appears across all documents and how prevalent this topic is within documents that it appears in.
+Williamson et al.
+\cite{IBP} proposed the 'focused topic model' to overcome this drawback.
+In their model, each topic $k = 1,2,...$ has a relative prevalence $\phi_k \sim Gamma(\gamma, 1)$ and a population frequency $\pi_k = \prod^k_{j=1} \mu_k$, where each $\mu_k \sim Beta(\alpha, 1) $.
+For each document $m$, whether topic $k$ appears is sampled as $b_{mk} \sim Bernoulli(\pi_k)$, and the topic proportions are sampled as $\theta_m \sim Dirichlet(b_m \cdot \phi)$.
+\\
+
+Since the code from the original IBP-DP paper was not available, we implemented an inference algorithm using collapsed Gibbs sampling \cite{IBP}.
+Due to the non-conjugacy of the model, sampling each latent variable from its full conditional required using another sampling method.
+To sample the topics parameters $\pi$ and $\phi$, we used slice sampling based on the semi-ordered stick-breaking representation of the model \cite{IBP2}.\\
+
+We tested our code on a subset of the Reuters-21578 dataset, using several different values of the concentration hyper-parameter $\alpha$, which influences the number of clusters.
+Although higher values of $\alpha$ yielded better log-likelihood values, we found that it resulted in a large number of very small topics, which are not very useful (Figure \ref{fig:ibp_words_in_topic}).
+Qualitatively, we did not find the topics from IBP-DP (Figure \ref{fig:ibptopics}) to be more coherent than topics than LDA (Figure \ref{fig:ldatopics}).
+The most prevalent topics from the IBP-DP each corresponded to similar topics from LDA; less prevalent topics tended to consist of a few unrelated words.
+These results discouraged us from optimizing the code to train this model on our scRNA-seq dataset, as we do not think it would yield more coherent gene modules than LDA.
+We emphasize that these results are not completely unexpected, as the authors of the IBP-DP paper did not show any topics from their model, nor did they assess the quality of their model with metrics other than perplexity.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/time}
+    \caption{Comparison of the running times of each of the posterior estimation methods across various numbers of topics.}
+    \label{fig:time}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/ll}
+    \caption{Log-likelihood of the held-out testing set, across various numbers of topics.}
+    \label{fig:ll}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1\textwidth]{figs/pathways}
+    \caption{Matches between the gene collections found in LDA topics and published gene sets in \texttt{MSigDB}. Cluster size refers to the number of topics in the model}
+    \label{fig:pathways}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/hdp-perplexity}
+    \caption{Comparison of log-likelihood of the held-out testing set, under various LDA models and the HDP model.}
+    \label{fig:hdp-perplexity}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/hdp-enrichment}
+    \caption{Comparison of distributions of p-values from gene set enrichment analysis between LDA models and the HDP model.}
+    \label{fig:hdp-enrichment}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/ibp_words_in_topic}
+    \caption{Beeswarm plot of number of words per topic, for 3 different IBP-DP models with different concentration parameters. Each point represents one topic from its model}
+    \label{fig:ibp_words_in_topic}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1\textwidth]{figs/ibptopics}
+    \caption{Topics from IBP-DP model trained on subset of Reuters dataset.}
+    \label{fig:ibptopics}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1\textwidth]{figs/ldatopics}
+    \caption{Topics from LDA model with 15 topics trained on subset of Reuters dataset.}
+    \label{fig:ldatopics}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.3\textwidth]{figs/mixture}
+    \caption{Finite \texttt{K}-sized mixture model currently implemented.
+$\theta$ is the parameter for every cluster component, represented from a categorical draw of over all genes.
+$z_i$ is the cluster assignment, and $\phi$ is the distribution of clusters.
+As usual, $\alpha, \beta$ are hyper-parameters.}
+    \label{fig:plate}
+\end{figure}
 

### 2017-05-10 finalreport/paper.tex
make some fixes

@@
 We eschewed dimensionality reduction techniques such as PCA because of loss of its direct feature interpretability.
-Biologists have hand-labeled every cell in our dataset using existing tool; there are a total of 9 cell categories.
-The complete dataset, as well our preprocessed and pruned versions, are available at \cite{dataset}.
+Prior researchers have labeled the cells in our dataset; there are a total of 9 cell categories.
+The complete dataset, as well our preprocessed and pruned versions, are openly available \cite{dataset}.
 
@@
 Yu et al.
-propose a unsupervised classifier ensemble as another approach to cell clustering \cite{consensus}.
+propose an unsupervised classifier ensemble as another approach to cell clustering \cite{consensus}.
 
@@
 Bayesian networks have also been used in an attempt to learn gene dependencies from expression data \cite{bayesge3}.
+Our work uses different models to explore gene expression data, but where appropriate (e.g.
+in mixture models in Section~\ref{mmsec}), we compare results.
 
@@
 We first discuss our experiments using Bayesian topic models to discover related sets of genes in a 'topic': LDA in Section~\ref{ldasec} and Dynamic Time Models in Section~\ref{dtmsec}.
-We discuss our experiments in clustering: Dirichlet Mixtures in Section~\ref{mmsec}, Integrated Topic-Clustering in Section~\ref{intsec}, and non-parametric models in Section~\ref{nonparametricsec}.
-Finally, we conclude our paper with our obversations from across all studies.
+Then, we discuss our experiments in clustering: Dirichlet Mixtures in Section~\ref{mmsec}, Integrated Topic-Clustering in Section~\ref{intsec}, and non-parametric models in Section~\ref{nonparametricsec}.
+We conclude our paper with our observations from across all our studies.
+
+For the purpose of reproducibility, all code can be found at \url{https://github.com/skoppula/882}.
 
@@
 \subsection{Model Description} 
-In the generative process for LDA, the topic assigned to a word in a document is a drawn from the document's topic distribution.
+In the generative process for LDA, the topic assigned to each word is a drawn from the document's topic distribution.
 The identity of the word is drawn from topic's word distribution.
@@
 \subsection{Implementation} 
-We explored two implementations of posterior inference for LDA on large datasets: one using online mean-field variational bayes for posterior estimate \cite{ovb, online}, and a broken \texttt{C++} Gibbs sampler for LDA that we modified to work \cite{plda}.
-We were able to get this latter sampler running parallel across multiple cores, with a burn-in of 100 iterations.
-We compare these two posterior estimation approaches using our entire dataset, with a 10\% held-out testing partition.
-We experimented with
-$k=5, 10, 25, 50$ topics. \\
+A first attempt using the built-in Python \texttt{lda} package resulting in early memory overflows during what we suspect was pre-allocation of per-document variables.
+The source code was not available, so we had few clues.
 
-More details about the Gibbs sampler: We had a fixed burn-in of number of samples (200), a fixed number of sampling iterations after that (500).
-We didn't extensively explore varying these values, but trying out significantly more iterations (700) didn't seem to change the topic's word distributions significantly.
-We could explore this in more detail for the final report.
-There was one sampling chain per core, with four cores.
-Not sure exactly what you had in mind by reporting
-'performance details', other than time for estimation/core, but maybe a somewhat more fair comparison, like test perplexity/instructions executed during posterior estimation?
+We switched to two open-source implementations: an online mean-field variational Bayes for posterior estimation \cite{ovb}, and a broken \texttt{C++} Gibbs sampler for LDA \cite{plda}.
+
+\nocite{online}
+
+We fixed portions of the sampler to compile properly and extended the sampler to run in across four cores.
+Details of our sampling procedure can be found in Appendix~\ref{ldaappendix}.
+We compared these two posterior estimation approaches using our entire dataset, using a 10\% held-out testing partition.
+We experimented using $k=5, 10, 25, 50$ topics.
+We did not require dataset pruning after these optimizations.
 
@@
 \subsection{Experiments}  
+- comparison with pe'er paper
 
@@
 We believe the authors contributed equally in this work.
+
+\subsection{Latent Dirichlet Allocation}
+\label{ldaappendix}
+For our Gibb's sampler, we had a fixed burn-in of number of samples (200), a fixed number of sampling iterations after that (500).
+We didn't extensively explore varying these values, but trying out significantly more iterations (700) didn't seem to change the topic's word distributions significantly.
+There was one sampling chain on each of four cores.
 

### 2017-05-10 finalreport/paper.tex
update lda

@@
 
-We fixed portions of the sampler to compile properly and extended the sampler to run in across four cores.
+We fixed portions of the sampler to compile properly and extended the sampler to run across four cores.
 Details of our sampling procedure can be found in Appendix~\ref{ldaappendix}.
@@
 \subsection{Experiments} 
+In lieue of intrusion testing, we evaluated the interpretability of our topics (ranked lists of genes) using using hypergeometric tests \cite{hg}.
+In brief, this determines the probability that our sets of genes would be clustered together in accordance to with existing collections of genes catalogued by biologists in the gene module database \texttt{MSigDB} \cite{msigdb}.
+
+Figure~\ref{fig:pathways} in the Appendix shows gene module matches in \texttt{MSigDB} for which the $p$-value of the match is at least less than 0.3.
+Notice that models with more topics tended to have more matches with higher significance.
+The $p$-values in our tests do \textit{not} factor for multiple hypothesis corrections, so at the moment we are only using them as relative measures of model quality.
+
+It is validating to see that many of the similar modules, such as the \texttt{BRCA1} and \texttt{DOXORUBICIN}, are cancer related, given that our expression assay was from tumor cells.
+
+Extracting the parameters of each model, we calculated log-likelihood for on a held-out test set.
+This is shown in Figure \ref{fig:ll}.
+We found that for this dataset increasing the topic count increases log-likelihood, roughly logarithmically from 5 to 50.
+Further experiments will need to test at what point log-likelihood drops off (one way to optimize the number of topics).
+
+\begin{figure}
+\vskip 0.2in
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/ll}
+    \caption{Log-likelihood of the held-out testing set, across various numbers of topics.}
+    \label{fig:ll}
+\vskip -0.2in
+\end{figure}
+
+As an interesting aside, Figure \ref{fig:time} shows the time until inference completion (collected during our experiments).
+Gibbs appears to scale poorly with the parameter dimensionality, in contrast to online variational Bayes \footnote{It is important to note that comparing the magnitude of the time to complete inference is not particularly meaningful; as described in Appendix~\ref{ldaappendix}, we chose a reasonable guess for the number of iterations after which to stop estimation}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/time}
+    \caption{Comparison of the running times of each of the posterior estimation methods across various numbers of topics.}
+    \label{fig:time}
+\end{figure}
+
+Posterior predictive checks to test the mutual information between the each cell and its words' topics is something that we did not have time, but would be interesting to examine.
+It's not clear that this independence assumption is true in the context of gene expression data, because of cross-talk and regulatory mechanisms between gene modules.
+
 - posterior predictive checks
@@
 - performance metrics
-
-We did gene set enrichment analysis using the minimum hypergeometric test \cite{hg} to compare our topics, which are ranked lists of genes, with existing collections of genes catalogued in the comprehensive gene module database, \texttt{MSigDB} \cite{msigdb}.
-Figure \ref{fig:pathways} in the Appendix shows topic-to-\texttt{MSigDB} module matches for which the $p$-value of the match is at least less than 0.3.
-Models with more topics tended to have more matches with higher significance.
-These results suggest that we need to train the model with more topics.
-We note that the $p$-values do \textit{not} factor for multiple hypothesis corrections, so at the moment we are only using them as relative measures of model quality.
-
-3.
-More details about biological significance: Good question about \# of topics affecting the results for the hypergeometric test for biological relevance/correlation with existing biologically annotated.
-This is something I'm not too sure about, and why we experience the results we do.
-That said, \# topics is something that we were experimented with, but soon realized we didn't consider a broad enough range, after seeing the results from our non-parametric model afterward.
-More
-investigation and thought will need to go into this for the final report.
-
-Figure \ref{fig:time} shows the time to complete each estimation method.
-As expected, Gibbs scales poorly with the parameter dimensionality and is strictly worse than online variational Bayes across all studied topic counts.
-We also calculated perplexity and log-likelihood for the OVB parameters, and very surprisingly, we found that log-likelihood decreased as number of topics increased on a held-out test set (Figure \ref{fig:ll}).
-We are currently trying to understand whether this is because of programming error.
-
-\begin{figure}[h]
-\vskip 0.2in
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{figs/examplegraph}}
-\caption{This is a demo figure.}
-\label{alabel}
-\end{center}
-\vskip -0.2in
-\end{figure}
 
@@
 
-\subsection{Team's Breakdown of Work}
-The implementation(s) and experiments involving LDA, Dirichlet Mixture, and Dynamic Time Model was completed by Skanda (Section~\ref{ldasec}, Section~\ref{mmsec}, and Section~\ref{dtmsec}).
+\subsection{Breakdown of Work}
+The implementation(s) and experiments involving LDA, Dirichlet Mixtures, and Dynamic Time Model was completed by Skanda (Section~\ref{ldasec}, Section~\ref{mmsec}, and Section~\ref{dtmsec}).
 Karren completed the implementations and experiments involved Integrated Topic-Clustering, Non-Parametric models, and the shared gene enrichment test implementation (Section~\ref{intsec} and Section~\ref{nonparametricsec}.
@@
 There was one sampling chain on each of four cores.
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=1\textwidth]{figs/pathways}
+    \caption{Matches between the gene collections found in LDA topics and published gene sets in \texttt{MSigDB}. Cluster size refers to the number of topics in the model}
+    \label{fig:pathways}
+\end{figure*}
 
@@
 We emphasize that these results are not completely unexpected, as the authors of the IBP-DP paper did not show any topics from their model, nor did they assess the quality of their model with metrics other than perplexity.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{figs/time}
-    \caption{Comparison of the running times of each of the posterior estimation methods across various numbers of topics.}
-    \label{fig:time}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{figs/ll}
-    \caption{Log-likelihood of the held-out testing set, across various numbers of topics.}
-    \label{fig:ll}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{figs/pathways}
-    \caption{Matches between the gene collections found in LDA topics and published gene sets in \texttt{MSigDB}. Cluster size refers to the number of topics in the model}
-    \label{fig:pathways}
-\end{figure}
 

### 2017-05-11 finalreport/paper.tex
do dtm section

@@
 \usepackage{subfigure} 
+\usepackage{amsfonts}
 
@@
 \subsection{Experiments} 
-In lieue of intrusion testing, we evaluated the interpretability of our topics (ranked lists of genes) using using hypergeometric tests \cite{hg}.
+In lieu of intrusion testing, we evaluated the interpretability of our topics (ranked lists of genes) using using hypergeometric tests \cite{hg}.
 In brief, this determines the probability that our sets of genes would be clustered together in accordance to with existing collections of genes catalogued by biologists in the gene module database \texttt{MSigDB} \cite{msigdb}.
@@
 
-Extracting the parameters of each model, we calculated log-likelihood for on a held-out test set.
+Extracting the parameters of each model, we calculated log-likelihood on our held-out test set.
 This is shown in Figure \ref{fig:ll}.
-We found that for this dataset increasing the topic count increases log-likelihood, roughly logarithmically from 5 to 50.
-Further experiments will need to test at what point log-likelihood drops off (one way to optimize the number of topics).
+We found that for this dataset increasing the topic count increases log-likelihood, roughly linear from 5 to 50.
+Further experiments will need to test at what point log-likelihood drops off; this is one way to optimize the number of topics.
 
 \begin{figure}
-\vskip 0.2in
     \centering
-    \includegraphics[width=0.5\textwidth]{figs/ll}
-    \caption{Log-likelihood of the held-out testing set, across various numbers of topics.}
+    \includegraphics[width=0.4\textwidth]{figs/ll}
+    \caption{Log-likelihood of the held-out testing set across numbers of topics. Higher topic sizes are explored in later models.}
     \label{fig:ll}
@@
 As an interesting aside, Figure \ref{fig:time} shows the time until inference completion (collected during our experiments).
-Gibbs appears to scale poorly with the parameter dimensionality, in contrast to online variational Bayes \footnote{It is important to note that comparing the magnitude of the time to complete inference is not particularly meaningful; as described in Appendix~\ref{ldaappendix}, we chose a reasonable guess for the number of iterations after which to stop estimation}.
+Gibbs appears to scale poorly with the parameter dimensionality, in contrast to online variational Bayes \footnote{It is important to note that comparing the magnitude of the time to complete inference in Figure~\ref{fig:time} is not particularly meaningful; as described in Appendix~\ref{ldaappendix}, we chose a reasonable guess for the number of iterations
+and optimization threshold after which to stop estimation}.
 
@@
     \centering
-    \includegraphics[width=0.5\textwidth]{figs/time}
-    \caption{Comparison of the running times of each of the posterior estimation methods across various numbers of topics.}
+    \includegraphics[width=0.4\textwidth]{figs/time}
+    \caption{Comparison of the running times of each of the posterior estimation methods across various numbers of topics.
+Termination conditions across all runs were constant, and described in Appendix~\ref{ldaappendix}}
     \label{fig:time}
@@
 It's not clear that this independence assumption is true in the context of gene expression data, because of cross-talk and regulatory mechanisms between gene modules.
+
+\section{Dynamic Topic Model} 
+\label{dtmsec}
+\subsection{Model Description} 
+As tumor cells proliferate, they undergo a process called \textit{differentiation}.
+The set of cells types during tumor emergence may differ from the set of types expressed days later.
+Correspondingly, patterns of gene expression may also change over time.
+The set of gene modules expressed in the cells, or the composition of each gene module (topic), may also change.
+
+To capture these evolving topics, we look to dynamic topic models \cite{dtm}.
+In brief, dynamic topic models establish a conditional distribution over the hyperparameters $\alpha$ and $\beta$, that govern the document's topic distribution and topic's word distribution, respectively.
+The distribution over each hyperparameter is conditioned on the prior in the previous time step, allowing changes to topic composition and assignment.
+This can be seen graphically in the plate
+model in Figure~\ref{fig:dtmplate}.
+For convenience, we provide a concise description of the generative process for the implemented heirarchical model for a time slice $t$ in Appendix~\ref{dtmappendix}.
+
+\begin{figure}[h]
+    \vspace{-0.1in}
+    \centering
+    \includegraphics[width=0.4\textwidth]{figs/dtmplate}
+    \caption{Plate diagram for the Dynamic Topic Model. Reproduced from \cite{dtm}.}
+    \label{fig:dtmplate}
+    \vspace{-0.1in}
+\end{figure}
+
+Note that in contrast to LDA, dynamic topic models use a logistic-normal to express proportion uncertainty. This is stated more explicitly in Appendix~\ref{dtmappendix}.
+
+\subsection{Implementation}
+
+Unfortunately, due to the non-conjugacy of the Gaussian/Multinomial in our logistic-normal setup, integrating out parameters for any sort of Gibbs sampling becomes hard.
+Instead, the original DTM paper uses a variational approximation based on a Kalman filter, that preserves time dependencies, unlike a mean-field approximation.
+
+Re-implementing the variational approximation, while educational and interesting, would be a project of itself, so our first attempt in employing dynamic time models to our gene expression data re-used the Kalman Filter-based variational inference used by the original authors, published recently by the Blei Lab \cite{dtmbleigit}.
+The results we show in the subsequent section are using a wrapper we've implemented around the group's inference code.
+
+The authors stumbled upon a recent arXiv pre-print that proposed a set of sampler update rules to create a correct DTM Gibbs sampler \cite{sdtm}.
+Using the probabalistic programming framework \texttt{Edward}, and modifying the in-built sampler to follow these updates, we were able to obtain a working Gibbs-sampler for sample time-sliced data for a dynamic topic model defined in Edward \cite{edward}.
+While we don't have time to conduct benchmarks or re-do results right now, after verifying that results are consistent, the authors hope to submit a merge for this Edward sketch into the mainstream Edward examples library this coming summer.
+
+\subsection{Experiments} 
+We subdivided our data into three seperate time slices, and pruned each slice, retaining the top 30\% of high-variance genes. We experimented using $k=15$, $30$, and $50$ topics.
+
+Similar to our analysis in Section~\ref{ldasec}, we used the hypergeometric test to evaluate interpretability.
+
+Figure~\ref{fig:pathwaysdtm} in the Appendix shows gene module matches for which the $p$-value of the match is at least less than 0.3, across the three time slices for each of the three topic count sizes.
+
+Notice that models with more topics tended to have more matches with higher significance.
+The $p$-values in our tests do \textit{not} factor for multiple hypothesis corrections, so at the moment we are only using them as relative measures of model quality.
+
+It is validating to see that many of the similar modules, such as the \texttt{BRCA1} and \texttt{DOXORUBICIN}, are cancer related, given that our expression assay was from tumor cells.
 
@@
 \subsection{Model Description} 
-    Algorithm~\ref{alg:example} describes the generative process for the mixture odel.
-
-    \begin{algorithm}[tb]
-       \caption{Mixture Model}
-       \label{alg:example}
-    \begin{algorithmic}
-       \STATE {\bfseries Input:} data $x_i$, size $m$
-       \REPEAT
-       \STATE Initialize $noChange = true$.
-       \FOR{$i=1$ {\bfseries to} $m-1$}
-       \IF{$x_i > x_{i+1}$} 
-       \STATE Swap $x_i$ and $x_{i+1}$
-       \STATE $noChange = false$
-       \ENDIF
-       \ENDFOR
-       \UNTIL{$noChange$ is $true$}
-    \end{algorithmic}
-    \end{algorithm}
 \subsection{Implementation} 
@@
 - comparison with pe'er paper
-
-\section{Dynamic Time Model} 
-\label{dtmsec}
-\subsection{Model Description} 
-\subsection{Implementation} 
-\subsection{Experiments}
 
@@
 
+\section{Conclusions} 
+\label{nonparametricsec}
+
 % In the unusual situation where you want a paper to appear in the
@@
 \subsection{Breakdown of Work}
-The implementation(s) and experiments involving LDA, Dirichlet Mixtures, and Dynamic Time Model was completed by Skanda (Section~\ref{ldasec}, Section~\ref{mmsec}, and Section~\ref{dtmsec}).
-Karren completed the implementations and experiments involved Integrated Topic-Clustering, Non-Parametric models, and the shared gene enrichment test implementation (Section~\ref{intsec} and Section~\ref{nonparametricsec}.
-We believe the authors contributed equally in this work.
+The implementation(s) and experiments involving LDA, Dirichlet Mixtures, and Dynamic Time Model was completed by SK (Section~\ref{ldasec}, Section~\ref{mmsec}, and Section~\ref{dtmsec}).
+KY completed the implementations and experiments involved Integrated Topic-Clustering, Non-Parametric models, and the shared gene enrichment test implementation (Section~\ref{intsec} and Section~\ref{nonparametricsec}.
+The authors contributed equally in this work.
 
@@
     \includegraphics[width=1\textwidth]{figs/pathways}
-    \caption{Matches between the gene collections found in LDA topics and published gene sets in \texttt{MSigDB}. Cluster size refers to the number of topics in the model}
+    \caption{Matches between the gene collections found in LDA topics and published gene sets in \texttt{MSigDB}. `Cluster size' refers to the number of topics in the model}
     \label{fig:pathways}
 \end{figure*}
+
+\subsection{Dynamic Topic Model}
+\label{dtmappendix}
+Here is a description of the generative process for our dynamic topic model:
+\begin{enumerate}
+    \item Draw a new topic composition hyperparameter \\ $\beta \leftarrow \mathbb{N}(\beta_{t-1}, \sigma I^2)$
+    \item Draw a new document composition hyperparameter \\ $\alpha \leftarrow \mathbb{N}(\alpha_{t-1}, \delta I^2)$
+    \item For each document: 
+    \begin{enumerate}
+        \item Draw a new document topic distribution \\ $\theta \leftarrow \pi(\mathbb{N}(\alpha_t))$
+        \item For every word in the document:
+        \begin{enumerate}
+            \item Draw the word topic assignment \\ $Z \leftarrow$ Mult$(\theta)$
+            \item Draw the word identity \\ $Z \leftarrow$ Mult$(\pi(\beta_{t,z}))$
+        \end{enumerate}
+    \end{enumerate}
+\end{enumerate}
+Here, $\pi(x_i)$ is the softmax function $\frac{exp(x_i)}{\sum_k exp(x_k)}$ \cite{dtm}.
 

### 2017-05-11 finalreport/paper.tex
finished dtm

@@
 Using the probabalistic programming framework \texttt{Edward}, and modifying the in-built sampler to follow these updates, we were able to obtain a working Gibbs-sampler for sample time-sliced data for a dynamic topic model defined in Edward \cite{edward}.
-While we don't have time to conduct benchmarks or re-do results right now, after verifying that results are consistent, the authors hope to submit a merge for this Edward sketch into the mainstream Edward examples library this coming summer.
+While we don't have time to conduct benchmarks or re-run our results right now, after verifying that results are consistent, the authors hope to submit a merge for this Edward sketch into the mainstream Edward examples library this coming summer.
 
 \subsection{Experiments} 
-We subdivided our data into three seperate time slices, and pruned each slice, retaining the top 30\% of high-variance genes. We experimented using $k=15$, $30$, and $50$ topics.
+We subdivided our data into three seperate time slices, and pruned each slice, retaining the top 20\% of high-variance genes. We experimented using $k=15$, $30$, and $50$ topics.
 
@@
 
-Figure~\ref{fig:pathwaysdtm} in the Appendix shows gene module matches for which the $p$-value of the match is at least less than 0.3, across the three time slices for each of the three topic count sizes.
+Figure~\ref{fig:pathwaysdtm} in the Appendix shows gene module matches for which the $p$-value is less than 0.3, across the three time slices and the three topic count sizes.
+Interestingly, we notice a very different set of matched gene modules, with exception of the \texttt{KLF1} module\footnote{This could very well be a result of our dataset pruning, which we did to keep inference runtimes manageable}.
+Like before, we notice that some of the identified topics are relevant to tumor cells:
+\texttt{ST\_TUMOR\_NECROSIS\_FACTOR\_PATHWAY} and \texttt{WEST\_ADRENOCORTICAL\_TUMOR\_MARKERS\_UP} to name the two most prominent topics.
 
-Notice that models with more topics tended to have more matches with higher significance.
-The $p$-values in our tests do \textit{not} factor for multiple hypothesis corrections, so at the moment we are only using them as relative measures of model quality.
+We do not see that much variation across time-slices, demonstrating that our time-slice partitions are largely uniform.
+For example, for model $k=50$, we see that topic 20 consistently matches with MSigDB module \texttt{PILON\_KLF1\_TARGETS\_UP}, suggesting that its gene composition is not varying.
 
-It is validating to see that many of the similar modules, such as the \texttt{BRCA1} and \texttt{DOXORUBICIN}, are cancer related, given that our expression assay was from tumor cells.
+Models with more topics tended to have more matches with higher significance.
+But again, the $p$-values in our tests do \textit{not} factor for multiple hypothesis corrections, so at the moment we are only using them as relative measures of model quality.
 
-- posterior predictive checks
-- no intrusion testing meaningful
-- hypergeometric tests
-- likelihood
-- performance metrics
+After extracting the mode of learned parameters, we evaluated the log-likelihood of our held-out test set, also partitioned by time.
+This is shown in Figure \ref{fig:lldtm}.
+We find a leveling of log-likelihood after $k=30$.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.4\textwidth]{figs/lldtm}
+    \caption{Log-likelihood of the held-out testing set across our time slices for varying numbers of topics.}
+    \label{fig:lldtm}
+\vskip -0.2in
+\end{figure}
+
+With more time, we would ideally repeat the experiments to obtain confidence intervals on these experiments, and examine the gene-module/cell/time-slice mutual information.
 
@@
 The implementation(s) and experiments involving LDA, Dirichlet Mixtures, and Dynamic Time Model was completed by SK (Section~\ref{ldasec}, Section~\ref{mmsec}, and Section~\ref{dtmsec}).
-KY completed the implementations and experiments involved Integrated Topic-Clustering, Non-Parametric models, and the shared gene enrichment test implementation (Section~\ref{intsec} and Section~\ref{nonparametricsec}.
+KY completed the implementations and experiments involved Integrated Topic-Clustering, Non-Parametric models, and the shared gene enrichment test implementation (Section~\ref{intsec} and Section~\ref{nonparametricsec}).
 The authors contributed equally in this work.
@@
 
-\subsection{LDA vs. IBP-DP}
-One drawback to the HDP is that there tends to be a correlation between how frequently a topic appears across all documents and how prevalent this topic is within documents that it appears in.
-Williamson et al.
-\cite{IBP} proposed the 'focused topic model' to overcome this drawback.
-In their model, each topic $k = 1,2,...$ has a relative prevalence $\phi_k \sim Gamma(\gamma, 1)$ and a population frequency $\pi_k = \prod^k_{j=1} \mu_k$, where each $\mu_k \sim Beta(\alpha, 1) $.
-For each document $m$, whether topic $k$ appears is sampled as $b_{mk} \sim Bernoulli(\pi_k)$, and the topic proportions are sampled as $\theta_m \sim Dirichlet(b_m \cdot \phi)$.
-\\
-
-Since the code from the original IBP-DP paper was not available, we implemented an inference algorithm using collapsed Gibbs sampling \cite{IBP}.
-Due to the non-conjugacy of the model, sampling each latent variable from its full conditional required using another sampling method.
-To sample the topics parameters $\pi$ and $\phi$, we used slice sampling based on the semi-ordered stick-breaking representation of the model \cite{IBP2}.\\
-
-We tested our code on a subset of the Reuters-21578 dataset, using several different values of the concentration hyper-parameter $\alpha$, which influences the number of clusters.
-Although higher values of $\alpha$ yielded better log-likelihood values, we found that it resulted in a large number of very small topics, which are not very useful (Figure \ref{fig:ibp_words_in_topic}).
-Qualitatively, we did not find the topics from IBP-DP (Figure \ref{fig:ibptopics}) to be more coherent than topics than LDA (Figure \ref{fig:ldatopics}).
-The most prevalent topics from the IBP-DP each corresponded to similar topics from LDA; less prevalent topics tended to consist of a few unrelated words.
-These results discouraged us from optimizing the code to train this model on our scRNA-seq dataset, as we do not think it would yield more coherent gene modules than LDA.
-We emphasize that these results are not completely unexpected, as the authors of the IBP-DP paper did not show any topics from their model, nor did they assess the quality of their model with metrics other than perplexity.
-
-\begin{figure}
+\begin{figure*}
     \centering
-    \includegraphics[width=0.5\textwidth]{figs/hdp-perplexity}
-    \caption{Comparison of log-likelihood of the held-out testing set, under various LDA models and the HDP model.}
-    \label{fig:hdp-perplexity}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{figs/hdp-enrichment}
-    \caption{Comparison of distributions of p-values from gene set enrichment analysis between LDA models and the HDP model.}
-    \label{fig:hdp-enrichment}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{figs/ibp_words_in_topic}
-    \caption{Beeswarm plot of number of words per topic, for 3 different IBP-DP models with different concentration parameters. Each point represents one topic from its model}
-    \label{fig:ibp_words_in_topic}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{figs/ibptopics}
-    \caption{Topics from IBP-DP model trained on subset of Reuters dataset.}
-    \label{fig:ibptopics}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{figs/ldatopics}
-    \caption{Topics from LDA model with 15 topics trained on subset of Reuters dataset.}
-    \label{fig:ldatopics}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.3\textwidth]{figs/mixture}
-    \caption{Finite \texttt{K}-sized mixture model currently implemented.
-$\theta$ is the parameter for every cluster component, represented from a categorical draw of over all genes.
-$z_i$ is the cluster assignment, and $\phi$ is the distribution of clusters.
-As usual, $\alpha, \beta$ are hyper-parameters.}
-    \label{fig:plate}
-\end{figure}
+    \includegraphics[width=1\textwidth]{figs/pathwaysdtm}
+    \caption{Matches between the gene collections found using a dynamic topic model and published gene sets in \texttt{MSigDB}.
+$k$ refers to the number of topics in the model.
+$t$ refers to the time slice in the model.}
+    \label{fig:pathwaysdtm}
+\end{figure*}
 

### 2017-05-11 finalreport/paper.tex
start mm

@@
 \usepackage{graphicx} % more modern
+\usepackage{booktabs}
 \usepackage{subfigure} 
@@
 
-\section{Dirichlet Mixture Model} 
+\section{Mixture Model} 
 \label{mmsec}
 \subsection{Model Description} 
+The second question we wanted to address
+- comparison with pe'er paper
 \subsection{Implementation} 
 \subsection{Experiments}  
-- comparison with pe'er paper
+- 500 sampling iterations per run
+\begin{table}[h]
+    \centering
+    \label{mmtable}
+    \begin{tabular}{@{}ll@{}}
+        \toprule
+        \# of Clusters & \# of Clusters Used \\ \midrule
+        10                 & 3                                                         \\
+        25                 & 9                                                         \\
+        50                 & 16                                                        \\
+        75                 & 18                                                        \\
+        100                & 20                                                        \\ \bottomrule
+    \end{tabular}
+    \caption{The number of clusters defined in the model appears to be more than the actual number of unique clusters assigned to the data points after posterior estimation.
+The estimated true number of categories is slightly above 10.}
+\end{table}
 
@@
 
+\section{Breakdown of Work}
+The implementation(s) and experiments involving LDA, Gaussian Mixtures, and Dynamic Time Model was completed by SK (Section~\ref{ldasec}, Section~\ref{mmsec}, and Section~\ref{dtmsec}).
+KY completed the implementations and experiments involved Integrated Topic-Clustering, non-parametric models, (Section~\ref{intsec} and Section~\ref{nonparametricsec}) and the shared gene enrichment test implementation.
+The authors contributed equally in this work.
+
 \bibliography{citations}
@@
 
-\subsection{Breakdown of Work}
-The implementation(s) and experiments involving LDA, Dirichlet Mixtures, and Dynamic Time Model was completed by SK (Section~\ref{ldasec}, Section~\ref{mmsec}, and Section~\ref{dtmsec}).
-KY completed the implementations and experiments involved Integrated Topic-Clustering, Non-Parametric models, and the shared gene enrichment test implementation (Section~\ref{intsec} and Section~\ref{nonparametricsec}).
-The authors contributed equally in this work.
-
 \subsection{Latent Dirichlet Allocation}
 \label{ldaappendix}
+
+This section includes Figure~\ref{fig:pathways} on the next page.
+
 For our Gibb's sampler, we had a fixed burn-in of number of samples (200), a fixed number of sampling iterations after that (500).
@@
 \label{dtmappendix}
+
+This section includes Figure~\ref{fig:pathwaysdtm} on the next page.
+
 Here is a description of the generative process for our dynamic topic model:
@@
 
+\subsection{Gaussian Mixture Model}
+
+For our Gibb's sampler, we had a fixed burn-in of number of samples (200), a fixed number of sampling iterations after that (500).
+For time reasons, we didn't extensively explore varying these values.
+Figure~\ref{fig:llmm} describes the log-likelihood evaluation of the estimated posteriors.
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=1\textwidth]{figs/llmm}
+    \caption{Log-likelihood of held-out test set, on three seperate runs of Gibbs sampling procedure, described above}
+    \label{fig:llmm}
+\end{figure*}
+
+\label{mmappendix}
+
 \end{document}

### 2017-05-11 finalreport/paper.tex
finish mm

@@
 \subsection{Model Description} 
-The second question we wanted to address
-- comparison with pe'er paper
+The second question -- whether we could learn cell groupings -- we tackled using a canonical Gaussian mixture model.
+In brief, the model assigns every data point to a cluster; every cluster is Gaussian whose parameters are learned through posterior inference.
+A plate diagram can be found at \cite{mmplate}.
+This model is different from the previously mentioned Pe'er papers which use Dirichlet Processes Mixtures to study expression data.
+By plotting
+histograms of raw gene expression values, we've found that for many genes the distribution is mostly normal across our data points.
+This hints that the generative process in a Gaussian mixture model is plausible.
+
 \subsection{Implementation} 
+We implemented the mixture model in \textbf{Edward}, the probabalistic programming framework on top of Tensorflow.
+This allowed us to experiment with different samplers and variational inference approximations.
+The results we show below are for our runs using a Gibbs sampler, with a fixed burn-in of number of samples (200), a fixed number of sampling iterations (500).
+
 \subsection{Experiments}  
-- 500 sampling iterations per run
+% As a pre-processing step, we normalized every gene to zero-mean, unit variance across the data points.
+To speed up inference, we cut the dimensionality of the data to 10 using PCA before feeding it into our mixture model.
+
+Figure~\ref{fig:mmplot} shows an estimated cluster assignment for one-hundred sampled data point.
+We see that while our learned assignments do capture some of the clustering, there is significant bias toward one group (the burgundy cluster).
+
+Following this trend, we examined the set of unique clusters assigned in the posterior.
+The results are shown in Table~\ref{mmtable}.
+Interestingly, the model consistently uses less clusters than is allocated in the model.
+We are unsure why this is the case.
+
+Finally, we examine the log-likelihood of a held-out test set three different runs of posterior estimation with our sampler.
+This is shown in Figure~\ref{fig:llmm}.
+The runs are consistent with each other, and the likelihood is largely invariant across the five different cluster counts we tried, $k=\{10,25,50,75,50\}$.
+There are an estimated 9 true cell types in the original dataset, which could explain this result.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.4\textwidth]{figs/llmm}
+    \caption{Log-likelihood of held-out test set, on three seperate runs of Gibbs sampling procedure, described above}
+    \label{fig:llmm}
+\end{figure}
+
 \begin{table}[h]
@@
     \caption{The number of clusters defined in the model appears to be more than the actual number of unique clusters assigned to the data points after posterior estimation.
-The estimated true number of categories is slightly above 10.}
+The estimated true number of categories is 9.}
 \end{table}
+
+\begin{figure*}[ht]
+    \centering
+    \includegraphics[width=0.9\textwidth]{figs/totalclusters}
+    \caption{2D-projection of sampled gene expression data points, colored by predicted cluster (left) and by true cluster assignments (right). Axis are the two learned PCA dimensions.}
+    \label{fig:mmplot}
+\vskip -0.2in
+\end{figure*}
 
@@
 \bibliographystyle{sty/icml2013}
+
+\newpage
 
@@
 
-\subsection{Gaussian Mixture Model}
-
-For our Gibb's sampler, we had a fixed burn-in of number of samples (200), a fixed number of sampling iterations after that (500).
-For time reasons, we didn't extensively explore varying these values.
-Figure~\ref{fig:llmm} describes the log-likelihood evaluation of the estimated posteriors.
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=1\textwidth]{figs/llmm}
-    \caption{Log-likelihood of held-out test set, on three seperate runs of Gibbs sampling procedure, described above}
-    \label{fig:llmm}
-\end{figure*}
-
-\label{mmappendix}
-
 \end{document}

### 2017-05-11 finalreport/paper.tex
finish shit
completed final report

@@
 \usepackage{graphicx} % more modern
-\usepackage{booktabs}
 \usepackage{subfigure} 
@@
 
+% For equations
+\usepackage{amsmath}
+
 % For citations
 % \usepackage{natbib}
+
+\usepackage{booktabs}
 
@@
 Tumors cell lines are composed of different sub-populations of cells which often exhibit shared patterns of gene expression.
-Biologists are interested in two key questions: given gene expressions values, (1) can we identify cell clusterings, and (2) can we identify clusters of biologically-related genes?
-Our goal with this project was to answer both these two questions using Bayesian methods.
+Biologists are interested in two key questions: given gene expressions values, (1) can we identify biologically meaningful gene modules, and (2) can we identify cell sub-populations and find gene modules that illuminate their differences?
+Our goal with this project was to address these questions using Bayesian methods.
 
-In brief, we explored the use of three Bayesian clustering methods (a vanilla mixture, integrated topic-mixture, and non-parametric models) to address the first question, and two topic models (vanilla and a dynamic-topic LDA) to address the second.
+To address the first question, we explored the use of Latent Dirichlet Allocation (LDA), two non-parametric topic models, and a dynamic-topic LDA.
+To address the second question, we explored the use of a finite mixture model as well as a clustering topic model.
 
 \subsection{Description of Data}
-Using a contemporary gene sequencing machine, we obtained samples of single-cell RNA-sequencing data (scRNA-seq) taken from tumors in mice.
-Our data consisted of the expression values of 22,712 genes for each of 4645 cells.
-In total, the dataset amounted to 0.86 GB, presenting significant computational challenges when attempting posterior estimation.
+To do our analysis, we used a single-cell RNA-sequencing (scRNA-seq) dataset obtained from human melanoma samples \cite{melanoma}.
+The data consisted of the expression values of 22,712 genes for each of 4645 cells.
+In total, the dataset amounted to 0.86 GB, presenting significant computational challenges when attempting posterior inference.
 Apart from a few computational tricks (online inference, multi-core parallelization), for most methods, we pruned non-informative genes from the dataset, ranking based on the deviation in value across cells \footnote{We recognize that this can bias towards noisy genes.
@@
 We eschewed dimensionality reduction techniques such as PCA because of loss of its direct feature interpretability.
-Prior researchers have labeled the cells in our dataset; there are a total of 9 cell categories.
+Prior researchers have labeled the cells in our dataset; there are a total of 9 cell categories \cite{melanoma}.
 The complete dataset, as well our preprocessed and pruned versions, are openly available \cite{dataset}.
@@
 Our work uses different models to explore gene expression data, but where appropriate (e.g.
-in mixture models in Section~\ref{mmsec}), we compare results.
+LDA vs.
+non-parametric models), we compare results.
 
 \subsection{Structure of Report}
-We first discuss our experiments using Bayesian topic models to discover related sets of genes in a 'topic': LDA in Section~\ref{ldasec} and Dynamic Time Models in Section~\ref{dtmsec}.
-Then, we discuss our experiments in clustering: Dirichlet Mixtures in Section~\ref{mmsec}, Integrated Topic-Clustering in Section~\ref{intsec}, and non-parametric models in Section~\ref{nonparametricsec}.
+We first discuss our experiments using Bayesian topic models to discover topics (i.e.
+related sets of genes) in scRNA-seq data: LDA in Section~\ref{ldasec}, non-parametric topic models in Section ~\ref{nonparametricsec} and Dynamic Time Models in Section~\ref{dtmsec}.
+Then, we discuss our experiments in clustering: Finite Mixtures in Section~\ref{mmsec}, and Integrated Topic-Clustering in Section~\ref{intsec}.
 We conclude our paper with our observations from across all our studies.
@@
 \subsection{Experiments} 
-In lieu of intrusion testing, we evaluated the interpretability of our topics (ranked lists of genes) using using hypergeometric tests \cite{hg}.
-In brief, this determines the probability that our sets of genes would be clustered together in accordance to with existing collections of genes catalogued by biologists in the gene module database \texttt{MSigDB} \cite{msigdb}.
+We evaluated the interpretability of our topics (ranked lists of genes) using hypergeometric tests \cite{hg}.
+In brief, this determines the enrichment of our topics for any existing collections of genes catalogued by biologists in the gene module database \texttt{MSigDB} \cite{msigdb}.
+More details about the score can be found in the Appendix.
 
@@
 It's not clear that this independence assumption is true in the context of gene expression data, because of cross-talk and regulatory mechanisms between gene modules.
+
+\section{Non-Parametric Topic Models} 
+\label{nonparametricsec}
+\subsection{Model Description}
+One limitation of LDA is the need to define the number of topics a priori.
+Non-parametric models overcome this limitation by assuming a countably infinite number of topics.
+We trained and compared two non-parametric topic models with LDA: the Hierarchical Dirichlet Process (HDP) and the Indian Buffet Process Compound Dirichlet Process (IDP-DP).
+On a high level, HDP is very similar to LDA, with the main difference being that the draws from a Dirichlet distribution in LDA are replaced with draws from a Dirichlet process in HDP \cite{HDP}.
+One drawback to the HDP is that there tends to be a correlation between how frequently a topic appears across all documents and how prevalent this topic is within documents that it appears in.
+This may be undesirable in circumstances where there are 'rare' topics with high prevalence in a small number of documents.
+IBP-DP tries to overcome this problem by separately modeling the frequency of documents that contain a topic and the prevalence of this topic within a document where it is present \cite{HDP}.
+Both generative models are described in greater detail in the Appendix (sections \ref{HDP-details} and \ref{IBP-details}).
+
+\subsection{Experiment: LDA vs. HDP}
+
+We trained an HDP topic model and compared it to LDA models with similar numbers of topics, using held-out perplexity and gene set enrichment analysis as described above as our evaluation metrics.
+We used an existing C++ implementation of the Gibbs sampler provided by the authors of the original paper \cite{HDP}.
+Since the available implementation of the posterior inference algorithm for HDP was too slow to run on the entire dataset, we did feature selection using the Seurat toolkit in R \cite{seurat} to reduce the number of genes in our dataset with low variance across cells, and trained all models on this subset of data.
+We tuned the concentration parameters in HDP to obtain a reasonable number of topics (i.e.
+~150); subsequently, we trained LDA models with similar numbers of topics for comparison.
+We found that the HDP model had higher perplexity on a held-out dataset (Figure \ref{fig:hdp-perplexity}).
+Moreover, the topics in the HDP model tended to have less significant enrichment for known gene sets from \texttt{MSigDB}, and we found that this difference to be significant (Figure \ref{fig:hdp-enrichment}, Mann-Whitney-U test, $p<0.0001$).
+Overall, the LDA model proved superior to the HDP model in this experiment.
+
+\subsection{Experiment: LDA vs. IBP-DP}
+One drawback to the HDP is that there tends to be a correlation between how frequently a topic appears across all documents and how prevalent this topic is within documents that it appears in.
+Williamson et al.
+\cite{IBP} proposed the 'focused topic model' to overcome this drawback.
+We implemented their model to compare it with LDA on the Reuters-21578 dataset.
+Since the code from the original IBP-DP paper was not available, we implemented an inference algorithm using collapsed Gibbs sampling \cite{IBP}.
+Due to the non-conjugacy of the model, sampling each latent variable from its full conditional required using another sampling method.
+To sample the topics parameters, the number of which changes depending on how many topics are represented in the dataset, we used slice sampling based on the semi-ordered stick-breaking representation of the model \cite{IBP2}.
+For more details about implementation, please refer to those papers.\\
+
+We tested our code on a subset of the Reuters-21578 dataset, using several different values of the concentration hyper-parameter $\alpha$, which influences the number of clusters.
+Although higher values of $\alpha$ yielded better log-likelihood values, we found that it resulted in a large number of very small topics, which are not very useful (Figure \ref{fig:ibp_words_in_topic}).
+Qualitatively, we did not find the topics from IBP-DP (Figure \ref{fig:ibptopics}) to be more coherent than topics than LDA (Figure \ref{fig:ldatopics}).
+The most prevalent topics from the IBP-DP each corresponded to similar topics from LDA; less prevalent topics tended to consist of a few unrelated words.
+These results discouraged us from optimizing the code to train this model on our scRNA-seq dataset, as we do not think it would yield more coherent gene modules than LDA.
+We emphasize that these results are not completely unexpected, as the authors of the IBP-DP paper did not show any topics from their model, nor did they assess the quality of their model with metrics other than perplexity.
 
@@
 By plotting
-histograms of raw gene expression values, we've found that for many genes the distribution is mostly normal across our data points.
+histograms of whitened gene expression values, we've found that for many genes the distribution is Gaussian across our data points.
 This hints that the generative process in a Gaussian mixture model is plausible.
@@
 \subsection{Model Description}
-\subsection{Implementation} 
+LDA is suitable for modeling corpuses with no special structure, but this is not the case of scRNA-seq data.
+Since the cells within a tissue sample may belong to many different sub-populations, we would expect cells within the same sub-population to have more similar topic proportions.
+We capture this additional structure using the clustering topic model (CTM), a unified framework for clustering and topic modeling based on LDA.
+In this model, which is similar to the one proposed by \cite{wallach_thesis}, the following generative process is assumed:
+
+\begin{enumerate}
+\item Choose a distribution over the clusters, $\pi \sim$ Dir($\pi_0/C$), where $\pi_0$ is a hyperparameter and $C$ is the number of clusters.
+\item Choose a global distribution over the topics, $\phi_0 \sim$ Dir($\alpha_0/T$), where $\alpha_0$ is a hyperparameter and $T$ is the number of topics.
+\item For each cluster $c$, choose a cluster distribution over the topics, $\phi_c \sim$ Dir($\alpha_1 \phi_0$), where $\alpha_1$ is a hyperparameter
+\item For each topic $k$, choose a distribution over the vocabulary, $\beta_k \sim$ Dir($\eta$), where $\eta$ is a hyperparameter.
+\item For each document $d$,
+\begin{enumerate}
+\item Choose a cluster assignment $\xi_d \sim$ Cat($\pi$)
+\item Choose a distribution over topics based on the cluster assignment, $\theta_d \sim$ Dir($\alpha \phi_{xi_d}$), where $\alpha$ is a hyperparameter.
+\item Choose the number of words in this document, $N_d \sim$ Poisson($\kappa$), where $\kappa$ is a hyperparameter.
+\item For the $i$th word in the $d$th document,
+\begin{enumerate}
+\item Choose a topic, $z_{di} \sim$ Categorical($\theta_d$), based on the distribution over topics
+\item Choose a word, $w_{di} \sim$ Categorical($\beta_{z_{di}}$), based on the distribution over the vocabulary for topic $z_{di}$
+\end{enumerate}
+\end{enumerate}
+\end{enumerate}
+
+This model is very similar to LDA, with the addition of cluster assignments that influence document topic proportions.
+Note that the documents in each cluster have more similar topic proportions, particularly if we choose a reasonably large value of $\alpha$.
+
+\subsection{Inference}
+We implemented a collapsed Gibbs sampling in \texttt{MATLAB} to perform inference for the CTM, integrating out all latent variables except the topic assignments $z$ and the cluster assignments $\xi$ to facilitate mixing.
+First, we establish the prior distribution of $z_{di}$:
+\begin{equation} \label{eq:z_prior}
+p(z_{di} = k | z_{-di}, \xi_d) = \frac{N_{d,k} + \alpha\frac{N_{\xi_d,k} + \alpha_1\frac{N_k + \alpha_0/K}{N + \alpha_0}}{N_{\xi_d}} + \alpha_1}{N_d + \alpha}
+\end{equation}
+where $z_{-di}$ refers to topic assignments other than that for word $i$ in document $d$, $N_{d,k}$ is the number of words in document $d$ assigned to topic $k$, $N_d$ is the total number of words in document $d$, $N_{\xi_d,k}$ is the number of words in the documents from cluster $\xi_d$ assigned to topic $k$, $N_{\xi_d}$ is the total number of words in the documents from cluster $\xi_d$, $N_k$ is the number of words in the corpus assigned to topic $k$, and $N$ is the total number of words in the corpus\footnote{When counting the number of words, we always exclude the word currently being considered, but we omit this in the equation to avoid clutter.}.
+It follows that the full conditional for each $z_{di}$ is:
+\begin{equation}
+\left.\begin{aligned}
+&p(z_{di} = k | w_{di}, w_{-di}, z_{-di}, \xi_d)\\
+&\propto p(w_{di} | z_{di} = k, w_{-di})p(z_{di} = k | z_{-di}, \xi_d)\\
+&= \frac{N_{w_{di},k}}{N_k}p(z_{di} = k | z_{-di}, \xi_d)
+\end{aligned}\right.
+\end{equation}
+where $N_{w_{di},k}$ is the number of times the word $w_{di}$ appears in topic $k$, $N_k$ is the total number of words assigned to topic $k$, and the prior of $z_{di}$ is given by equation \ref{eq:z_prior}.
+Finally, the full conditional for each $\xi_d$ is:
+\begin{equation}
+\left.\begin{aligned}
+&p(\xi_d = c | \xi_{-d},z_d, z_{-d}; \pi)\\
+&\propto p(z_d | \xi_d = c, z_{-d})p(\xi_d = c | \xi_{-d})\\
+&= Multi(z_d, p(z_{d}| z_{-d}, \xi_d = c)) \frac{M_{d,c} + \pi_0/C}{M_d + \pi_0}
+\end{aligned}\right.
+\end{equation}
+where $M_{c}$ is the number of documents assigned to topic $c$ and $M$ is the total number of documents in the corpus\footnote{When counting the number of documents, we exclude the document currently being considered, but omit this in the equation to avoid clutter}.
+$Multi(z_d, p(z_{d}| \cdots))$ is the probability under the multinomial distribution of getting topic assignments $z_d$ when the topic proportions are given by $p(z_{d}| \cdots)$.
+The topic proportions are obtained essentially as described in equation \ref{eq:z_prior}, except we set the number of words in document $d$ to be $0$ and adjust all the other counts accordingly.
+
 \subsection{Experiments}
 
-\section{Non-parametric Models: IBP and HDP} 
-\label{nonparametricsec}
-\subsection{Model Description} 
-\subsection{Implementation} 
-\subsection{Experiments}
+We trained the CTM on a subset of 500 cells and 3000 genes from the human melanoma scRNA-seq dataset using the collapsed Gibbs sampler until convergence (about 500 iterations).
+All hyperparameters were set to $1$ with the exception of $\alpha$, which we set to $10$ to encourage similarity of topic proportions for cells assigned to the same cluster, and the numbers of clusters and topics were set to $10$ and $20$ respectively.
+To determine the efficacy of the clustering, we generated a tSNE plot of the cells using the Euclidean distance between their topic proportions (Figure \ref{fig:clusters_CTM}).
+Compared to a tSNE plot based on the Euclidean distance between the counts of all 3000 genes (Figure \ref{fig:unclustered}), in which there was no obvious grouping of cells, we found that the topic proportions were nicely grouped based on their learned cluster assignments.
+To determine if our clusters had any biological significance, we compared them with cell labels published by the original authors (Figure \ref{fig:clusters_true})\footnote{The authors published the cell labels based on prior biological knowledge, i.e.
+thresholding the cells based on known gene markers of different cell types.
+Due to noise and dropout in scRNA-seq data, they were not able to label all cells using this approach.}.
+Interestingly, our clusters corresponded nearly exactly to the different cell types, and we were even able to classify the cells with unknown labels.
+\\
 
-\section{Conclusions} 
-\label{nonparametricsec}
+\begin{figure}
+    \centering
+    \includegraphics[width=0.35\textwidth]{figs/clusters_CTM}
+    \caption{tSNE plot of cells from human melanoma, plotted using Euclidean distances between topic proportions, with cluster assignments shown in color.
+The cells nicely group into their cluster assignments, suggesting that the clustering topic model inference algorthm worked as intended.
+Moreover, the cluster assignments correpond to the cell labels found by the group that published the dataset, see Figure \ref{fig:clusters_true} }
+    \label{fig:clusters_CTM}
+\end{figure}
+\begin{figure}
+    \centering
+    \includegraphics[width=0.35\textwidth]{figs/clusters_true}
+    \caption{tSNE plot of cells from human melanoma, plotted using Euclidean distances between topic proportions, with labels published by the group that collected the dataset.
+0 - label unknown; 1 - T-cell; 2 - B-cell; 3 - Macrophage; 4 - Endoderm; 5 - Cancer-associated fibroblast; 6 - Natural Killer; 7 - Malignant.
+The labels correpond to the cluster assignments learned using the CTM model.}
+    \label{fig:clusters_true}
+\end{figure}
+
+Subsequently, we wanted to determine if the learned topics were meaningful.
+We first used the biological relevance score described earlier and found that the topics were significantly enriched for some known gene sets from \texttt{MSigDB} (Figure \ref{fig:ctm_pvals}).
+We then did some examination of the specific genes appearing in select topics using a combination of the \texttt{MSigDB} analysis and \texttt{GO}-enrichment analysis\footnote{\texttt{GO} enrichment analysis looks for enrichment of genes associated with cell processes.
+Whereas \texttt{MSigDB} is a database of various published gene sets, the \texttt{GO} database contains only those well-characterized gene modules that are generally accepted as being involved in a particular cell process}.
+We found that topic $18$, which was prevalent in the T-cells and not in other cells (Figure \ref{fig:tcelltopics}), was highly enriched for genes involved in "myeloid dendritic cell activation involved in immune response".
+Topic $3$, which was prevalent in the B-cells and not elsewhere (Figure \ref{fig:bcelltopics}), was highly enriched for genes involved in "lymphocyte activation" and "inflammatory response".
+Topic $4$, which was highly prevalent in the T-cells and B-cells only (Figures \ref{fig:tcelltopics} and \ref{fig:bcelltopics}), was highly enriched for genes involved in "somatic diversification of immune receptors".
+Finally, topic $8$ which was mainly present in the tumor cells (\ref{fig:tumortopics}), was enriched for genes involved in "G1 to S transition of mitotic cell cycle" and "cell division".
+The correspondence between the genes enriched in the topics and the cell sub-populations with those topics suggests that our learned topics have biological significance.
+As a result, the highly-ranked genes in some of these topics may be of interest to experimental biologists as candidate gene markers for specific cell sub-populations.
+
+\section{Discussion}
+
+In this paper, we sought to address the following two questions using Bayesian topic models: given gene expression data, (1) can we identify biologically meaningful gene modules, and (2) can we identify cell sub-populations and find gene modules that illuminate their differences.
+\\
+
+\subsection{Can we identify biologically meaningful gene modules?}
+We addressed the first question by trying to model scRNA-seq data with LDA as well as two non-parametric topic models, the HDP and the IBP-DP.
+The purpose of doing this was mainly to determine if topic models based on LDA are suitable for modeling gene expression; as far as we are aware, there are no publications applying LDA to scRNA-seq data.
+Our results from section \ref{ldasec} suggest that LDA could potentially be useful for scRNA-seq, since our topics were enriched for previously discovered biological gene sets.
+One major consideration is computational cost: although scRNA-seq data is sparse, it is not nearly as sparse as the bag-of-words representation of a corpus, so most inference algorthms take a long time to run on the entire dataset.
+Based on our results, however, it appears that this problem can be addressed either by using a faster posterior inference method such as online variational inference or by pre-processing the dataset for those genes with the greatest variability.
+
+Compared to LDA, we found that HDP and IBP-DP were not as suitable for modeling scRNA-seq data (Section \ref{nonparametricsec}).
+The HDP underperformed in comparison to LDA in both log-likelihood on a held-out dataset and the biological significance scores of its topics.
+The IBP-DP did not seem to work better than LDA on a subset of the Reuters-21578 corpus, based on a qualitative analysis of the topics, so we did not try it on the gene expression data, as we do not expect it to perform better than LDA.
+The underperformance of the IBP-DP compared to LDA on the Reuters-21578 dataset does not surprise us, as the authors of the IBP-DP paper did not show any topics from their model, even though they claimed it may produce more 'focused' topics.
+
+After we were confident that LDA could be useful for modeling scRNA-seq data, we wanted to try models that could capture higher-level structure of gene expression data - for instance, time.
+As cells proliferate, they may differentiate into different cell types.
+A useful topic model might learn not just the topics at a given time point, but also how the topics are evolving over time.
+We have presented some preliminary results applying a dynamic topic model to our scRNA-seq dataset and selecting an appropriate number of topics (Section \ref{dtmsec}).
+To further test this model, experiments could be done on simulated time-series scRNA-seq data with known evolving topics, and one could evaluate how well those evolving topics are picked up by the model.
+
+\subsection{Can we identify cell sub-populations and find gene modules that illuminate their differences?}
+In addition to time, another source of structure in the scRNA-seq dataset is the clustering of cells into sub-populations.
+In biology, there is much interest in identifying these sub-populations and finding gene modules or markers that are characteristic of different sub-populations.
+We first wanted to determine if a Bayesian finite mixture model would be appropriate for separating the cells into clusters (Section \ref{mmsec}).
+We found that we were approximately able to learn the cell type labels published by \cite{melanoma}\\
+
+Finally, we wanted to create a unified framework for cell clustering and topic modeling.
+We implemented the clustering topic model (Section \ref{intsec}) and found that it was able to correctly separate the cells into sub-populations.
+It appeared to outperform the vanilla mixture model without clustering, suggesting that knowledge of the topics helps the model learn better clusters.
+We also found that several topics corresponded to the cell types that they were prevalent in; we believe these results suggest that biologists can use the topics - which are ranked lists of genes - as candidate gene markers for cell sub-populations of interest within the melanoma, such as malignant cells, T-cells and B-cells.
+Whether these candidate gene markers are truly useful must be confirmed by biologists experimentally.
+Looking forward, it would be very interesting to combine the dynamic topic model and the clustering topic model, so that we can capture the time-evolution of the cell sub-populations within a scRNA-seq dataset.
 
@@
 
-\section{Breakdown of Work}
-The implementation(s) and experiments involving LDA, Gaussian Mixtures, and Dynamic Time Model was completed by SK (Section~\ref{ldasec}, Section~\ref{mmsec}, and Section~\ref{dtmsec}).
-KY completed the implementations and experiments involved Integrated Topic-Clustering, non-parametric models, (Section~\ref{intsec} and Section~\ref{nonparametricsec}) and the shared gene enrichment test implementation.
+\section{Author Contributions}
+The implementation(s) and experiments involving LDA, Dirichlet Mixtures, and Dynamic Time Model was completed by SK (Section~\ref{ldasec}, Section~\ref{dtmsec}, and Section~\ref{mmsec}).
+KY completed the implementations and experiments involved non-parametric topic models, the clustering topic model, and the gene set enrichment test implementation (Section~\ref{nonparametricsec}, Section~\ref{intsec}, and Section~\ref{enrichment_score}).
 The authors contributed equally in this work.
@@
 
-\newpage
+\section{Appendix}
 
-\section{Appendix}
+\subsection{Biological Significance Score / Gene Set Enrichment} \label{enrichment_score}
+In general, it is difficult to evaluate the 'coherence' of topics of genes, since unlike topics of words, we cannot just read them and get a sense of how similar they are.
+Therefore, nothing like the intrusion test exists for evaluating topics of genes.
+To overcome this, we developed our own test for the biological significance of gene topics.
+The \texttt{MSigDB} contains thousands of gene sets discovered through previous research, catalogued by biologists from the Broad Institute.
+We deduced that if a topic is 'coherent', then it is likely significantly enriched for some of these gene sets.
+For each topic, our biological significance scoring method iterates through the \texttt{MSigDB} database and tests each gene set for enrichment using the minimum hyper-geometric test \cite{hg}.
+The most significant p-values for each topic are recorded and used as a proxy for its biological coherence.
 
@@
 \label{ldaappendix}
-
-This section includes Figure~\ref{fig:pathways} on the next page.
-
 For our Gibb's sampler, we had a fixed burn-in of number of samples (200), a fixed number of sampling iterations after that (500).
@@
 There was one sampling chain on each of four cores.
+
+\subsection{HDP}
+\label{HDP-details}
+HDP is an infinite topic extension of LDA based on the Dirichlet process \cite{HDP}. Briefly, words in a corpus are assumed to be generated as follows:
+\begin{enumerate}
+\item Sample the global distribution over topics, $G_0 \sim$ DP($\gamma, H$), from a Dirichlet process with concentration $\gamma$ and base Dirichlet distribution $H$\footnote{More specifically, $H$ is a Dirichlet distribution over the $(V-1)$-dimensional simplex, where $V$ is the size of the vocabulary, and $G_0$ is a countably infinite set of point masses over this simplex whose weights sum to $1$.
+$G_0$ in HDP plays a role analogous to $\alpha$ in LDA.}.
+\item For each document $d$,
+\begin{enumerate}
+\item Sample the local distribution over topics, $G_d \sim$ DP($\alpha_0, G_0$), from a Dirichlet process with concentration $\alpha_0$ and base distribution $G_0$.
+\item Choose the number of words in this document, $N_d \sim$ Poisson($\xi$), where $\xi$ is a hyperparameter.
+\item For the $i$th word in the $d$th document,
+\begin{enumerate}
+\item Choose a topic (i.e. distribution over words in vocab), $\beta_{di} \sim$ $G_d$, based on the local distribution over topics
+\item Choose a word, $w_{di} \sim$ Categorical($\beta_{di}$), based on the distribution over the vocabulary in the topic
+\end{enumerate}
+\end{enumerate}
+\end{enumerate}
+
+\subsection{IBP-DP}
+\label{IBP-details}
+One drawback to the HDP is that there tends to be a correlation between how frequently a topic appears across all documents and how prevalent this topic is within documents that it appears in.
+Williamson et al.
+\cite{IBP} proposed the 'focused topic model' to overcome this drawback.
+In their model, the frequency of a topic $k$ depends on two separate variables, its relative prevalence within a document ($\phi_k$) and the probability that a given document contains this topic ($\pi_k$), thus reducing correlation between the two.
+Briefly, the words in the corpus are assumed to be generated as follows:
+
+\begin{enumerate}
+\item For each topic $k = 1, 2, ...$
+\begin{enumerate}
+\item Choose a population frequency (i.e. probability that a document contains topic) $\pi_k = \prod^k_{j=1} \mu_k$, where each $\mu_k \sim Beta(\alpha, 1)$, where $\alpha$ is a hyperparameter.
+\item Choose a relative prevalence (i.e. how often the topic appears within a document containing the topic) $\phi_k \sim Gamma(\gamma, 1)$, where $\gamma$ is a hyperparameter.
+\item Choose a distribution over the vocabulary, $\beta_k \sim$ Dir($\eta$), where $\eta$ is a hyperparameter.
+\end{enumerate}
+\item For each document $d$,
+\begin{enumerate}
+\item Choose the topics that will appear as a binary vector, $b_d \sim Bernoulli(\pi)$
+\item Choose the topic proportion as $\theta_d \sim Dirichlet(b_d\cdot \phi)$
+\item Choose the number of words in this document, $N_d \sim$ NegativeBin($\sum_k b_{dk}phi_k$, $\delta$), where $\delta$ is a hyperparameter.
+\item For the $i$th word in the $d$th document,
+\begin{enumerate}
+\item Choose a topic, $z_{di} \sim$ Categorical($\theta_d$), based on the distribution over topics
+\item Choose a word, $w_{di} \sim$ Categorical($\beta_{z_{di}}$), based on the distribution over the vocabulary for topic $z_{di}$
+\end{enumerate}
+\end{enumerate}
+\end{enumerate}
 
@@
 \label{dtmappendix}
-
-This section includes Figure~\ref{fig:pathwaysdtm} on the next page.
-
 Here is a description of the generative process for our dynamic topic model:
@@
     \centering
-    \includegraphics[width=1\textwidth]{figs/pathwaysdtm}
-    \caption{Matches between the gene collections found using a dynamic topic model and published gene sets in \texttt{MSigDB}.
-$k$ refers to the number of topics in the model.
-$t$ refers to the time slice in the model.}
-    \label{fig:pathwaysdtm}
+    \includegraphics[width=0.5\textwidth]{figs/hdp-perplexity}
+    \caption{Comparison of log-likelihood of the held-out testing set, under various LDA models and the HDP model.}
+    \label{fig:hdp-perplexity}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/hdp-enrichment}
+    \caption{Comparison of distributions of p-values from gene set enrichment analysis between LDA models and the HDP model.}
+    \label{fig:hdp-enrichment}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/ibp_words_in_topic}
+    \caption{Beeswarm plot of number of words per topic, for 3 different IBP-DP models with different concentration parameters. Each point represents one topic from its model}
+    \label{fig:ibp_words_in_topic}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=1\textwidth]{figs/ibptopics}
+    \caption{Topics from IBP-DP model trained on subset of Reuters dataset.}
+    \label{fig:ibptopics}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=1\textwidth]{figs/ldatopics}
+    \caption{Topics from LDA model with 15 topics trained on subset of Reuters dataset.}
+    \label{fig:ldatopics}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.35\textwidth]{figs/unclustered}
+    \caption{tSNE plot of cells from human melanoma, plotted using Euclidean distances between gene count vectors, with cluster assignments shown in color.}
+    \label{fig:unclustered}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.35\textwidth]{figs/ctm_pvals}
+    \caption{Histogram of topics from CTM model based on p-value of enrichment of \texttt{MSigDB} gene sets.}
+    \label{fig:ctm_pvals}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.35\textwidth]{figs/tumortopics}
+    \caption{Bar charts of topic proportions for malignant cell clusters.}
+    \label{fig:tumortopics}
+\end{figure*}
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/tcelltopics}
+    \caption{Bar charts of topic proportions for T-cell clusters.}
+    \label{fig:tcelltopics}
+\end{figure*}
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.35\textwidth]{figs/bcelltopics}
+    \caption{Bar charts of topic proportions for B-cell clusters}
+    \label{fig:bcelltopics}
 \end{figure*}

### 2017-05-11 finalreport/paper.tex
fixed figure

@@
 \subsection{Structure of Report}
-We first discuss our experiments using Bayesian topic models to discover topics (i.e.
-related sets of genes) in scRNA-seq data: LDA in Section~\ref{ldasec}, non-parametric topic models in Section ~\ref{nonparametricsec} and Dynamic Time Models in Section~\ref{dtmsec}.
+We first discuss our experiments using Bayesian topic models to discover topics in scRNA-seq data: LDA in Section~\ref{ldasec}, non-parametric topic models in Section ~\ref{nonparametricsec} and Dynamic Time Models in Section~\ref{dtmsec}.
 Then, we discuss our experiments in clustering: Finite Mixtures in Section~\ref{mmsec}, and Integrated Topic-Clustering in Section~\ref{intsec}.
@@
 \section{Author Contributions}
-The implementation(s) and experiments involving LDA, Dirichlet Mixtures, and Dynamic Time Model was completed by SK (Section~\ref{ldasec}, Section~\ref{dtmsec}, and Section~\ref{mmsec}).
+The implementations and experiments involving LDA, Gaussian Mixtures, and the Dynamic Topic Model was completed by SK (Section~\ref{ldasec}, Section~\ref{dtmsec}, and Section~\ref{mmsec}).
 KY completed the implementations and experiments involved non-parametric topic models, the clustering topic model, and the gene set enrichment test implementation (Section~\ref{nonparametricsec}, Section~\ref{intsec}, and Section~\ref{enrichment_score}).
@@
     \centering
+    \includegraphics[width=1\textwidth]{figs/pathwaysdtm}
+    \caption{Matches between the gene collections found in DTM topics and published gene sets in \texttt{MSigDB}.
+$k$ refers to the number of topics in the model, $t$ refers to one of three time slices.}
+    \label{fig:pathwaysdtm}
+\end{figure*}
+
+\begin{figure*}
+    \centering
     \includegraphics[width=0.5\textwidth]{figs/hdp-perplexity}
